# Data Sources — Sweden

Documentation of all external datasets used in the Sweden gap finder pipeline.

> **Location:** All raw data in `data/SE/raw/`. Never edited manually.
> All processing outputs go to `data/SE/proc/`.

```{r ds-setup, include=FALSE}
# Citations are resolved + validated once by scripts/01b_resolve_data_sources.R
# and cached here. This document, the app, and the reports all read this same
# object, so DOIs never drift. If it's missing, run 01b first.
ds_meta_path <- here::here("data", "SE", "proc", "data_sources_meta.rds")
ds <- if (file.exists(ds_meta_path)) readRDS(ds_meta_path) else NULL

# Inline helpers so checklist citations are read from the resolved metadata
# (config download/dataset key -> GBIF API) instead of being hand-typed in prose.
ck_doi <- function(key) {
  x <- if (!is.null(ds)) ds$checklists[[key]] else NULL
  if (is.null(x) || is.null(x$doi) || is.na(x$doi)) return("\u2014")
  sprintf("[%s](%s)", sub("https://doi.org/", "", x$doi), x$doi)
}
ck_dataset <- function(key) {
  x <- if (!is.null(ds)) ds$checklists[[key]] else NULL
  if (is.null(x) || is.null(x$dataset_key) || is.na(x$dataset_key)) return("\u2014")
  sprintf("[%s](https://www.gbif.org/dataset/%s)", x$dataset_key, x$dataset_key)
}
```

---

## 1. GBIF Occurrence Cubes (SQL API)

**Purpose:** Species-level, gridded occurrence summaries for Sweden.

**Source:** GBIF SQL Download API
**Downloaded:** automatically by `scripts/01a_download_raw_data.R` (download dates in the table below)
**Format:** SQL_TSV_ZIP → CSV → Parquet (converted by script 04)

### How a cube is made

- **One query file.** The cube is defined by `sql/gbif_occurrence_cube.sql`. Both cubes use it
  and differ only in grid resolution (10,000 vs 50,000 m). `render_cube_sql()` (R/globals.R)
  fills in the country code, the resolution and the Catalogue of Life checklist key.
- **Automated download.** When GBIF credentials are set, 01a submits the query with
  `rgbif::occ_download_sql()`, waits for GBIF and imports the result into `data/SE/raw/cubes/`.
  Without credentials it prints the same query to run by hand at the
  [SQL API](https://www.gbif.org/occurrence/download/sql). 01a only downloads when fewer than
  two cube files are on disk.
- **Provenance records (version-controlled, written by the pipeline, never by hand):**
  - `provenance/cube_downloads_SE.yml`: the GBIF download key of each cube, written by 01a
    after every automated download.
  - `provenance/upstream_versions_SE.yml`: download key, DOI and creation date of each cube,
    and the release of every checklist, written by `scripts/01b_resolve_data_sources.R` on
    every run. `git diff provenance/` after a run shows whether anything upstream changed.
- **Citation.** 01b resolves each key on the GBIF API (DOI, record count, licence) and fails
  loudly if a key does not exist on gbif.org. This document, the app and the reports read the
  result, so DOIs are never typed by hand.
- **Pinning an older cube.** To cite a specific historical download instead of the latest,
  add `cubes.<grid>.download_key` to `configs/config_SE.yml`; it takes precedence over the
  provenance file.

### SQL query

The query as run (comments removed; `${...}` are filled in by `render_cube_sql()`):

```{r ds-sql, echo=FALSE, results='asis'}
sql_path <- here::here("sql", "gbif_occurrence_cube.sql")
if (file.exists(sql_path)) {
  sql <- readLines(sql_path, encoding = "UTF-8")
  sql <- sql[!grepl("^\\s*--", sql) & nzchar(trimws(sql))]
  cat("```sql", sql, "```", sep = "\n")
} else {
  cat("> *`sql/gbif_occurrence_cube.sql` not found.*\n")
}
```

- `specieskey` is pinned to the Catalogue of Life through `classificationdetails`, so it is a
  COL taxonID, not a GBIF Backbone key. Most are alphanumeric (e.g. `6VFN8`); some are
  numeric (e.g. `67343` = *Anemone nemorosa*).
- Grid radius 0: cell assignment is deterministic.

### Columns (17)

The 14 grouping and count columns, plus three measures that make the cube compatible with
b-cubed tools (`b3gbi::process_cube()`).

| # | Column | Description |
|--:|--------|-------------|
| 1 | specieskey | Catalogue of Life taxonID of the species (pinned via `classificationdetails`) |
| 2 | species | Scientific name |
| 3 | kingdom | Kingdom name |
| 4 | phylum | Phylum name |
| 5 | class | Class name |
| 6 | order | Order name |
| 7 | family | Family name |
| 8 | basisofrecord | Record type (HUMAN_OBSERVATION, PRESERVED_SPECIMEN, etc.) |
| 9 | publishingorgkey | UUID of the publishing organisation |
| 10 | datasetkey | UUID of the source dataset |
| 11 | eeacellcode | EEA reference-grid cell code (10 km or 50 km) |
| 12 | year | Observation/collection year |
| 13 | month | Observation/collection month |
| 14 | occurrences | Record count for the group (species × cell × year × month × record type × publisher × dataset) |
| 15 | mincoordinateuncertaintyinmeters | Smallest coordinate uncertainty in the group, in metres; a missing value counts as 1,000 m |
| 16 | mintemporaluncertainty | Smallest temporal uncertainty of the event date in the group, in seconds (`GBIF_TemporalUncertainty`) |
| 17 | distinctobservers | Number of distinct `recordedBy` values in the group |

Older 14-column cubes still convert; scripts 04 and 05 report the three measures as absent.

### Downloads

```{r ds-cubes, echo=FALSE, results='asis'}
if (is.null(ds)) {
  cat("> *Run `scripts/01b_resolve_data_sources.R` to populate resolved DOIs,",
      "record counts and validation status.*\n")
} else {
  cat("| Resolution | Created | Records | Download Key | DOI | Resolves |\n")
  cat("|-----------|---------|--------:|--------------|-----|:--------:|\n")
  for (nm in names(ds$cubes)) {
    cb  <- ds$cubes[[nm]]
    res <- sub("grid", "", nm); res <- sub("km", " km", res)
    created <- if (isTRUE(cb$resolves) && !is.null(cb$created) && !is.na(cb$created)) format(as.Date(cb$created)) else "—"
    recs <- if (isTRUE(cb$resolves) && !is.na(cb$records)) format(cb$records, big.mark = ",") else "—"
    doi  <- if (isTRUE(cb$resolves)) sprintf("[%s](%s)", sub("https://doi.org/", "", cb$doi), cb$doi) else "—"
    ok   <- if (isTRUE(cb$resolves)) "✅" else "❌"
    cat(sprintf("| %s | %s | %s | `%s` | %s | %s |\n", res, created, recs, cb$download_key, doi, ok))
  }
  cat(sprintf("\n*Resolved from download keys via the GBIF API on %s.*\n",
              format(as.Date(ds$resolved_at))))
}
```

### Files

```
data/SE/raw/cubes/
├── <download key>.zip     # as delivered by GBIF (one per resolution)
├── cube_10km.csv          # extracted 10 km cube
└── cube_50km.csv          # extracted 50 km cube
data/SE/proc/cubes/        # Parquet versions (script 04)
```

---

## 2. EEA Reference Grids (shared)

Both grids are Europe-wide and shared across all European countries. They live in `data/shared/grids/` and the pipeline clips to the country extent automatically.

### 2.1 EEA 10km Grid (Europe)

**Source:** [EEA GeoNetwork](https://sdi.eea.europa.eu/datastore/public?path=/eea_v_3035_10_km_eea-ref-grid-europe_p_2011_v01_r00)
**Version:** 2011
**CRS:** ETRS89-LAEA Europe (EPSG:3035)
**Coverage:** All of Europe
**Sweden cells:** 7,693 (after clipping)

```
data/shared/grids/
├── Grid_ETRS89-LAEA_10K.shp (+ .dbf, .prj, .shx)  # Europe-wide
```

### 2.2 EEA 50km Grid (Europe)

**Source:** [EEA GeoNetwork](https://sdi.eea.europa.eu/catalogue/srv/eng/catalog.search#/metadata/aac8379a-5c4e-445c-b2ef-23a6a2701ef0)
**Version:** v2024

```
data/shared/grids/
└── EEA_50km_grid_v2024.gpkg
```

---

## 3. Taxonomic Reference Datasets

### 3.1 Dyntaxa — Swedish Taxonomic Database

**Purpose:** National taxonomy backbone for Swedish taxa.
**DOI:** `r ck_doi("taxonomy")`
**GBIF Dataset:** `r ck_dataset("taxonomy")`
**Downloaded:** 2026-02-05
**Taxa:** ~110,000

```
data/SE/raw/taxonomy/
├── Taxon.csv                 # Taxonomic backbone
├── SpeciesDistribution.csv   # establishmentMeans, occurrenceStatus, threatStatus
├── VernacularName.csv        # Swedish and English common names
├── Reference.csv
├── meta.xml
└── eml.xml
```

Key columns in SpeciesDistribution.csv:

| Field | Values | Count |
|-------|--------|-------|
| establishmentMeans | native (62,394), introduced (4,755), invasive (450), naturalised (176), uncertain (2,819), blank (39,416) |
| occurrenceStatus | present (64,553), absent (9,892), doubtful (2,122), irregular (5,091), blank (28,352) |

### 3.2 Swedish Red List (SLU Artdatabanken)

**Purpose:** Threat status enrichment for the taxonomy backbone.
**DOI:** `r ck_doi("redlist")`
**GBIF Dataset:** `r ck_dataset("redlist")`
**Version:** 1.9
**Downloaded:** 2026-03-25
**Taxa:** ~11,000

```
data/SE/raw/redlist/
├── taxon.txt
├── distribution.txt
├── meta.xml
└── eml.xml
```

---

## 4. Invasive Species Registry

**Purpose:** Flag invasive alien species in the taxonomy backbone.
**DOI:** `r ck_doi("invasives")`
**GBIF Dataset:** `r ck_dataset("invasives")`
**Taxa:** ~450 invasive species

```
data/SE/raw/invasives/
├── taxon.txt
├── distribution.txt
├── meta.xml
└── eml.xml
```

Joined to the taxonomy backbone in script 03 as `is_invasive` (TRUE/FALSE) by `scientificName`.

---

## 5. Sensitive Species List (SLU Artdatabanken)

**Purpose:** Flag restricted access species with coordinate generalization categories.
**DOI:** `r ck_doi("sensitive")`
**GBIF Dataset:** `r ck_dataset("sensitive")`
**Published by:** SLU Artdatabanken
**Version:** 1.9 (March 2026)
**Taxa:** 203 restricted access species
**Last revised:** 2023

```
data/SE/raw/sensitive/
├── taxon.txt          # 203 species with sensitivity category in taxonRemarks
├── meta.xml
└── eml.xml
```

**Note:** No distribution file — the sensitivity category (5km, 25km, or 50km generalization)
is stored in the `taxonRemarks` column of `taxon.txt`. Script 03 detects this automatically.

Joined to the taxonomy backbone as `is_sensitive` (TRUE/FALSE) + `sensitivity_category` (e.g. "50km").

Taxonomic coverage: birds, mammals, Hymenoptera, butterflies, beetles, spiders,
vascular plants, lichens and fungi.

---

## 6. Administrative Boundaries (GADM)

**Purpose:** Region and municipality overlays for maps and regional filtering.
**Source:** [GADM v4.1](https://gadm.org/)
**Downloaded:** 2026-03-17

```
data/SE/raw/admin/
├── admin_level1.gpkg    # 21 regions (län)
└── admin_level2.gpkg    # 290 municipalities (kommuner)
```

Downloaded automatically via the `geodata` R package.

---

## Updating Data

**Red list / Invasive / Sensitive species update:**
1. Delete files in the relevant `data/SE/raw/{source}/` directory
2. Update the `dataset_key` in `configs/config_SE.yml` if the source moved to a new GBIF dataset
3. Run `source("scripts/01a_download_raw_data.R")`
4. Re-run pipeline from script 03 (DOIs in this document resolve automatically via `01b`)

**New occurrence cubes:**
1. Move the files in `data/SE/raw/cubes/` aside (01a downloads only when fewer than two cube files are there)
2. Run `targets::tar_invalidate(raw_data)`, then `targets::tar_make()`. 01a submits and imports both cubes,
   writes their keys to `provenance/cube_downloads_SE.yml`, and 01b records DOIs and dates in
   `provenance/upstream_versions_SE.yml`; script 04 rebuilds the parquets because the CSVs are newer
3. Compare the new cube with the previous one before it replaces it (ROADMAP: "Data refresh after the launch")
4. Commit the two provenance files with the new app bundle

**Taxonomy update:**
1. Delete files in `data/SE/raw/taxonomy/`
2. Run `source("scripts/01a_download_raw_data.R")`
3. Re-run full pipeline from script 03

---

**Last updated:** 2026-10-07
**Maintainer:** Lena Thöle
