# Data Sources — Sweden

Documentation of all external datasets used in the Sweden gap finder pipeline.

> **Location:** All raw data in `data/SE/raw/`. Never edited manually.
> All processing outputs go to `data/SE/proc/`.

```{r ds-setup, include=FALSE}
# Citations are resolved + validated once by scripts/01b_resolve_data_sources.R
# and cached here. This document, the app, and the reports all read this same
# object, so DOIs never drift. If it's missing, run 01b first.
ds_meta_path <- here::here("data", "SE", "proc", "data_sources_meta.rds")
ds <- if (file.exists(ds_meta_path)) readRDS(ds_meta_path) else NULL

# Inline helpers so checklist citations are read from the resolved metadata
# (config download/dataset key -> GBIF API) instead of being hand-typed in prose.
ck_doi <- function(key) {
  x <- if (!is.null(ds)) ds$checklists[[key]] else NULL
  if (is.null(x) || is.null(x$doi) || is.na(x$doi)) return("\u2014")
  sprintf("[%s](%s)", sub("https://doi.org/", "", x$doi), x$doi)
}
ck_dataset <- function(key) {
  x <- if (!is.null(ds)) ds$checklists[[key]] else NULL
  if (is.null(x) || is.null(x$dataset_key) || is.na(x$dataset_key)) return("\u2014")
  sprintf("[%s](https://www.gbif.org/dataset/%s)", x$dataset_key, x$dataset_key)
}
```

---

## 1. GBIF Occurrence Cubes (SQL API)

**Purpose:** Species-level, gridded occurrence summaries for Sweden.

**Source:** GBIF SQL Download API
**Downloaded:** 2026-03-25
**Format:** SQL_TSV_ZIP → CSV → Parquet (converted by pipeline)

### SQL Query

Both cubes use the same query, differing only in grid resolution (10000 vs 50000 meters):

```sql
SELECT
  specieskey, species, kingdom, phylum, class, "order", family,
  basisofrecord, publishingorgkey, datasetkey,
  GBIF_EEARGCode(10000, decimallatitude, decimallongitude, 0) AS eeacellcode,
  "year", "month",
  COUNT(*) AS occurrences
FROM occurrence
WHERE countrycode = 'SE'
  AND hascoordinate = TRUE
  AND hasgeospatialissues = FALSE
  AND occurrencestatus = 'PRESENT'
  AND specieskey IS NOT NULL
GROUP BY specieskey, species, kingdom, phylum, class, "order", family,
  basisofrecord, publishingorgkey, datasetkey,
  GBIF_EEARGCode(10000, decimallatitude, decimallongitude, 0),
  "year", "month"
```

### Columns (14)

| # | Column | Description |
|--:|--------|-------------|
| 1 | specieskey | GBIF species identifier (backbone key) |
| 2 | species | Scientific name (GBIF backbone) |
| 3 | kingdom | Kingdom name |
| 4 | phylum | Phylum name |
| 5 | class | Class name |
| 6 | order | Order name |
| 7 | family | Family name |
| 8 | basisofrecord | Record type (HUMAN_OBSERVATION, PRESERVED_SPECIMEN, etc.) |
| 9 | publishingorgkey | UUID of the publishing organisation |
| 10 | datasetkey | UUID of the source dataset |
| 11 | eeacellcode | EEA reference-grid cell code (10 km or 50 km) |
| 12 | year | Observation/collection year |
| 13 | month | Observation/collection month |
| 14 | occurrences | Record count for the species × cell × year × month |

### Downloads

```{r ds-cubes, echo=FALSE, results='asis'}
if (is.null(ds)) {
  cat("> *Run `scripts/01b_resolve_data_sources.R` to populate resolved DOIs,",
      "record counts and validation status.*\n")
} else {
  cat("| Resolution | Records | Download Key | DOI | Resolves |\n")
  cat("|-----------|--------:|--------------|-----|:--------:|\n")
  for (nm in names(ds$cubes)) {
    cb  <- ds$cubes[[nm]]
    res <- sub("grid", "", nm); res <- sub("km", " km", res)
    recs <- if (isTRUE(cb$resolves) && !is.na(cb$records)) format(cb$records, big.mark = ",") else "—"
    doi  <- if (isTRUE(cb$resolves)) sprintf("[%s](%s)", sub("https://doi.org/", "", cb$doi), cb$doi) else "—"
    ok   <- if (isTRUE(cb$resolves)) "✅" else "❌"
    cat(sprintf("| %s | %s | `%s` | %s | %s |\n", res, recs, cb$download_key, doi, ok))
  }
  cat(sprintf("\n*Resolved from download keys via the GBIF API on %s.*\n",
              format(as.Date(ds$resolved_at))))
}
```


### Files

```
data/SE/raw/cubes/
├── cube_10km.csv          # 49M rows, ~2.5 GB
└── cube_50km.csv          # 26M rows, ~1.3 GB
```

### Key differences from previous cube downloads

- **Single file per resolution** instead of 9 files per basis of record
- **Publisher and dataset** columns included (publishingorgkey, datasetkey)
- **All basis of record types** in one file (basisofrecord as a column)
- Downloaded via SQL API instead of pre-built cube DOIs

---

## 2. EEA Reference Grids (shared)

Both grids are Europe-wide and shared across all European countries. They live in `data/shared/grids/` and the pipeline clips to the country extent automatically.

### 2.1 EEA 10km Grid (Europe)

**Source:** [EEA GeoNetwork](https://sdi.eea.europa.eu/datastore/public?path=/eea_v_3035_10_km_eea-ref-grid-europe_p_2011_v01_r00)
**Version:** 2011
**CRS:** ETRS89-LAEA Europe (EPSG:3035)
**Coverage:** All of Europe
**Sweden cells:** 7,693 (after clipping)

```
data/shared/grids/
├── Grid_ETRS89-LAEA_10K.shp (+ .dbf, .prj, .shx)  # Europe-wide
```

### 2.2 EEA 50km Grid (Europe)

**Source:** [EEA GeoNetwork](https://sdi.eea.europa.eu/catalogue/srv/eng/catalog.search#/metadata/aac8379a-5c4e-445c-b2ef-23a6a2701ef0)
**Version:** v2024

```
data/shared/grids/
└── EEA_50km_grid_v2024.gpkg
```

---

## 3. Taxonomic Reference Datasets

### 3.1 Dyntaxa — Swedish Taxonomic Database

**Purpose:** National taxonomy backbone for Swedish taxa.
**DOI:** `r ck_doi("taxonomy")`
**GBIF Dataset:** `r ck_dataset("taxonomy")`
**Downloaded:** 2026-02-05
**Taxa:** ~110,000

```
data/SE/raw/taxonomy/
├── Taxon.csv                 # Taxonomic backbone
├── SpeciesDistribution.csv   # establishmentMeans, occurrenceStatus, threatStatus
├── VernacularName.csv        # Swedish and English common names
├── Reference.csv
├── meta.xml
└── eml.xml
```

Key columns in SpeciesDistribution.csv:

| Field | Values | Count |
|-------|--------|-------|
| establishmentMeans | native (62,394), introduced (4,755), invasive (450), naturalised (176), uncertain (2,819), blank (39,416) |
| occurrenceStatus | present (64,553), absent (9,892), doubtful (2,122), irregular (5,091), blank (28,352) |

### 3.2 Swedish Red List (SLU Artdatabanken)

**Purpose:** Threat status enrichment for the taxonomy backbone.
**DOI:** `r ck_doi("redlist")`
**GBIF Dataset:** `r ck_dataset("redlist")`
**Version:** 1.9
**Downloaded:** 2026-03-25
**Taxa:** ~11,000

```
data/SE/raw/redlist/
├── taxon.txt
├── distribution.txt
├── meta.xml
└── eml.xml
```

---

## 4. Invasive Species Registry

**Purpose:** Flag invasive alien species in the taxonomy backbone.
**DOI:** `r ck_doi("invasives")`
**GBIF Dataset:** `r ck_dataset("invasives")`
**Taxa:** ~450 invasive species

```
data/SE/raw/invasives/
├── taxon.txt
├── distribution.txt
├── meta.xml
└── eml.xml
```

Joined to the taxonomy backbone in script 03 as `is_invasive` (TRUE/FALSE) by `scientificName`.

---

## 5. Sensitive Species List (SLU Artdatabanken)

**Purpose:** Flag restricted access species with coordinate generalization categories.
**DOI:** `r ck_doi("sensitive")`
**GBIF Dataset:** `r ck_dataset("sensitive")`
**Published by:** SLU Artdatabanken
**Version:** 1.9 (March 2026)
**Taxa:** 203 restricted access species
**Last revised:** 2023

```
data/SE/raw/sensitive/
├── taxon.txt          # 203 species with sensitivity category in taxonRemarks
├── meta.xml
└── eml.xml
```

**Note:** No distribution file — the sensitivity category (5km, 25km, or 50km generalization)
is stored in the `taxonRemarks` column of `taxon.txt`. Script 03 detects this automatically.

Joined to the taxonomy backbone as `is_sensitive` (TRUE/FALSE) + `sensitivity_category` (e.g. "50km").

Taxonomic coverage: birds, mammals, Hymenoptera, butterflies, beetles, spiders,
vascular plants, lichens and fungi.

---

## 6. Administrative Boundaries (GADM)

**Purpose:** Region and municipality overlays for maps and regional filtering.
**Source:** [GADM v4.1](https://gadm.org/)
**Downloaded:** 2026-03-17

```
data/SE/raw/admin/
├── admin_level1.gpkg    # 21 regions (län)
└── admin_level2.gpkg    # 290 municipalities (kommuner)
```

Downloaded automatically via the `geodata` R package.

---

## Updating Data

**Red list / Invasive / Sensitive species update:**
1. Delete files in the relevant `data/SE/raw/{source}/` directory
2. Update the `dataset_key` in `configs/config_SE.yml` if the source moved to a new GBIF dataset
3. Run `source("scripts/01a_download_raw_data.R")`
4. Re-run pipeline from script 03 (DOIs in this document resolve automatically via `01b`)

**New occurrence cubes:**
1. Submit SQL query at https://www.gbif.org/occurrence/download/sql
2. Unzip downloaded file, rename to `cube_10km.csv` / `cube_50km.csv`
3. Place in `data/SE/raw/cubes/`
4. Delete old parquets in `data/SE/proc/cubes/`
5. Update the two `download_key` values in `configs/config_SE.yml` (the DOIs in this document resolve automatically from those keys via `01b`)
6. Re-run pipeline from script 04

**Taxonomy update:**
1. Delete files in `data/SE/raw/taxonomy/`
2. Run `source("scripts/01a_download_raw_data.R")`
3. Re-run full pipeline from script 03

---

**Last updated:** 2026-06-21
**Maintainer:** Lena Thöle
