# Data Sources — Norway

Documentation of all external datasets for the Norway gap finder pipeline.

> **Location:** All raw data in `data/NO/raw/`. Never edited manually.
> All processing outputs go to `data/NO/proc/`.

```{r ds-setup, include=FALSE}
# Citations are resolved + validated once by scripts/01b_resolve_data_sources.R
# and cached here. This document, the app, and the reports all read this same
# object, so DOIs never drift. If it's missing, run 01b first.
ds_meta_path <- here::here("data", "NO", "proc", "data_sources_meta.rds")
ds <- if (file.exists(ds_meta_path)) readRDS(ds_meta_path) else NULL

# Inline helpers so checklist citations are read from the resolved metadata
# (config download/dataset key -> GBIF API) instead of being hand-typed in prose.
ck_doi <- function(key) {
  x <- if (!is.null(ds)) ds$checklists[[key]] else NULL
  if (is.null(x) || is.null(x$doi) || is.na(x$doi)) return("\u2014")
  sprintf("[%s](%s)", sub("https://doi.org/", "", x$doi), x$doi)
}
ck_dataset <- function(key) {
  x <- if (!is.null(ds)) ds$checklists[[key]] else NULL
  if (is.null(x) || is.null(x$dataset_key) || is.na(x$dataset_key)) return("\u2014")
  sprintf("[%s](https://www.gbif.org/dataset/%s)", x$dataset_key, x$dataset_key)
}
```

---

## 1. GBIF Occurrence Cubes (SQL API)

**Purpose:** Species-level, gridded occurrence summaries for Norway.

**Source:** GBIF SQL Download API
**Downloaded:** 2026-04-17
**Format:** SQL_TSV_ZIP → CSV → Parquet (converted by pipeline)

### SQL Query

Both cubes use the same query, differing only in grid resolution (10000 vs 50000 meters):

```sql
SELECT
  specieskey, species, kingdom, phylum, class, "order", family,
  basisofrecord, publishingorgkey, datasetkey,
  GBIF_EEARGCode(10000, decimallatitude, decimallongitude, 0) AS eeacellcode,
  "year", "month",
  COUNT(*) AS occurrences
FROM occurrence
WHERE countrycode = 'NO'
  AND hascoordinate = TRUE
  AND hasgeospatialissues = FALSE
  AND occurrencestatus = 'PRESENT'
  AND specieskey IS NOT NULL
GROUP BY specieskey, species, kingdom, phylum, class, "order", family,
  basisofrecord, publishingorgkey, datasetkey,
  GBIF_EEARGCode(10000, decimallatitude, decimallongitude, 0),
  "year", "month"
```

### Columns (14)

| # | Column | Description |
|--:|--------|-------------|
| 1 | specieskey | GBIF species identifier (backbone key) |
| 2 | species | Scientific name (GBIF backbone) |
| 3 | kingdom | Kingdom name |
| 4 | phylum | Phylum name |
| 5 | class | Class name |
| 6 | order | Order name |
| 7 | family | Family name |
| 8 | basisofrecord | Record type (HUMAN_OBSERVATION, PRESERVED_SPECIMEN, etc.) |
| 9 | publishingorgkey | UUID of the publishing organisation |
| 10 | datasetkey | UUID of the source dataset |
| 11 | eeacellcode | EEA reference-grid cell code (10 km or 50 km) |
| 12 | year | Observation/collection year |
| 13 | month | Observation/collection month |
| 14 | occurrences | Record count for the species × cell × year × month |

### Downloads

```{r ds-cubes, echo=FALSE, results='asis'}
if (is.null(ds)) {
  cat("> *Run `scripts/01b_resolve_data_sources.R` to populate resolved DOIs,",
      "record counts and validation status.*\n")
} else {
  cat("| Resolution | Records | Download Key | DOI | Resolves |\n")
  cat("|-----------|--------:|--------------|-----|:--------:|\n")
  for (nm in names(ds$cubes)) {
    cb  <- ds$cubes[[nm]]
    res <- sub("grid", "", nm); res <- sub("km", " km", res)
    recs <- if (isTRUE(cb$resolves) && !is.na(cb$records)) format(cb$records, big.mark = ",") else "—"
    doi  <- if (isTRUE(cb$resolves)) sprintf("[%s](%s)", sub("https://doi.org/", "", cb$doi), cb$doi) else "—"
    ok   <- if (isTRUE(cb$resolves)) "✅" else "❌"
    cat(sprintf("| %s | %s | `%s` | %s | %s |\n", res, recs, cb$download_key, doi, ok))
  }
  cat(sprintf("\n*Resolved from download keys via the GBIF API on %s.*\n",
              format(as.Date(ds$resolved_at))))
}
```


### Files

```
data/NO/raw/cubes/
├── cube_10km.csv
└── cube_50km.csv
```

> **Note:** the Norway cubes were downloaded with an earlier cube SQL that also
> carried two publication-date columns (`year_published`, `month_published`).
> Those columns are now retired — the current 14-column SQL above omits them, and
> the pipeline ignores any extra columns, so these downloads still load unchanged.
> Re-download with the query above whenever you next refresh the cubes.

### Key differences from previous cube downloads

- **Single file per resolution** instead of 9 files per basis of record
- **Publisher and dataset** columns included (publishingorgkey, datasetkey)
- **All basis of record types** in one file (basisofrecord as a column)
- Downloaded via SQL API instead of pre-built cube DOIs

---

## 2. EEA Reference Grids (shared)

Both grids are Europe-wide and shared across all European countries. They live in `data/shared/grids/` and the pipeline clips to Norway's extent automatically.

### 2.1 EEA 10km Grid (Europe)

**Source:** [EEA GeoNetwork](https://sdi.eea.europa.eu/datastore/public?path=/eea_v_3035_10_km_eea-ref-grid-europe_p_2011_v01_r00)
**Version:** 2011
**CRS:** ETRS89-LAEA Europe (EPSG:3035)
**Coverage:** All of Europe (clipped to Norway at runtime)

```
data/shared/grids/
├── Grid_ETRS89-LAEA_10K.shp (+ .dbf, .prj, .shx)  # Europe-wide
```

### 2.2 EEA 50km Grid (Europe)

**Source:** [EEA GeoNetwork](https://sdi.eea.europa.eu/catalogue/srv/eng/catalog.search#/metadata/aac8379a-5c4e-445c-b2ef-23a6a2701ef0)
**Version:** v2024

```
data/shared/grids/
└── EEA_50km_grid_v2024.gpkg
```

---

## 3. Taxonomic Reference Datasets

### 3.1 Nortaxa — Norwegian Species Nomenclature Database

**Purpose:** National taxonomy backbone for Norwegian taxa.
**DOI:** `r ck_doi("taxonomy")`
**GBIF Dataset:** `r ck_dataset("taxonomy")`
**Published by:** Norwegian Biodiversity Information Centre (Artsdatabanken)
**Export URL:** https://ipt.artsdatabanken.no/archive.do?r=artsnavnebase

```
data/NO/raw/taxonomy/
├── taxon.txt
├── distribution.txt
├── meta.xml
└── eml.xml
```

### 3.2 Norwegian Red List 2015

**Purpose:** Threat status enrichment for the taxonomy backbone.
**DOI:** `r ck_doi("redlist")`
**GBIF Dataset:** `r ck_dataset("redlist")`
**Published by:** Norwegian Biodiversity Information Centre (Artsdatabanken)
**Export URL:** https://ipt.gbif.no/archive.do?r=redlist-norway

```
data/NO/raw/redlist/
├── taxon.txt
├── distribution.txt
├── meta.xml
└── eml.xml
```

**Note:** The 2021 Norwegian Red List exists (4,957 species assessed) but is not yet
published as a DwC-A on GBIF. The 2015 edition (21,402 species assessed) is used here.
Swap the `dataset_key` in `config_NO.yml` once the 2021 edition is available — the DOI
and citation above resolve automatically from the new key via `01b`.

---

## 4. Invasive Species Registry

### GRIIS — Global Register of Introduced and Invasive Species - Norway

**Purpose:** Flag invasive alien species in the taxonomy backbone.
**DOI:** `r ck_doi("invasives")`
**GBIF Dataset:** `r ck_dataset("invasives")`
**Published by:** Invasive Species Specialist Group (ISSG)
**Export URL:** https://cloud.gbif.org/griis/archive.do?r=griis-norway

```
data/NO/raw/invasives/
├── taxon.txt
├── distribution.txt
├── speciesprofile.txt   # isInvasive flag (Invasive / Null)
├── meta.xml
└── eml.xml
```

This is the GRIIS international checklist for Norway, not a national Norwegian
registry. It contains validated records of introduced (alien) and invasive alien
species. The `isInvasive` field (in `speciesprofile.txt`) marks species with
documented negative impact on biodiversity. Script 03 joins it to the taxonomy
backbone as `is_invasive` (TRUE/FALSE), flagging a species when any of its listed
forms appears as invasive.

> **Note:** Norway's national alien species assessment (Fremmedartslista) is
> maintained by Artsdatabanken but is not currently published as a DwC-A on GBIF.
> The GRIIS checklist is the best available alternative.

---

## 5. Sensitive Species List

**DOI:** —
**Status:** ❌ Not available

No Norwegian sensitive/restricted species DwC-A checklist was found on GBIF.
The pipeline runs without it — `is_sensitive = FALSE` for all taxa, and the
sensitive scope is automatically skipped by script 09c.

Set `sensitive.enabled: true` and add a `dataset_key` in `config_NO.yml` when a
list becomes available. Search for future availability:
https://www.gbif.org/dataset/search?q=sensitive&publishingCountry=NO&type=CHECKLIST

---

## 6. Administrative Boundaries (GADM)

**Purpose:** Region and municipality overlays for maps and regional filtering.
**Source:** [GADM v4.1](https://gadm.org/)

```
data/NO/raw/admin/
├── admin_level1.gpkg    # Fylker (counties)
└── admin_level2.gpkg    # Kommuner (municipalities)
```

Downloaded automatically via the `geodata` R package.

---

## Updating Data

**Red list / Invasive species update:**
1. Delete files in the relevant `data/NO/raw/{source}/` directory
2. Update the `dataset_key` in `configs/config_NO.yml` if the source moved to a new GBIF dataset
3. Run `source("scripts/01a_download_raw_data.R")`
4. Re-run pipeline from script 03 (DOIs in this document resolve automatically via `01b`)

**New occurrence cubes:**
1. Submit the SQL query above at https://www.gbif.org/occurrence/download/sql with `countrycode = 'NO'`
2. Unzip downloaded file, rename to `cube_10km.csv` / `cube_50km.csv`
3. Place in `data/NO/raw/cubes/`
4. Delete old parquets in `data/NO/proc/cubes/`
5. Update the two `download_key` values in `configs/config_NO.yml` (the DOIs in this document resolve automatically from those keys via `01b`)
6. Re-run pipeline from script 04

**Taxonomy update:**
1. Delete files in `data/NO/raw/taxonomy/`
2. Run `source("scripts/01a_download_raw_data.R")`
3. Re-run full pipeline from script 03

---

**Last updated:** 2026-06-21
**Maintainer:** Lena Thöle
