---
title: "Derived Summaries Overview"
subtitle: "GBIF Gap Finder - Data Structure and Usage Guide"
author: "Generated by 06a_make_core_summaries.R + 06b_make_species_summaries.R"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(scales)
library(data.table)
library(DT)
library(glue)

# Paths
p_derived <- here("data_proc", "derived")
p_by_order <- here(p_derived, "by_order")

# Theme for plots
theme_set(theme_minimal(base_size = 12))
```

# Introduction

This document provides an overview of the derived summary tables created by `06a_make_core_summaries.R` and `06b_make_species_summaries.R`. These tables aggregate GBIF occurrence cube data into analysis-ready formats for gap analysis.

**Note on architecture (April 2026 refactor):** 06a and 06b now produce only
taxonomy-agnostic aggregations. Scope-filtered variants (`_all`, `_dyntaxa`,
`_threatened`, `_invasive`, `_sensitive`) and the recent-period layer are
produced by a dedicated downstream script, `09c_scope_summaries.R`, using the
reconciliation table from 09a. See the CHANGELOG for details.

## Key Features

- **Pure cube aggregations**: 06a produces one variant of each core summary (cell, time, cell × time, order × cell, order × time, family × time, published × time) with no taxonomy filtering. Scope-filtered variants live in files named `<summary>_<scope>_<grid>.csv` (produced by 09c).
- **Bias correction columns**: Species-level tables from 06b include `familycount`, `ordercount`, `classcount` and their corresponding `relative_*` ratios.
- **Split by order**: Species-level tables are split by taxonomic order for manageable file sizes.
- **Multiple granularities**: From simple cell counts to full species × cell × time detail.
- **Taxonomy flags applied downstream**: `in_dyntaxa`, `is_invasive`, `is_sensitive`, and scope-filtered aggregations are all produced by 09c, which uses the 4-tier matching from 09a (not the simple name join that 06a/06b previously relied on).

---

# File Inventory

```{r file_inventory}
# Core summaries
core_files <- list.files(p_derived, pattern = "^(cell|time)_summary.*\\.csv$", full.names = TRUE)
cell_time_files <- list.files(p_derived, pattern = "^cell_time_summary.*\\.csv$", full.names = TRUE)
order_level_files <- list.files(p_derived, pattern = "^(order|family).*summary.*\\.csv$", full.names = TRUE)

# Species by order
species_summary_files <- list.files(here(p_by_order, "species_summary"), pattern = "\\.csv$", full.names = TRUE)
species_cell_files <- list.files(here(p_by_order, "species_cell"), pattern = "\\.csv$", full.names = TRUE)
species_time_files <- list.files(here(p_by_order, "species_time"), pattern = "\\.csv$", full.names = TRUE)
species_cell_time_files <- list.files(here(p_by_order, "species_cell_time"), pattern = "\\.csv$", full.names = TRUE)

# Create inventory table
inventory <- tibble::tribble(

~Category, ~Description, ~Files, ~Location,
"Cell Summary", "Occurrences & species richness per cell", length(core_files[str_detect(core_files, "cell_summary")]), "derived/",
"Time Summary", "Occurrences & species richness per month", length(core_files[str_detect(core_files, "time_summary")]), "derived/",
"Cell × Time", "Occurrences per cell per month", length(cell_time_files), "derived/",
"Order × Cell", "Occurrences per order per cell", length(order_level_files[str_detect(order_level_files, "order_cell")]), "derived/",
"Order × Time", "Occurrences per order per month", length(order_level_files[str_detect(order_level_files, "order_time")]), "derived/",
"Family × Time", "Occurrences per family per month", length(order_level_files[str_detect(order_level_files, "family_time")]), "derived/",
"Species Summary", "Totals per species (by order)", length(species_summary_files), "derived/by_order/species_summary/",
"Species × Cell", "Species spatial distribution (by order)", length(species_cell_files), "derived/by_order/species_cell/",
"Species × Time", "Species temporal trends (by order)", length(species_time_files), "derived/by_order/species_time/",
"Species × Cell × Time", "Full species detail (by order)", length(species_cell_time_files), "derived/by_order/species_cell_time/"
)

knitr::kable(inventory, caption = "Summary of derived tables")
```

---
# Table Structures

## Core Summaries

### Cell Summary

Aggregated occurrence counts per grid cell.

```{r cell_summary_structure}
if (length(core_files[str_detect(core_files, "cell_summary")]) > 0) {
  cell_file <- core_files[str_detect(core_files, "cell_summary")][1]
  cell_sample <- fread(cell_file, nrows = 5)
  
  cat("**File:**", basename(cell_file), "\n\n")
  cat("**Columns:**", paste(names(cell_sample), collapse = ", "), "\n\n")
  
  knitr::kable(cell_sample, caption = "Sample rows from cell_summary")
} else {
  cat("No cell summary files found.\n")
}
```

| Column | Description |
|--------|-------------|
| `grid` | Grid resolution (grid10km, grid50km) |
| `basisofrecord` | Record type or "all" for combined |
| `eeacellcode` | EEA grid cell identifier |
| `occurrences` | Total occurrence count |
| `n_species` | Number of unique species |

**Use cases:**
- Identify zero-coverage cells
- Map sampling intensity
- Compare basisOfRecord spatial patterns

---

### Time Summary

Aggregated occurrence counts per year-month.

```{r time_summary_structure}
if (length(core_files[str_detect(core_files, "time_summary")]) > 0) {
  time_file <- core_files[str_detect(core_files, "^(?!cell_).*time_summary")][1]
  if (!is.na(time_file)) {
    time_sample <- fread(time_file, nrows = 5)
    
    cat("**File:**", basename(time_file), "\n\n")
    knitr::kable(time_sample, caption = "Sample rows from time_summary")
  }
}
```

| Column | Description |
|--------|-------------|
| `grid` | Grid resolution |
| `basisofrecord` | Record type or "all" |
| `yearmonth` | Year-month (YYYY-MM or YYYYMM) |
| `occurrences` | Total occurrence count |
| `n_species` | Number of unique species |
| `n_cells` | Number of cells with data |

**Use cases:**
- Identify temporal gaps
- Analyze seasonal patterns
- Track data growth over time

---

### Cell × Time Summary

Occurrences per cell per month — for data recency analysis.

```{r cell_time_structure}
if (length(cell_time_files) > 0) {
  ct_sample <- fread(cell_time_files[1], nrows = 5)
  
  cat("**File:**", basename(cell_time_files[1]), "\n\n")
  knitr::kable(ct_sample, caption = "Sample rows from cell_time_summary")
}
```

| Column | Description |
|--------|-------------|
| `eeacellcode` | Grid cell |
| `yearmonth` | Time period |
| `occurrences` | Occurrence count |
| `n_species` | Species in this cell-month |

**Use cases:**
- Calculate data recency per cell (when was last observation?)
- Identify cells with seasonal gaps
- Temporal completeness analysis

---

## Order-Level Summaries

### Order × Cell Summary

```{r order_cell_structure}
order_cell_files <- list.files(p_derived, pattern = "order_cell_summary", full.names = TRUE)
if (length(order_cell_files) > 0) {
  oc_sample <- fread(order_cell_files[1], nrows = 5)
  knitr::kable(oc_sample, caption = "Sample rows from order_cell_summary")
}
```

**Use cases:**
- Compare spatial coverage across taxonomic orders
- Identify which orders are under-sampled in specific regions

### Order × Time Summary

```{r order_time_structure}
order_time_files <- list.files(p_derived, pattern = "order_time_summary", full.names = TRUE)
if (length(order_time_files) > 0) {
  ot_sample <- fread(order_time_files[1], nrows = 5)
  knitr::kable(ot_sample, caption = "Sample rows from order_time_summary")
}
```

**Use cases:**
- Compare temporal trends across orders
- Identify orders with declining sampling

---

## Species-Level Summaries (by Order)

All species-level tables are split by taxonomic order and include **bias correction columns**.

### Available Orders

```{r list_orders}
# Extract order names from filenames
if (length(species_summary_files) > 0) {
  orders_available <- species_summary_files |>
    basename() |>
    str_extract("species_summary_(.+)_\\d+km\\.csv", group = 1) |>
    unique() |>
    sort()
  
  cat("**Orders with species-level data:**\n\n")
  cat(paste("-", orders_available), sep = "\n")
} else {
  cat("No species summary files found.\n")
}
```

### Species Summary (by Order)

Totals per species within each order.

```{r species_summary_structure}
if (length(species_summary_files) > 0) {
  # Pick a common order like Passeriformes if available
  example_file <- species_summary_files[str_detect(species_summary_files, "Passeriformes|Coleoptera")][1]
  if (is.na(example_file)) example_file <- species_summary_files[1]
  
  ss_sample <- fread(example_file, nrows = 5)
  
  cat("**Example file:**", basename(example_file), "\n\n")
  knitr::kable(ss_sample, caption = "Sample rows from species_summary")
}
```

| Column | Description |
|--------|-------------|
| `specieskey` | GBIF species identifier |
| `species` | Scientific name |
| `family` | Family name |
| `occurrences` | Total occurrences |
| `familycount` | Total occurrences in family (for bias correction) |
| `ordercount` | Total occurrences in order |
| `classcount` | Total occurrences in class |
| `relative_family` | `occurrences / familycount` |
| `relative_order` | `occurrences / ordercount` |
| `relative_class` | `occurrences / classcount` |

Taxonomy flags (`in_dyntaxa`, `is_invasive`, `is_sensitive`, `establishmentMeans`) are not present on 06b output files. Join to `species_scope_summary.csv` (produced by 09c) via `specieskey` to add them:

```r
library(data.table)
ss <- fread("data/SE/proc/derived/by_order/species_summary/species_summary_Passeriformes_10km.csv")
scope <- fread("data/SE/proc/derived/species_scope_summary.csv")
ss <- merge(ss, scope[, .(specieskey, in_dyntaxa, is_invasive, is_sensitive,
                          is_threatened, establishmentMeans)],
            by = "specieskey", all.x = TRUE)
```

---

### Species × Cell (by Order)

Spatial distribution per species.

```{r species_cell_structure}
if (length(species_cell_files) > 0) {
  example_file <- species_cell_files[str_detect(species_cell_files, "Passeriformes|Coleoptera")][1]
  if (is.na(example_file)) example_file <- species_cell_files[1]
  
  sc_sample <- fread(example_file, nrows = 5)
  
  cat("**Example file:**", basename(example_file), "\n\n")
  knitr::kable(sc_sample, caption = "Sample rows from species_cell")
}
```

**Use cases:**
- Map individual species distributions
- Calculate species range sizes
- Identify spatial gaps for specific species

---

### Species × Time (by Order)

Temporal trends per species.

```{r species_time_structure}
if (length(species_time_files) > 0) {
  example_file <- species_time_files[str_detect(species_time_files, "Passeriformes|Coleoptera")][1]
  if (is.na(example_file)) example_file <- species_time_files[1]
  
  st_sample <- fread(example_file, nrows = 5)
  
  cat("**Example file:**", basename(example_file), "\n\n")
  knitr::kable(st_sample, caption = "Sample rows from species_time")
}
```

**Use cases:**
- Analyze species temporal trends
- Compare raw vs bias-corrected trends
- Identify species with declining records

---

### Species × Cell × Time (by Order)

Full detail — the most granular table.

```{r species_cell_time_structure}
if (length(species_cell_time_files) > 0) {
  example_file <- species_cell_time_files[str_detect(species_cell_time_files, "Passeriformes|Coleoptera")][1]
  if (is.na(example_file)) example_file <- species_cell_time_files[1]
  
  sct_sample <- fread(example_file, nrows = 5)
  
  cat("**Example file:**", basename(example_file), "\n\n")
  cat("**Columns:**", paste(names(sct_sample), collapse = ", "), "\n\n")
  
  # File size
  file_size_mb <- file.info(example_file)$size / 1024^2
  cat("**File size:**", round(file_size_mb, 1), "MB\n\n")
  
  knitr::kable(sct_sample, caption = "Sample rows from species_cell_time")
}
```

**Use cases:**
- Full single-species analysis
- Spatio-temporal modeling
- Occupancy modeling inputs

---

# Bias Correction

## The Problem

Raw occurrence counts are biased by sampling effort:

- Urban areas have more observers
- Weekends have more citizen science records
- Charismatic species are over-reported
- Some time periods have more digitization effort

## The Solution: Relative Occurrence

Instead of comparing raw counts, compare **what fraction of the higher taxon** each species represents:

$$\text{relative\_family} = \frac{\text{occurrences}}{\text{familycount}}$$

This normalizes for sampling effort because observers who record one species in a family typically record others too.

## Example: Comparing Two Cells

```{r bias_example}
example_data <- tibble::tribble(
  ~cell, ~species, ~occurrences, ~familycount, ~relative_family,
  "Cell A (urban)", "Parus major", 500, 2000, 0.25,
  "Cell B (rural)", "Parus major", 50, 200, 0.25,
  "Cell A (urban)", "Parus caeruleus", 300, 2000, 0.15,
  "Cell B (rural)", "Parus caeruleus", 40, 200, 0.20
)

knitr::kable(example_data, caption = "Raw vs relative occurrence comparison")
```
**Interpretation:**

- Raw counts suggest Parus major is 10× more common in Cell A
- Relative occurrence shows it's equally dominant (25%) in both cells
- Parus caeruleus is actually *relatively* more common in Cell B (20% vs 15%)

---

# Quick Analysis Examples

## Example 1: Load One Order

```{r example_load, eval=FALSE}
library(data.table)

# Load all Passeriformes spatial data
birds <- fread("data_proc/derived/by_order/species_cell/species_cell_Passeriformes_10km.csv")

# Filter to one species
great_tit <- birds[species == "Parus major"]

# Summarize
great_tit[, .(
  total_occurrences = sum(occurrences),
  n_cells = .N,
  mean_relative_family = mean(relative_family, na.rm = TRUE)
), by = basisofrecord]
```

## Example 2: Compare Raw vs Corrected Trends

```{r example_trends, eval=FALSE}
# Load temporal data
birds_time <- fread("data_proc/derived/by_order/species_time/species_time_Passeriformes_10km.csv")

# Get one species, all basisOfRecord combined
gt_trends <- birds_time[species == "Parus major" & basisofrecord == "all"]

# Extract year
gt_trends[, year := as.integer(substr(yearmonth, 1, 4))]

# Compare raw vs corrected annual trends
annual <- gt_trends[, .(
  raw_total = sum(occurrences),
  corrected_mean = mean(relative_family, na.rm = TRUE)
), by = year]

# Plot both
library(ggplot2)
ggplot(annual, aes(x = year)) +
  geom_line(aes(y = raw_total, color = "Raw occurrences")) +
  geom_line(aes(y = corrected_mean * max(raw_total), color = "Relative (scaled)")) +
  labs(title = "Parus major: Raw vs Bias-Corrected Trend")
```

## Example 3: Identify Under-sampled Species

```{r example_undersampled, eval=FALSE}
# Load species summaries for an order
mammals <- fread("data_proc/derived/by_order/species_summary/species_summary_Rodentia_10km.csv")

# Find species with low relative occurrence (potentially under-sampled)
undersampled <- mammals[
  basisofrecord == "all" & 
  !is.na(relative_order) & 
  occurrences >= 10  # At least some records
][order(relative_order)][1:20]

print(undersampled[, .(species, occurrences, ordercount, relative_order)])
```

---

# File Size Summary

```{r file_sizes}
# Calculate file sizes
all_files <- c(
  list.files(p_derived, pattern = "\\.csv$", full.names = TRUE),
  list.files(p_by_order, pattern = "\\.csv$", full.names = TRUE, recursive = TRUE)
)

if (length(all_files) > 0) {
  file_info <- data.table(
    path = all_files,
    size_mb = file.info(all_files)$size / 1024^2
  )
  file_info[, category := fcase(
    str_detect(path, "species_cell_time"), "Species × Cell × Time",
    str_detect(path, "species_cell"), "Species × Cell",
    str_detect(path, "species_time"), "Species × Time",
    str_detect(path, "species_summary"), "Species Summary",
    str_detect(path, "cell_time"), "Cell × Time",
    str_detect(path, "order_|family_"), "Order/Family Level",
    default = "Core"
  )]
  
  size_summary <- file_info[, .(
    n_files = .N,
    total_mb = sum(size_mb),
    avg_mb = mean(size_mb),
    max_mb = max(size_mb)
  ), by = category][order(-total_mb)]
  
  knitr::kable(size_summary, digits = 1, caption = "File sizes by category")
  
  cat("\n**Total size:**", round(sum(file_info$size_mb), 1), "MB\n")
}
```

---

# Next Steps

1. **Validate data**: Run `05_validate_inputs.R` to check for issues
2. **Match to Dyntaxa**: Join GBIF species to your taxonomic backbone
3. **Gap analysis**: Use these tables for spatial, temporal, and taxonomic gap analysis
4. **Visualization**: Create maps and charts from the summaries

---

# Session Info

```{r session_info}
sessionInfo()
```
