---
title: "GBIF Gap Finder — Overview"
subtitle: "Integrated Summary of Spatial, Temporal, Taxonomic, and Institutional Gaps"
author: "GBIF Gap Finder"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: hide
    theme: flatly
  pdf_document:
    toc: true
    latex_engine: xelatex
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6, dpi = 150
)

library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(sf)
library(knitr)
library(kableExtra)

# Load the same data bundle as the Shiny app
data_path <- here("shiny_app", "gap_finder", "data", "shiny_data.rds")
app_data <- readRDS(data_path)
safe_get <- function(name) if (name %in% names(app_data)) app_data[[name]] else NULL

# Extract common datasets
dashboard       <- safe_get("dashboard")
grid_10km       <- safe_get("grid_10km")
spatial_gaps    <- safe_get("spatial_gaps_10km")
tax_by_threat   <- safe_get("tax_by_threat")
tax_by_order    <- safe_get("tax_by_order")
overview_ly     <- safe_get("overview_last_year")
metadata        <- safe_get("metadata")
match_summary   <- safe_get("taxonomic_match_summary")
species_scope   <- safe_get("species_scope_lookup")
time_summary    <- safe_get("time_summary_10km")
all_basis_recent <- safe_get("all_basis_recent")
publisher_summary <- safe_get("publisher_summary")
publisher_cell_dep <- safe_get("publisher_cell_dependency")
last_year_label <- safe_get("recent_label") %||% "last 12 months"

# Normalise: script 10 uses n_ref_total, script 11 uses n_taxa
norm_tax <- function(df) {
  if (is.null(df)) return(NULL)
  if ("n_ref_total" %in% names(df) && !"n_taxa" %in% names(df))
    df <- df |> rename(n_taxa = n_ref_total)
  df
}
tax_by_threat <- norm_tax(tax_by_threat)

# ---- Sources of truth: computed exactly as the Shiny app's truth_* reactives
# and ov_concern, so this report ties out to the dashboard. Spatial from
# spatial_gaps[basisofrecord == "all"]; taxonomic + concern from the match
# summary (matched_any = in GBIF); year range from time_summary.
truth_spatial <- if (!is.null(spatial_gaps)) {
  sf <- spatial_gaps |> filter(basisofrecord == "all")
  total <- nrow(sf); with_data <- sum(sf$has_data, na.rm = TRUE)
  list(total_cells = total, cells_data = with_data, cells_zero = total - with_data,
       pct_coverage = if (total > 0) round(100 * mean(sf$has_data, na.rm = TRUE), 1) else 0,
       total_occ = sum(as.numeric(sf$occurrences), na.rm = TRUE))
} else NULL

truth_taxonomic <- if (!is.null(match_summary)) {
  n_ref <- nrow(match_summary); n_gbif <- sum(match_summary$matched_any, na.rm = TRUE)
  list(n_reference = n_ref, n_in_gbif = n_gbif, n_missing = n_ref - n_gbif,
       pct_coverage = if (n_ref > 0) round(100 * n_gbif / n_ref, 1) else 0)
} else NULL

truth_years <- if (!is.null(time_summary)) {
  yrs <- time_summary$year[!is.na(time_summary$year)]
  list(min = if (length(yrs)) min(yrs) else NA, max = if (length(yrs)) max(yrs) else NA)
} else NULL

concern_stats <- if (!is.null(match_summary)) {
  tc  <- intersect(c("threatStatus","threatStatus_redlist","threatStatus_backbone"), names(match_summary))[1]
  thr <- if (!is.na(tc)) match_summary[[tc]] else rep(NA_character_, nrow(match_summary))
  matched <- if ("matched_any"  %in% names(match_summary)) match_summary$matched_any else rep(FALSE, nrow(match_summary))
  inv <- if ("is_invasive"  %in% names(match_summary)) match_summary$is_invasive  else rep(FALSE, nrow(match_summary))
  sen <- if ("is_sensitive" %in% names(match_summary)) match_summary$is_sensitive else rep(FALSE, nrow(match_summary))
  is_thr <- thr %in% c("CR","EN","VU","NT")
  list(thr_total = sum(is_thr, na.rm = TRUE), thr_gbif = sum(is_thr & matched, na.rm = TRUE),
       inv_total = sum(inv, na.rm = TRUE),    inv_gbif = sum(inv & matched, na.rm = TRUE),
       sen_total = sum(sen, na.rm = TRUE),    sen_gbif = sum(sen & matched, na.rm = TRUE),
       unmon_cr  = sum(thr == "CR" & !matched, na.rm = TRUE),
       unmon_en  = sum(thr == "EN" & !matched, na.rm = TRUE),
       unmon_inv = sum(inv & !matched, na.rm = TRUE))
} else NULL

# Theme
pal <- list(sage = "#2A7F62", slate = "#4477AA", sand = "#CCBB44",
            coral = "#EE6677", plum = "#AA3377")

theme_gaps <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "#666", size = 11),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
theme_set(theme_gaps)
```

# About

This report summarises the GBIF biodiversity data gap analysis for **`r metadata$taxonomy_name %||% "the configured country"`**, using the national taxonomy backbone as a reference. It synthesises findings across spatial, temporal, taxonomic, and institutional dimensions.

**Data sources:**

- **Taxonomy backbone:** `r metadata$taxonomy_name %||% "National checklist"`
- **Grid resolution:** EEA 10 km reference grid (EPSG:3035)
- **Analysis date:** `r if (!is.null(metadata$created_at)) format(metadata$created_at, "%Y-%m-%d") else "unknown"`

# Key Statistics

```{r key-stats}
if (!is.null(truth_spatial) && !is.null(truth_taxonomic)) {
  stats <- tibble(
    Metric = c("Total occurrences", "Species in GBIF", "Species in backbone",
               "Taxonomic coverage", "Grid cells (10 km)",
               "Cells with data", "Spatial coverage",
               "Year range"),
    Value = c(
      comma(truth_spatial$total_occ),
      comma(truth_taxonomic$n_in_gbif),
      comma(truth_taxonomic$n_reference),
      paste0(truth_taxonomic$pct_coverage, "%"),
      comma(truth_spatial$total_cells),
      comma(truth_spatial$cells_data),
      paste0(truth_spatial$pct_coverage, "%"),
      if (!is.null(truth_years)) paste0(truth_years$min, "–", truth_years$max) else "—"
    )
  )
  kable(stats, align = c("l", "r")) |> kable_styling(full_width = FALSE)
}
```

# Gap Summary

```{r gap-panels, fig.height=4}
if (!is.null(truth_spatial) && !is.null(truth_taxonomic)) {
  gap_df <- tibble(
    Dimension = c("Spatial (10 km)", "Taxonomic"),
    Coverage = c(truth_spatial$pct_coverage, truth_taxonomic$pct_coverage),
    Gap = 100 - Coverage
  ) |>
    pivot_longer(c(Coverage, Gap), names_to = "type", values_to = "pct")

  ggplot(gap_df, aes(x = pct, y = Dimension, fill = type)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = c(Coverage = pal$sage, Gap = pal$coral)) +
    scale_x_continuous(labels = label_percent(scale = 1), limits = c(0, 100)) +
    labs(title = "Coverage vs Gap", x = "Percentage", y = "", fill = "") +
    theme(legend.position = "bottom")
}
```

## Threatened Species

```{r threatened-summary}
if (!is.null(tax_by_threat) && "threatStatus" %in% names(tax_by_threat)) {
  df <- tax_by_threat |> filter(threatStatus %in% c("CR", "EN", "VU", "NT", "DD"))
  kable(df |> select(Status = threatStatus, `In backbone` = n_taxa,
    `In GBIF` = n_in_gbif, Missing = n_missing, `Coverage %` = pct_coverage),
    align = c("l", "r", "r", "r", "r")) |> kable_styling(full_width = FALSE)
}
```

## Species of Concern

```{r concern-summary}
if (!is.null(concern_stats)) {
  cs <- concern_stats
  concern <- tibble(
    Category       = c("Threatened (CR/EN/VU/NT)", "Invasive", "Sensitive"),
    `In backbone`  = c(cs$thr_total, cs$inv_total, cs$sen_total),
    `In GBIF`      = c(cs$thr_gbif,  cs$inv_gbif,  cs$sen_gbif),
    Unmonitored    = c(cs$thr_total - cs$thr_gbif, cs$inv_total - cs$inv_gbif, cs$sen_total - cs$sen_gbif)
  )
  kable(concern, align = c("l", "r", "r", "r")) |> kable_styling(full_width = FALSE)
}
```

Of the threatened species with **no GBIF records at all**, `r if (!is.null(concern_stats)) concern_stats$unmon_cr else 0` are Critically Endangered (CR) and `r if (!is.null(concern_stats)) concern_stats$unmon_en else 0` Endangered (EN); `r if (!is.null(concern_stats)) concern_stats$unmon_inv else 0` invasive species are entirely unmonitored.

## Publishers

```{r publisher-summary}
if (!is.null(publisher_summary)) {
  n_pub <- nrow(publisher_summary)
  n_single <- if (!is.null(publisher_cell_dep)) sum(publisher_cell_dep$n_publishers == 1) else NA
  n_cells <- if (!is.null(grid_10km)) nrow(grid_10km) else NA
  top_pub <- publisher_summary |> arrange(desc(total_occurrences)) |> slice(1)
  top_pct <- round(100 * top_pub$total_occurrences / sum(publisher_summary$total_occurrences), 1)

  pub_stats <- tibble(
    Metric = c("Total publishers", "Single-publisher cells", "Largest publisher", "Top publisher share"),
    Value = c(comma(n_pub),
              paste0(comma(n_single), " / ", comma(n_cells)),
              if ("publisher_name" %in% names(top_pub)) top_pub$publisher_name else top_pub$publishingorgkey,
              paste0(top_pct, "%"))
  )
  kable(pub_stats, align = c("l", "r")) |> kable_styling(full_width = FALSE)
}
```

## Basis of Record

```{r bor-summary, fig.height=4}
bor_src <- if (!is.null(all_basis_recent)) {
  all_basis_recent |> filter(basisofrecord != "all") |>
    transmute(basisofrecord, occurrences = as.numeric(occ_total))
} else if (!is.null(spatial_gaps)) {
  spatial_gaps |> filter(basisofrecord != "all") |>
    group_by(basisofrecord) |>
    summarise(occurrences = sum(as.numeric(occurrences), na.rm = TRUE), .groups = "drop")
} else NULL
if (!is.null(bor_src)) {
  sg <- bor_src |>
    mutate(pct = round(100 * occurrences / sum(occurrences), 1),
           bor_label = gsub("_", " ", basisofrecord) |> tools::toTitleCase()) |>
    arrange(desc(occurrences))

  ggplot(sg |> head(6), aes(x = reorder(bor_label, occurrences), y = occurrences)) +
    geom_col(fill = pal$sage) +
    geom_text(aes(label = paste0(pct, "%")), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.15))) +
    labs(title = "Occurrences by Basis of Record", x = "", y = "Total occurrences")
}
```

# The Last 12 Months in Review

```{r last-12-months}
if (!is.null(overview_ly)) {
  ly_stats <- tibble(
    Metric = c("Observations dated in period",
               "Cells with active recording",
               "Newly covered cells",
               "Priority cells resolved"),
    Value = c(
      comma(overview_ly$occ_last_year),
      comma(overview_ly$cells_active_last_year),
      comma(overview_ly$cells_newly_covered),
      comma(overview_ly$cells_resolved %||% 0)
    )
  )
  kable(ly_stats, caption = paste0("Period: ", last_year_label),
    align = c("l", "r")) |> kable_styling(full_width = FALSE)
}
```

## Establishment Means

```{r establishment, fig.height=4}
tax_estab <- norm_tax(safe_get("tax_by_establishment"))
if (!is.null(tax_estab)) {
  df <- tax_estab |>
    mutate(label = case_when(
      is.na(establishmentMeans) | establishmentMeans == "" ~ "Unclassified",
      TRUE ~ tools::toTitleCase(establishmentMeans)
    )) |>
    arrange(desc(n_taxa))

  ggplot(df, aes(x = reorder(label, pct_coverage), y = pct_coverage)) +
    geom_col(aes(fill = label), show.legend = FALSE) +
    geom_text(aes(label = paste0(pct_coverage, "%")), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_fill_manual(values = c(Native = pal$sage, Introduced = pal$sand,
      Invasive = pal$coral, Naturalised = pal$slate,
      Uncertain = pal$plum, Unclassified = "#ccc")) +
    scale_y_continuous(limits = c(0, 105)) +
    labs(title = "GBIF Coverage by Establishment Means", x = "", y = "Coverage (%)")
}
```

---

*Report generated by `gbif_gap_finder` on `r Sys.time()`.*
