---
title: "Quick Sanity Checks - GBIF Sweden Data"
subtitle: "Visual and statistical QA after ingestion and derived summaries"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  error = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load project setup
library(here)
source(here("scripts", "00_setup.R"))

# Load packages
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(purrr)
library(sf)
library(viridis)
library(scales)
library(glue)
library(knitr)
library(forcats)
library(tibble)

# Set ggplot theme
theme_set(theme_minimal(base_size = 12))

# Paths
p_derived <- here(p_data_proc, "derived")
```

# Overview

This notebook provides visual and statistical quality checks for the GBIF Sweden occurrence data after ingestion and creation of derived summary tables.

**Run this after:** Scripts 01-06 (ingestion + derived summaries)

**Last run:** `r Sys.time()`

---

# 0. File Existence Checks

## Core processed files

```{r check-core-files}
core_files <- tribble(
  ~file, ~location, ~required,
  "grids_10km.gpkg", p_data_proc, TRUE,
  "grids_50km.gpkg", p_data_proc, TRUE,
  "taxa_reference_current.rds", p_data_proc, TRUE,
  "cube_manifest.csv", p_data_proc, TRUE,
  "cube_totals_by_basisOfRecord.csv", p_data_proc, TRUE
) |>
  mutate(
    path = here(location, file),
    exists = file.exists(path),
    status = case_when(
      exists ~ "✅",
      required ~ "❌ MISSING",
      TRUE ~ "⚠️ Optional"
    )
  )

core_files |>
  select(file, status) |>
  kable(caption = "Core Processed Files")
```

## Derived summary files

```{r check-derived-files}
derived_files <- tribble(
  ~file, ~required,
  "cell_summary_10km.csv", TRUE,
  "cell_summary_50km.csv", TRUE,
  "time_summary_10km.csv", TRUE,
  "time_summary_50km.csv", TRUE,
  "species_summary_10km.csv", TRUE,
  "species_summary_50km.csv", TRUE,
  "cube_key_summary.csv", TRUE,
  "grid_lookup_10km.csv", FALSE,
  "grid_lookup_50km.csv", FALSE,
  "family_time_summary_10km.csv", FALSE,
  "family_time_summary_50km.csv", FALSE,
  "order_time_summary_10km.csv", FALSE,
  "order_time_summary_50km.csv", FALSE,
  "species_time_summary_topN_10km.csv", FALSE,
  "species_time_summary_topN_50km.csv", FALSE
) |>
  mutate(
    path = here(p_derived, file),
    exists = file.exists(path),
    size_mb = map_dbl(path, ~{
      if (file.exists(.x)) file.size(.x) / 1024^2 else NA_real_
    }),
    status = case_when(
      exists ~ "✅",
      required ~ "❌ MISSING",
      TRUE ~ "⚠️ Optional"
    )
  )

derived_files |>
  select(file, status, size_mb) |>
  mutate(size_mb = round(size_mb, 2)) |>
  kable(caption = "Derived Summary Files")
```

---

# 1. Cube Totals by Basis of Record

```{r load-totals}
totals <- read_csv(
  here(p_data_proc, "cube_totals_by_basisOfRecord.csv"),
  show_col_types = FALSE
)

# Summary table
totals |>
  arrange(desc(total_occurrences)) |>
  mutate(
    total_occurrences = comma(total_occurrences),
    file_size_gb = round(file_size_gb, 2)
  ) |>
  kable(caption = "Occurrence Totals by Basis of Record and Grid")
```

## Total occurrences by basis of record

```{r plot-totals-all, fig.height=8}
totals |>
  mutate(basisOfRecord = fct_reorder(basisOfRecord, total_occurrences)) |>
  ggplot(aes(x = total_occurrences, y = basisOfRecord, fill = grid)) +
  geom_col(position = "dodge") +
  scale_x_continuous(labels = comma) +
  scale_fill_viridis_d(option = "C") +
  labs(
    title = "Total Occurrences by Basis of Record",
    subtitle = "Comparison between 10km and 50km grids",
    x = "Total Occurrences",
    y = NULL,
    fill = "Grid Resolution"
  ) +
  theme(legend.position = "bottom")
```

## Excluding human observations

Human observations often dominate; let's view without them:

```{r plot-totals-no-human, fig.height=7}
totals |>
  filter(basisOfRecord != "humanObservation") |>
  mutate(basisOfRecord = fct_reorder(basisOfRecord, total_occurrences)) |>
  ggplot(aes(x = total_occurrences, y = basisOfRecord, fill = grid)) +
  geom_col(position = "dodge") +
  scale_x_continuous(labels = comma) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Total Occurrences by Basis of Record (excluding humanObservation)",
    x = "Total Occurrences",
    y = NULL,
    fill = "Grid Resolution"
  ) +
  theme(legend.position = "bottom")
```

---

# 2. Spatial Coverage

## Top cells by occurrence count

```{r load-spatial}
cell10 <- read_csv(here(p_derived, "cell_summary_10km.csv"), show_col_types = FALSE)
cell50 <- read_csv(here(p_derived, "cell_summary_50km.csv"), show_col_types = FALSE)
```

### 10km grid - Top 20 cells

```{r top-cells-10km}
cell10 |>
  filter(basisofrecord == "all") |>
  arrange(desc(occurrences)) |>
  slice_head(n = 20) |>
  mutate(
    occurrences = comma(occurrences),
    rank = row_number()
  ) |>
  select(rank, eeacellcode, occurrences) |>
  kable(caption = "Top 20 Most Sampled Cells (10km)")
```

### 50km grid - Top 20 cells

```{r top-cells-50km}
cell50 |>
  filter(basisofrecord == "all") |>
  arrange(desc(occurrences)) |>
  slice_head(n = 20) |>
  mutate(
    occurrences = comma(occurrences),
    rank = row_number()
  ) |>
  select(rank, eeacellcode, occurrences) |>
  kable(caption = "Top 20 Most Sampled Cells (50km)")
```

## Coverage statistics

```{r spatial-coverage-stats}
spatial_stats <- bind_rows(
  cell10 |>
    filter(basisofrecord == "all") |>
    summarise(
      grid = "10km",
      total_cells = n(),
      cells_with_data = sum(occurrences > 0),
      total_occurrences = sum(occurrences),
      mean_occ_per_cell = mean(occurrences),
      median_occ_per_cell = median(occurrences),
      max_occ_per_cell = max(occurrences)
    ),
  cell50 |>
    filter(basisofrecord == "all") |>
    summarise(
      grid = "50km",
      total_cells = n(),
      cells_with_data = sum(occurrences > 0),
      total_occurrences = sum(occurrences),
      mean_occ_per_cell = mean(occurrences),
      median_occ_per_cell = median(occurrences),
      max_occ_per_cell = max(occurrences)
    )
) |>
  mutate(
    pct_coverage = round(100 * cells_with_data / total_cells, 1),
    across(c(total_occurrences, mean_occ_per_cell, median_occ_per_cell, max_occ_per_cell), 
           ~comma(round(.x)))
  )

spatial_stats |>
  kable(caption = "Spatial Coverage Statistics")
```

---

# 3. Temporal Patterns

```{r load-temporal}
time10 <- read_csv(here(p_derived, "time_summary_10km.csv"), show_col_types = FALSE) |>
  filter(basisofrecord == "all") |>
  mutate(
    yearmonth_chr = str_trim(as.character(yearmonth)),
    year = as.integer(str_sub(yearmonth_chr, 1, 4)),
    month = as.integer(str_sub(yearmonth_chr, 6, 7)),
    date = as.Date(glue("{yearmonth_chr}-01"))
  )
```

## Overall temporal trend

```{r temporal-trend}
time10 |>
  ggplot(aes(x = date, y = occurrences)) +
  geom_line(color = "steelblue") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Temporal Trend in Occurrence Records",
    subtitle = "10km grid, all basis of record",
    x = "Year-Month",
    y = "Occurrences"
  )
```

## Annual trend

```{r annual-trend}
time10 |>
  group_by(year) |>
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = occurrences)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  geom_point(color = "darkgreen") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Annual Occurrence Trend",
    subtitle = "10km grid, all basis of record",
    x = "Year",
    y = "Total Occurrences"
  )
```

## Seasonal pattern

```{r seasonal-pattern}
time10 |>
  group_by(month) |>
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = month, y = occurrences)) +
  geom_col(fill = "coral") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Seasonal Sampling Pattern",
    subtitle = "Aggregated across all years",
    x = "Month",
    y = "Total Occurrences"
  )
```

## Year × Month heatmap

```{r temporal-heatmap, fig.height=8}
time10 |>
  group_by(year, month) |>
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = month, y = year, fill = log10(occurrences + 1))) +
  geom_tile() +
  scale_fill_viridis(option = "inferno", name = "log10(occ+1)") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = "Sampling Intensity Over Time",
    subtitle = "Heatmap of occurrences by year and month",
    x = "Month",
    y = "Year"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Last 50 years (detailed view)

```{r temporal-heatmap-recent, fig.height=6}
time10 |>
  filter(year >= year(Sys.Date()) - 50) |>
  group_by(year, month) |>
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = month, y = year, fill = log10(occurrences + 1))) +
  geom_tile() +
  scale_fill_viridis(option = "magma", name = "log10(occ+1)") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = "Recent Sampling Intensity (Last 50 Years)",
    x = "Month",
    y = "Year"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 4. Taxonomic Patterns

## Top species overall

```{r load-species}
sp10 <- read_csv(here(p_derived, "species_summary_10km.csv"), show_col_types = FALSE) |>
  filter(basisofrecord == "all") |>
  arrange(desc(occurrences))
```

```{r top-species}
sp10 |>
  slice_head(n = 30) |>
  mutate(
    rank = row_number(),
    occurrences = comma(occurrences)
  ) |>
  select(rank, specieskey, species, occurrences) |>
  kable(caption = "Top 30 Species by Occurrence Count")
```

## Family-level patterns (if available)

```{r family-patterns}
fam10_path <- here(p_derived, "family_time_summary_10km.csv")

if (file.exists(fam10_path)) {
  fam10 <- read_csv(fam10_path, show_col_types = FALSE) |>
    filter(basisofrecord == "all") |>
    mutate(
      yearmonth_chr = str_trim(as.character(yearmonth)),
      year = as.integer(str_sub(yearmonth_chr, 1, 4)),
      month = as.integer(str_sub(yearmonth_chr, 6, 7))
    )
  
  # Top families
  fam10_top <- fam10 |>
    group_by(family) |>
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(occurrences)) |>
    slice_head(n = 20)
  
  fam10_top |>
    mutate(
      rank = row_number(),
      occurrences = comma(occurrences)
    ) |>
    select(rank, family, occurrences) |>
    kable(caption = "Top 20 Families by Occurrence Count")
  
  # Seasonality of top 10 families
  fam10_season <- fam10 |>
    semi_join(fam10_top |> slice_head(n = 10), by = "family") |>
    group_by(family, month) |>
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")
  
  fam10_season |>
    ggplot(aes(x = month, y = occurrences, color = family)) +
    geom_line(linewidth = 1) +
    scale_x_continuous(breaks = 1:12, labels = month.abb) +
    scale_y_continuous(labels = comma) +
    scale_color_viridis_d() +
    labs(
      title = "Seasonal Patterns of Top 10 Families",
      x = "Month",
      y = "Occurrences",
      color = "Family"
    ) +
    theme(
      legend.position = "right",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
} else {
  cat("Family-level summaries not available (optional file)")
}
```

## Order-level patterns (if available)

```{r order-patterns}
ord10_path <- here(p_derived, "order_time_summary_10km.csv")

if (file.exists(ord10_path)) {
  ord10 <- read_csv(ord10_path, show_col_types = FALSE) |>
    filter(basisofrecord == "all") |>
    mutate(
      yearmonth_chr = str_trim(as.character(yearmonth)),
      year = as.integer(str_sub(yearmonth_chr, 1, 4)),
      month = as.integer(str_sub(yearmonth_chr, 6, 7))
    )
  
  # Top orders
  ord10_top <- ord10 |>
    group_by(order) |>
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(occurrences)) |>
    slice_head(n = 20)
  
  ord10_top |>
    mutate(
      rank = row_number(),
      occurrences = comma(occurrences)
    ) |>
    select(rank, order, occurrences) |>
    kable(caption = "Top 20 Orders by Occurrence Count")
  
  # Temporal trend of top 8 orders
  ord10_year <- ord10 |>
    semi_join(ord10_top |> slice_head(n = 8), by = "order") |>
    group_by(order, year) |>
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")
  
  ord10_year |>
    ggplot(aes(x = year, y = occurrences, color = order)) +
    geom_line(linewidth = 1) +
    scale_y_continuous(labels = comma) +
    scale_color_viridis_d() +
    labs(
      title = "Temporal Trends of Top 8 Orders",
      x = "Year",
      y = "Occurrences",
      color = "Order"
    ) +
    theme(legend.position = "right")
} else {
  cat("Order-level summaries not available (optional file)")
}
```

---

# 5. Spatial Visualization

## Map: 10km grid coverage

```{r map-10km, fig.width=12, fig.height=10}
# Load grid
grid10 <- st_read(here(p_data_proc, "grids_10km.gpkg"), quiet = TRUE)

# Load cell summary
cell10_map <- cell10 |>
  filter(basisofrecord == "all") |>
  select(eeacellcode, occurrences)

# Identify cell code field
guess_cellcode_field <- function(field_names) {
  names_lower <- str_to_lower(field_names)
  
  # Prefer fields with both "eea" and "code"
  candidates <- field_names[
    str_detect(names_lower, "eea") & str_detect(names_lower, "code")
  ]
  
  if (length(candidates) > 0) return(candidates[1])
  
  # Fallback to any field with relevant keywords
  candidates <- field_names[str_detect(names_lower, "eea|cell|code|grid")]
  
  if (length(candidates) > 0) return(candidates[1])
  
  NA_character_
}

code_field <- guess_cellcode_field(names(grid10))

if (is.na(code_field)) {
  cat("❌ Could not identify cell code field in grid.\n")
  cat("Available fields:", paste(names(grid10), collapse = ", "))
} else {
  # Create join key
  grid10$eeacellcode <- as.character(grid10[[code_field]])
  
  # Join with occurrence data
  grid10_map <- grid10 |>
    left_join(cell10_map, by = "eeacellcode") |>
    mutate(
      occurrences = coalesce(occurrences, 0),
      log_occ = log10(occurrences + 1)
    )
  
  # Join diagnostics
  diag <- grid10_map |>
    st_drop_geometry() |>
    summarise(
      total_cells = n(),
      cells_with_data = sum(occurrences > 0),
      pct_coverage = round(100 * cells_with_data / total_cells, 1)
    )
  
  cat(glue("Grid cells: {comma(diag$total_cells)}"), "\n")
  cat(glue("Cells with data: {comma(diag$cells_with_data)} ({diag$pct_coverage}%)"), "\n\n")
  
  # Map
  ggplot(grid10_map) +
    geom_sf(aes(fill = log_occ), color = NA) +
    scale_fill_viridis(
      option = "plasma",
      name = "log10(occ+1)",
      guide = guide_colorbar(barwidth = 15, barheight = 0.5)
    ) +
    labs(
      title = "Spatial Coverage: Occurrence Density Across 10km Grid Cells",
      subtitle = glue("Cell code field: {code_field} | Coverage: {diag$pct_coverage}%")
    ) +
    theme_void() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 10)
    )
}
```

---

# 6. Data Quality Summary

```{r quality-summary}
quality_checks <- tribble(
  ~check, ~status, ~details,
  "Core files present", 
    ifelse(all(core_files$exists[core_files$required]), "✅", "❌"),
    glue("{sum(core_files$exists)} / {nrow(core_files)} files found"),
  
  "Derived summaries present",
    ifelse(all(derived_files$exists[derived_files$required]), "✅", "❌"),
    glue("{sum(derived_files$exists & derived_files$required)} / {sum(derived_files$required)} required files"),
  
  "Spatial coverage adequate",
    ifelse(spatial_stats$pct_coverage[1] > 50, "✅", "⚠️"),
    glue("{spatial_stats$pct_coverage[1]}% of 10km cells have data"),
  
  "Temporal data present",
    ifelse(nrow(time10) > 0, "✅", "❌"),
    glue("{comma(nrow(time10))} time periods with data"),
  
  "Species data present",
    ifelse(nrow(sp10) > 0, "✅", "❌"),
    glue("{comma(nrow(sp10))} unique species")
)

quality_checks |>
  kable(caption = "Data Quality Summary")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
