---
title: "Quick sanity checks after scripts 1-6"
output: html_document
date: "`r format(Sys.Date())`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,     # show code
  eval = TRUE,     # execute code
  error = TRUE,    # show errors (and keep knitting)
  warning = TRUE,  # show warnings
  message = TRUE,   # show messages (e.g., package startup)
  collapse = TRUE,  # show code + output together
  comment = "#>"    # prefix output like reprex
)

source(here::here("scripts", "00_setup.R"))

data_proc_rel <- cfg_get("paths.data_proc", "data_proc")
p_data_proc <- here::here(data_proc_rel)
p_derived <- file.path(p_data_proc, "derived")

library(dplyr)
library(ggplot2)
library(readr)
library(viridis)
library(stringr)
library(tidyr)
library(sf)

```

## 0) Check Phase 2 derived files exist

```{r}
derived_files <- c(
  "family_time_summary_10km.csv",
  "order_time_summary_10km.csv",
  "family_time_summary_50km.csv",
  "order_time_summary_50km.csv",
  "species_time_summary_topN_10km.csv",
  "species_time_summary_topN_50km.csv"
)

exists_tbl <- tibble::tibble(
  file = derived_files,
  exists = file.exists(file.path(p_derived, derived_files))
)

exists_tbl
```

## 1) Cube totals by basisOfRecord

```{r}
totals <- read_csv(file.path(p_data_proc, "cube_totals_by_basisOfRecord.csv"), show_col_types = FALSE)

totals
totals %>%
  arrange(desc(total_occurrences)) %>%
  print(n = 50)
```

### total occurrences by basisOfRecord
```{r}
ggplot(totals, aes(x = basisOfRecord, y = total_occurrences, fill = basisOfRecord)) +
  geom_col() +
  #coord_flip() +
  facet_wrap(~ grid, scales = "free_x") +
  scale_fill_viridis_d() +
  labs(
    title = "Total occurrences by basisOfRecord (per grid)",
    x = "basisOfRecord",
    y = "Total occurrences"
  )

```


### total occurrences by basisOfRecord without humanObservation
```{r}

totals_noHumObs <- totals %>%
  filter(basisOfRecord != "humanObservation")

ggplot(totals_noHumObs, aes(x = basisOfRecord, y = total_occurrences, fill = basisOfRecord)) +
  geom_col() +
  #coord_flip() +
  facet_wrap(~ grid, scales = "free_x") +
  scale_fill_viridis_d(option = "B") +
  labs(
    title = "Total occurrences by basisOfRecord (per grid)",
    x = "basisOfRecord",
    y = "Total occurrences" )

```

---

## 2) Spatial sanity: top cells

```{r}
cell10 <- read_csv(file.path(p_derived, "cell_summary_10km.csv"), show_col_types = FALSE)
cell50 <- read_csv(file.path(p_derived, "cell_summary_50km.csv"), show_col_types = FALSE)

cell10 %>% 
  filter(basisofrecord == "all") %>% 
  arrange(desc(occurrences)) %>% 
  head(20)
```

```{r}
cell50 %>% 
  filter(basisofrecord == "all") %>% 
  arrange(desc(occurrences)) %>% 
  head(20)
```

---

## 3) Temporal sanity: time trend

This section assumes that `yearmonth` is encoded as `YYYYMM` (e.g., `202312`).
If your format is different, adjust the parsing.

```{r}
time10 <- read_csv(file.path(p_derived, "time_summary_10km.csv"), show_col_types = FALSE) %>%
  filter(basisofrecord == "all")

time10 <- time10 %>%
  mutate(
    yearmonth_chr = stringr::str_trim(as.character(yearmonth)),
    year = as.integer(substr(yearmonth_chr, 1, 4)),
    month = as.integer(substr(yearmonth_chr, 6, 7)),
    ym = as.Date(paste0(yearmonth_chr, "-01"))
  )


ggplot(time10, aes(x = ym, y = occurrences)) +
  geom_line() +
  labs(x = "Year-Month", y = "Occurrences")

```
### 3.1 Plot by Year

```{r}
time10_year <- time10 %>%
  group_by(year) %>%
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")

ggplot(time10_year, aes(x = year, y = occurrences)) +
  geom_line() +
  labs(x = "Year", y = "Occurrences", title = "Occurrences per year (10km, all)")
```
### 3.2 Plot by Month

```{r}
time10_month <- time10 %>%
  group_by(month) %>%
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")

ggplot(time10_month, aes(x = month, y = occurrences)) +
  geom_line() +
  scale_x_continuous(breaks = 1:12, limits = c(1, 12)) +
  labs(x = "Month", y = "Occurrences", title = "Occurrences per year (10km, all)")
```

### 3.3 Heatmap

```{r}
time10_heat <- time10 %>%
  group_by(year, month) %>%
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")

ggplot(time10_heat, aes(x = month, y = year, fill = occurrences)) +
  geom_tile() +
  scale_fill_viridis() +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Year", title = "Sampling intensity by year and month")


time10_heat_last50yrs <- time10 %>%
  filter(year > 1974) %>%
  group_by(year, month) %>%
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")

ggplot(time10_heat_last50yrs, aes(x = month, y = year, fill = occurrences)) +
  geom_tile() +
  scale_fill_viridis() +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Year", title = "Sampling intensity by year and month")
```
---

## 4) Family & Order time summaries (10km) ----

```{r}
fam10_path <- file.path(p_derived, "family_time_summary_10km.csv")
ord10_path <- file.path(p_derived, "order_time_summary_10km.csv")

if (file.exists(fam10_path)) {
  fam10 <- readr::read_csv(fam10_path, show_col_types = FALSE) %>%
    filter(basisofrecord == "all") %>%
    mutate(
      yearmonth_chr = str_trim(as.character(yearmonth)),
      year = as.integer(substr(yearmonth_chr, 1, 4)),
      month = as.integer(substr(yearmonth_chr, 6, 7))
    )

  # Top families overall
  fam10_top <- fam10 %>%
    group_by(family) %>%
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(occurrences)) %>%
    slice_head(n = 20)

  fam10_top

  # Seasonality of top 10 families (month profile)
  fam10_season <- fam10 %>%
    semi_join(fam10_top %>% slice_head(n = 10), by = "family") %>%
    group_by(family, month) %>%
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")

  ggplot(fam10_season, aes(x = month, y = occurrences, group = family, color = family)) +
    geom_line() +
    scale_x_continuous(breaks = 1:12) +
    scale_color_viridis_d() +
    labs(
      title = "Seasonality (top 10 families, 10km, all basisOfRecord)",
      x = "Month", y = "Occurrences"
    )
} else {
  message("Missing: ", fam10_path)
}

if (file.exists(ord10_path)) {
  ord10 <- readr::read_csv(ord10_path, show_col_types = FALSE) %>%
    filter(basisofrecord == "all") %>%
    mutate(
      yearmonth_chr = str_trim(as.character(yearmonth)),
      year = as.integer(substr(yearmonth_chr, 1, 4)),
      month = as.integer(substr(yearmonth_chr, 6, 7))
    )

  # Top orders overall
  ord10_top <- ord10 %>%
    group_by(order) %>%
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(occurrences)) %>%
    slice_head(n = 20)

  ord10_top

  # Year trend of top 8 orders
  ord10_year <- ord10 %>%
    semi_join(ord10_top %>% slice_head(n = 8), by = "order") %>%
    group_by(order, year) %>%
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")

  ggplot(ord10_year, aes(x = year, y = occurrences, group = order, color = order)) +
    geom_line() +
    scale_color_viridis_d() +
    labs(
      title = "Yearly trend (top 8 orders, 10km, all basisOfRecord)",
      x = "Year", y = "Occurrences"
    )
} else {
  message("Missing: ", ord10_path)
}
```

## 5) Species × time (Top N) quick look (10km) ----

```{r}
sp10t_path <- file.path(p_derived, "species_time_summary_topN_10km.csv")

if (file.exists(sp10t_path)) {
  sp10t <- readr::read_csv(sp10t_path, show_col_types = FALSE) %>%
    filter(basisofrecord == "all") %>%
    mutate(
      yearmonth_chr = str_trim(as.character(yearmonth)),
      year = as.integer(substr(yearmonth_chr, 1, 4)),
      month = as.integer(substr(yearmonth_chr, 6, 7))
    )

  # Top 20 species overall (within the Top-N set)
  sp10t_top <- sp10t %>%
    group_by(specieskey, species) %>%
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(occurrences)) %>%
    slice_head(n = 20)

  sp10t_top

  # Plot time series for top 6 species (monthly)
  sp10t_top6 <- sp10t_top %>% slice_head(n = 6)

  sp10t_plot <- sp10t %>%
    semi_join(sp10t_top6, by = c("specieskey", "species")) %>%
    mutate(ym = as.Date(paste0(yearmonth_chr, "-01")))

  ggplot(sp10t_plot, aes(x = ym, y = occurrences, group = species)) +
    geom_line() +
    labs(
      title = "Monthly time series (top 6 species within Top-N, 10km)",
      x = "Month", y = "Occurrences"
    )
} else {
  message("Missing (optional): ", sp10t_path)
}
```


## 6 Map check (10km): join cell_summary to grid polygons ----

```{r}
grid10_path <- file.path(p_data_proc, "grids_10km.gpkg")
cell10_path <- file.path(p_derived, "cell_summary_10km.csv")

stopifnot(file.exists(grid10_path), file.exists(cell10_path))

g10 <- sf::st_read(grid10_path, quiet = TRUE)

cell10_all <- readr::read_csv(cell10_path, show_col_types = FALSE) %>%
  dplyr::filter(basisofrecord == "all") %>%
  dplyr::select(eeacellcode, occurrences)

# Auto-detect likely cell-code field in the grid file
guess_code_field <- function(nms) {
  n <- tolower(nms)
  # prefer EEA code fields
  cand <- nms[grepl("eea", n) & grepl("code", n)]
  if (length(cand) > 0) return(cand[1])
  # then cellcode-ish fields
  cand <- nms[grepl("cell", n) & grepl("code", n)]
  if (length(cand) > 0) return(cand[1])
  # fallback: any grid/cell/code column
  cand <- nms[grepl("eea|cell|code|grid", n)]
  if (length(cand) > 0) return(cand[1])
  NA_character_
}

code_field <- guess_code_field(names(g10))
if (is.na(code_field)) {
  stop("Could not identify cell code field in grids_10km.gpkg. Columns are:\n",
       paste(names(g10), collapse = ", "))
}

# Make a join key called eeacellcode
g10$eeacellcode <- as.character(g10[[code_field]])

# Join and compute log intensity
g10_join <- g10 %>%
  dplyr::left_join(cell10_all, by = "eeacellcode") %>%
  dplyr::mutate(
    occurrences = dplyr::coalesce(occurrences, 0),
    log_occ = log10(occurrences + 1)
  )

# Quick join diagnostics
diag <- g10_join %>%
  sf::st_drop_geometry() %>%
  dplyr::summarise(
    n_cells = dplyr::n(),
    n_matched_nonzero = sum(occurrences > 0),
    n_zero = sum(occurrences == 0),
    pct_matched_nonzero = round(100 * n_matched_nonzero / n_cells, 2)
  )
diag

# Map (can be a bit heavy for 10km; still fine as a check)
ggplot(g10_join) +
  geom_sf(aes(fill = log_occ), linewidth = 0) +
  labs(
    title = "Map check: 10km grid joined with cube cell coverage (log10(occ+1))",
    subtitle = paste0("Cell code field in grid: ", code_field),
    fill = "log10(occ+1)"
  )
```

## 6) Taxonomic sanity: top species

```{r}
sp10 <- read_csv(file.path(p_derived, "species_summary_10km.csv"), show_col_types = FALSE) %>%
  filter(basisofrecord == "all") %>%
  arrange(desc(occurrences))

sp10 %>%
  select(specieskey, species, occurrences) %>%
  head(30)
```
