---
title: "Comprehensive Sanity Checks"
subtitle: "GBIF Gap Analysis - Data Quality & Readiness Assessment"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: flatly
    df_print: paged
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 3
    fig_width: 10
    fig_height: 6
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)

# Hide code in PDF output, show (with folding) in HTML
if (!knitr::is_html_output()) {
  knitr::opts_chunk$set(echo = FALSE)
}

library(here)
source(here("scripts", "00_setup.R"))

library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(purrr)
library(sf)
library(viridis)
library(scales)
library(glue)
library(knitr)
library(forcats)
library(tibble)
library(data.table)
library(DT)

theme_set(
 theme_minimal(base_size = 12) +
   theme(
     plot.title = element_text(face = "bold"),
     legend.position = "bottom"
   )
)

p_derived <- here(p_data_proc, "derived")
p_by_order <- here(p_derived, "by_order")
p_by_family <- here(p_derived, "by_family")
```

# Overview

This notebook provides comprehensive quality checks for the GBIF occurrence data, validating readiness for **spatial**, **temporal**, and **taxonomic** gap analysis.

**Run this after:** Scripts 01-06 (ingestion + derived summaries)

**Purpose:** 

1. Verify all required files exist and are properly formatted
2. Identify data quality issues before gap analysis
3. Characterize spatial, temporal, and taxonomic coverage
4. Flag potential biases that may affect gap interpretation

**Last run:** `r Sys.time()`

---

# 1. File Inventory & Validation {.tabset}

## 1.1 Core Processed Files

```{r check-core-files}
core_files <- tribble(
 ~file, ~location, ~required, ~description,
 "grids_10km.gpkg", p_data_proc, TRUE, "EEA 10km reference grid",
 "grids_50km.gpkg", p_data_proc, TRUE, "EEA 50km reference grid",
 "taxa_reference_current.rds", p_data_proc, TRUE, "National taxonomy backbone",
 "cube_manifest.csv", p_data_proc, FALSE, "Cube file metadata",
 "cube_totals_by_basisOfRecord.csv", p_data_proc, FALSE, "Occurrence totals summary"
) |>
 mutate(
   path = here(location, file),
   exists = file.exists(path),
   size_mb = map_dbl(path, ~if (file.exists(.x)) round(file.size(.x) / 1024^2, 2) else NA_real_),
   status = case_when(
     exists ~ "✅ OK",
     required ~ "❌ MISSING",
     TRUE ~ "⚠️ Optional"
   )
 )

core_files |>
 select(file, description, status, size_mb) |>
 kable(caption = "Core Processed Files")
```

## 1.2 Derived Summary Files (Script 06a)

```{r check-derived-core}
derived_core <- tribble(
 ~file, ~required, ~description,
 "cell_summary_10km.csv", TRUE, "Occurrences per cell (10km)",
 "cell_summary_50km.csv", TRUE, "Occurrences per cell (50km)",
 "time_summary_10km.csv", TRUE, "Occurrences per month (10km)",
 "time_summary_50km.csv", TRUE, "Occurrences per month (50km)",
 "cell_time_summary_10km.csv", TRUE, "Cell × month matrix (10km)",
 "cell_time_summary_50km.csv", TRUE, "Cell × month matrix (50km)",
 "order_cell_summary_10km.csv", TRUE, "Order × cell (10km)",
 "order_cell_summary_50km.csv", TRUE, "Order × cell (50km)",
 "order_time_summary_10km.csv", TRUE, "Order × month (10km)",
 "order_time_summary_50km.csv", TRUE, "Order × month (50km)",
 "family_time_summary_10km.csv", TRUE, "Family × month (10km)",
 "family_time_summary_50km.csv", TRUE, "Family × month (50km)",
 "cube_key_summary.csv", TRUE, "Cube file statistics"
) |>
 mutate(
   path = here(p_derived, file),
   exists = file.exists(path),
   size_mb = map_dbl(path, ~if (file.exists(.x)) round(file.size(.x) / 1024^2, 2) else NA_real_),
   rows = map_dbl(path, ~{
     if (file.exists(.x)) {
       tryCatch(nrow(fread(.x, select = 1L)), error = function(e) NA_real_)
     } else NA_real_
   }),
   status = case_when(
     exists ~ "✅",
     required ~ "❌",
     TRUE ~ "⚠️"
   )
 )

derived_core |>
 select(file, description, status, size_mb, rows) |>
 mutate(rows = scales::comma(rows)) |>
 kable(caption = "Core Derived Files (from 06a)")
```

## 1.3 Species-Level Files (Script 06b)

```{r check-species-files}
# Count files in by_order and by_family directories
by_order_counts <- tibble(
 type = c("species_summary", "species_cell", "species_time", "species_cell_time"),
 dir = here(p_by_order, type)
) |>
 mutate(
   n_files = map_int(dir, ~{
     if (dir.exists(.x)) length(list.files(.x, pattern = "\\.csv$")) else 0L
   }),
   total_size_mb = map_dbl(dir, ~{
     if (dir.exists(.x)) {
       files <- list.files(.x, pattern = "\\.csv$", full.names = TRUE)
       sum(file.size(files)) / 1024^2
     } else 0
   })
 )

by_family_counts <- tibble(
 type = c("species_summary", "species_cell", "species_time", "species_cell_time"),
 dir = here(p_by_family, type)
) |>
 mutate(
   n_files = map_int(dir, ~{
     if (dir.exists(.x)) length(list.files(.x, pattern = "\\.csv$")) else 0L
   }),
   total_size_mb = map_dbl(dir, ~{
     if (dir.exists(.x)) {
       files <- list.files(.x, pattern = "\\.csv$", full.names = TRUE)
       sum(file.size(files)) / 1024^2
     } else 0
   })
 )

species_summary <- bind_rows(
 by_order_counts |> mutate(location = "by_order"),
 by_family_counts |> mutate(location = "by_family")
) |>
 select(location, type, n_files, total_size_mb) |>
 mutate(total_size_mb = round(total_size_mb, 1))

species_summary |>
 pivot_wider(names_from = location, values_from = c(n_files, total_size_mb)) |>
 kable(caption = "Species-Level Files (from 06b)")

# Total counts
cat("\n**Total species-level files:**", sum(species_summary$n_files), "\n")
cat("**Total size:**", round(sum(species_summary$total_size_mb), 1), "MB\n")
```

## 1.4 Validation Summary

```{r validation-summary}
validation <- tribble(
 ~check, ~status, ~details,
 "Core processed files",
   ifelse(all(core_files$exists[core_files$required]), "✅ PASS", "❌ FAIL"),
   glue("{sum(core_files$exists)} / {nrow(core_files)} files"),
 "Derived core files (06a)",
   ifelse(all(derived_core$exists[derived_core$required]), "✅ PASS", "❌ FAIL"),
   glue("{sum(derived_core$exists)} / {sum(derived_core$required)} required"),
 "Species by_order files (06b)",
   ifelse(sum(by_order_counts$n_files) > 0, "✅ PASS", "❌ FAIL"),
   glue("{sum(by_order_counts$n_files)} files"),
 "Species by_family files (06b)",
   ifelse(sum(by_family_counts$n_files) >= 0, "✅ PASS", "⚠️ None (OK if no large orders)"),
   glue("{sum(by_family_counts$n_files)} files")
)

validation |> kable(caption = "File Validation Summary")
```

---

# 2. Spatial Coverage Assessment {.tabset}

```{r load-spatial-data}
cell10 <- fread(here(p_derived, "cell_summary_10km.csv"))
cell50 <- fread(here(p_derived, "cell_summary_50km.csv"))

# Load grids for mapping
grid10 <- tryCatch(
 st_read(here(p_data_proc, "grids_10km.gpkg"), quiet = TRUE),
 error = function(e) NULL
)
grid50 <- tryCatch(
 st_read(here(p_data_proc, "grids_50km.gpkg"), quiet = TRUE),
 error = function(e) NULL
)
```

## 2.1 Coverage Statistics

```{r spatial-stats}
calc_spatial_stats <- function(cell_data, grid_name) {
 cell_data |>
   filter(basisofrecord == "all") |>
   summarise(
     grid = grid_name,
     total_cells = n(),
     cells_with_data = sum(occurrences > 0),
     zero_cells = sum(occurrences == 0),
     pct_coverage = round(100 * cells_with_data / total_cells, 1),
     total_occurrences = sum(occurrences),
     mean_occ = round(mean(occurrences)),
     median_occ = median(occurrences),
     sd_occ = round(sd(occurrences)),
     min_occ = min(occurrences),
     max_occ = max(occurrences),
     q10 = quantile(occurrences, 0.10),
     q25 = quantile(occurrences, 0.25),
     q75 = quantile(occurrences, 0.75),
     q90 = quantile(occurrences, 0.90)
   )
}

spatial_stats <- bind_rows(
 calc_spatial_stats(cell10, "10km"),
 calc_spatial_stats(cell50, "50km")
)

spatial_stats |>
 mutate(across(c(total_occurrences, mean_occ, sd_occ, max_occ), comma)) |>
 kable(caption = "Spatial Coverage Statistics")
```

## 2.2 Occurrence Distribution by Cell

```{r spatial-distribution, fig.height=5}
cell10_all <- cell10[basisofrecord == "all"]

# Count zeros
n_zero <- sum(cell10_all$occurrences == 0, na.rm = TRUE)
n_nonzero <- sum(cell10_all$occurrences > 0, na.rm = TRUE)
n_total <- nrow(cell10_all)

cat("**Cell counts:** Total:", n_total, "| With data:", n_nonzero, "| Zero:", n_zero, "\n\n")

# Plot 1: Linear scale to show the zero cells and skew
p1 <- ggplot(cell10_all, aes(x = occurrences)) +
 geom_histogram(bins = 100, fill = viridis(1, begin = 0.3), alpha = 0.8) +
 scale_x_continuous(labels = comma) +
 labs(title = "Distribution of Occurrences per Cell (10km)",
      subtitle = paste0("Linear scale - showing all ", comma(n_total), " cells (", comma(n_zero), " with zero)"),
      x = "Occurrences", y = "Number of Cells")

# Plot 2: Log scale for non-zero cells only
cell10_nonzero <- cell10_all[occurrences > 0]

p2 <- ggplot(cell10_nonzero, aes(x = occurrences)) +
 geom_histogram(bins = 50, fill = viridis(1, begin = 0.5), alpha = 0.8) +
 scale_x_log10(labels = comma) +
 labs(title = "Distribution (non-zero cells only)",
      subtitle = paste0("Log scale - ", comma(n_nonzero), " cells with data"),
      x = "Occurrences (log scale)", y = "Number of Cells")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## 2.3 Zero-Coverage Cells (Gap Candidates)
 
```{r zero-cells}
zero_cells_10km <- cell10[basisofrecord == "all" & occurrences == 0]
zero_cells_50km <- cell50[basisofrecord == "all" & occurrences == 0]

cat("**Zero-coverage cells (10km):**", nrow(zero_cells_10km), 
   "(", round(100 * nrow(zero_cells_10km) / nrow(cell10[basisofrecord == "all"]), 1), "%)\n")
cat("**Zero-coverage cells (50km):**", nrow(zero_cells_50km),
   "(", round(100 * nrow(zero_cells_50km) / nrow(cell50[basisofrecord == "all"]), 1), "%)\n")

if (nrow(zero_cells_10km) > 0) {
 cat("\nFirst 10 zero-coverage cell codes (10km):\n")
 print(head(zero_cells_10km$eeacellcode, 10))
}
```

## 2.4 Low-Coverage Cells (Bottom 10%)

```{r low-coverage}
q10_threshold <- quantile(cell10[basisofrecord == "all"]$occurrences, 0.10)
low_cells <- cell10[basisofrecord == "all" & occurrences > 0 & occurrences <= q10_threshold]

cat("**10th percentile threshold:**", q10_threshold, "occurrences\n")
cat("**Cells at or below threshold:**", nrow(low_cells), "\n")

low_cells |>
 summarise(
   n_cells = n(),
   min_occ = min(occurrences),
   max_occ = max(occurrences),
   mean_occ = round(mean(occurrences), 1)
 ) |>
 kable(caption = "Low-Coverage Cells (bottom 10%)")
```

## 2.5 Spatial Map (if grid available)

```{r spatial-map, fig.height=8, eval=!is.null(grid10)}
if (!is.null(grid10)) {
  # Find the cell code column name in the grid (could be cellcode, CELLCODE, eea_cellco, etc.)
  grid_names <- names(grid10)
  cellcode_col <- grid_names[str_detect(tolower(grid_names), "cell")][1]
  
  if (!is.na(cellcode_col)) {
    # Rename to match for join
    grid10_temp <- grid10
    grid10_temp$join_cellcode <- grid10_temp[[cellcode_col]]
    
    # Join occurrence data to grid
    cell10_for_join <- cell10[basisofrecord == "all", .(eeacellcode, occurrences, n_species)]
    
    grid10_occ <- grid10_temp |>
      left_join(
        cell10_for_join,
        by = c("join_cellcode" = "eeacellcode")
      ) |>
      mutate(
        occurrences = replace_na(occurrences, 0),
        log_occ = log10(occurrences + 1)
      )
    
    ggplot(grid10_occ) +
      geom_sf(aes(fill = log_occ), color = NA) +
      scale_fill_viridis(option = "inferno", name = "log10(occ+1)") +
      labs(title = "Spatial Distribution of Occurrences (10km Grid)",
           subtitle = "Log scale coloring") +
      theme_void() +
      theme(legend.position = "right")
  } else {
    cat("Could not find cell code column in grid. Available columns:", paste(grid_names, collapse = ", "))
  }
}
```

## 2.6 Coverage by Basis of Record

```{r coverage-by-basis}
cell10_by_basis <- cell10[, .(
 n_cells_with_data = sum(occurrences > 0),
 total_occurrences = sum(occurrences),
 mean_occ = round(mean(occurrences[occurrences > 0]), 1)
), by = basisofrecord][order(-total_occurrences)]

cell10_by_basis |>
 mutate(total_occurrences = comma(total_occurrences)) |>
 kable(caption = "Spatial Coverage by Basis of Record (10km)")
```

---

# 3. Temporal Coverage Assessment {.tabset}

```{r load-temporal-data}
time10 <- fread(here(p_derived, "time_summary_10km.csv"))
time50 <- fread(here(p_derived, "time_summary_50km.csv"))

# Parse yearmonth
time10[, `:=`(
 yearmonth_chr = str_trim(as.character(yearmonth)),
 year = as.integer(str_sub(as.character(yearmonth), 1, 4)),
 month = as.integer(str_sub(as.character(yearmonth), 6, 7))
)]
time10[, date := as.Date(paste0(yearmonth_chr, "-01"))]
```

## 3.1 Temporal Range

```{r temporal-range}
time10_all <- time10[basisofrecord == "all"]

temporal_range <- time10_all[, .(
 min_year = min(year, na.rm = TRUE),
 max_year = max(year, na.rm = TRUE),
 n_years = uniqueN(year),
 n_months = .N,
 total_occurrences = sum(occurrences)
)]

temporal_range |> kable(caption = "Temporal Range Summary")
```

## 3.2 Annual Trend

```{r annual-trend, fig.height=5}
annual <- time10_all[, .(
 occurrences = sum(occurrences),
 n_months = .N,
 n_cells = sum(n_cells, na.rm = TRUE)
), by = year][order(year)]

ggplot(annual, aes(x = year, y = occurrences)) +
 geom_line(color = viridis(1, begin = 0.4), linewidth = 1) +
 geom_point(color = viridis(1, begin = 0.4), size = 2) +
 scale_y_continuous(labels = comma) +
 labs(title = "Annual Occurrence Trend",
      subtitle = "Total occurrences per year",
      x = "Year", y = "Occurrences")
```

## 3.3 Seasonal Pattern

```{r seasonal-pattern, fig.height=5}
monthly <- time10_all[, .(
 occurrences = sum(occurrences),
 n_years = uniqueN(year)
), by = month][order(month)]

ggplot(monthly, aes(x = factor(month), y = occurrences)) +
 geom_col(fill = viridis(1, begin = 0.5)) +
 scale_y_continuous(labels = comma) +
 scale_x_discrete(labels = month.abb) +
 labs(title = "Seasonal Distribution of Occurrences",
      subtitle = "Aggregated across all years",
      x = "Month", y = "Total Occurrences")
```

## 3.4 Year × Month Heatmap

```{r temporal-heatmap, fig.height=10}
year_month <- time10_all[, .(occurrences = sum(occurrences)), by = .(year, month)]

ggplot(year_month, aes(x = factor(month), y = year, fill = log10(occurrences + 1))) +
 geom_tile(color = "white", size = 0.1) +
 scale_fill_viridis(option = "inferno", name = "log10(occ+1)") +
 scale_x_discrete(labels = month.abb) +
 labs(title = "Sampling Intensity: Year × Month",
      x = "Month", y = "Year") +
 theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 3.5 Temporal Gaps

```{r temporal-gaps}
# Find years with no data
all_years <- seq(min(time10_all$year, na.rm = TRUE), max(time10_all$year, na.rm = TRUE))
years_with_data <- unique(time10_all$year)
missing_years <- setdiff(all_years, years_with_data)

cat("**Years with no data:**", length(missing_years), "\n")
if (length(missing_years) > 0 && length(missing_years) <= 20) {
 cat("Missing years:", paste(missing_years, collapse = ", "), "\n")
}

# Find months with unusually low data (< 10% of median)
monthly_median <- median(monthly$occurrences)
low_months <- monthly[occurrences < monthly_median * 0.1]

if (nrow(low_months) > 0) {
 cat("\n**Months with unusually low data (<10% of median):**\n")
 print(low_months)
}
```
## 3.6 Recent Data Availability

```{r recent-data}
current_year <- as.integer(format(Sys.Date(), "%Y"))
recent_years <- (current_year - 5):current_year

recent_data <- annual[year %in% recent_years]

if (nrow(recent_data) > 0) {
 recent_data |>
   mutate(occurrences = comma(occurrences)) |>
   kable(caption = "Recent Years Data Availability")
} else {
 cat("No data from the last 5 years\n")
}

# Most recent data point
most_recent <- time10_all[which.max(date)]
cat("\n**Most recent data point:**", as.character(most_recent$date), 
   "with", comma(most_recent$occurrences), "occurrences\n")
```

## 3.7 Temporal Coverage by Basis of Record

```{r temporal-by-basis, fig.height=6}
time_by_basis <- time10[basisofrecord != "all", .(
 occurrences = sum(occurrences)
), by = .(year, basisofrecord)]

ggplot(time_by_basis, aes(x = year, y = occurrences, fill = basisofrecord)) +
  geom_area(alpha = 0.8) +
  scale_y_continuous(labels = comma) +
  scale_fill_viridis_d(option = "turbo") +
  facet_wrap(~basisofrecord, scales = "free_y", ncol = 2) +
  labs(title = "Temporal Trend by Basis of Record",
       x = "Year", y = "Occurrences") +
  theme(legend.position = "none")

ggplot(time_by_basis, aes(x = year, y = occurrences, color = basisofrecord)) +
  geom_line(linewidth = 1) +
  scale_y_log10(labels = comma) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Temporal Trend by Basis of Record",
       subtitle = "Log scale",
       x = "Year", y = "Occurrences (log)", color = "Basis of Record")

```

---

# 4. Taxonomic Coverage Assessment {.tabset}

```{r load-taxonomic-data}
# Load taxa reference
taxa_ref <- tryCatch(
 readRDS(here(p_data_proc, "taxa_reference_current.rds")),
 error = function(e) NULL
)

# Load order-level summary for quick stats
order_time <- fread(here(p_derived, "order_time_summary_10km.csv"))
family_time <- fread(here(p_derived, "family_time_summary_10km.csv"))

# Get species counts from by_order files
species_files <- list.files(here(p_by_order, "species_summary"), pattern = "_10km\\.csv$", full.names = TRUE)
```

## 4.1 Taxonomic Backbone Summary

```{r taxa-backbone}
if (!is.null(taxa_ref)) {
 taxa_summary <- taxa_ref |>
   filter(taxonomicStatus == "accepted" | is.na(taxonomicStatus)) |>
   summarise(
     total_taxa = n(),
     n_species = sum(taxonRank == "species", na.rm = TRUE),
     n_genus = sum(taxonRank == "genus", na.rm = TRUE),
     n_family = sum(taxonRank == "family", na.rm = TRUE),
     n_with_threat_dyntaxa = sum(!is.na(threatStatus_dyntaxa)),
     n_with_threat_redlist = sum(!is.na(threatStatus_redlist)),
     n_kingdoms = uniqueN(kingdom)
   )
 
 taxa_summary |> 
   pivot_longer(everything(), names_to = "metric", values_to = "count") |>
   mutate(count = comma(count)) |>
   kable(caption = "Taxonomic Backbone Summary")
 
 # Kingdom breakdown
 taxa_ref |>
   filter(taxonomicStatus == "accepted" | is.na(taxonomicStatus)) |>
   count(kingdom, sort = TRUE) |>
   kable(caption = "Taxa by Kingdom")
}
```

## 4.2 Order-Level Coverage in GBIF

```{r order-coverage}
order_summary <- order_time[basisofrecord == "all", .(
 total_occurrences = sum(occurrences),
 n_months = uniqueN(yearmonth),
 mean_monthly_occ = round(mean(occurrences))
), by = order][order(-total_occurrences)]

cat("**Number of orders in GBIF data:**", nrow(order_summary), "\n\n")

order_summary |>
 head(20) |>
 mutate(total_occurrences = comma(total_occurrences)) |>
 kable(caption = "Top 20 Orders by Occurrence Count")
```

## 4.3 Order Rank-Abundance

```{r order-rank, fig.height=5}
order_summary[, rank := .I]

ggplot(order_summary, aes(x = rank, y = total_occurrences)) +
 geom_point(color = viridis(1, begin = 0.4), alpha = 0.7) +
 geom_line(color = viridis(1, begin = 0.4), alpha = 0.5) +
 scale_y_log10(labels = comma) +
 labs(title = "Order Rank-Abundance Distribution",
      subtitle = "Log scale - showing dominance pattern",
      x = "Order Rank", y = "Total Occurrences (log)")
```

## 4.4 Family-Level Coverage

```{r family-coverage}
family_summary <- family_time[basisofrecord == "all", .(
 total_occurrences = sum(occurrences),
 n_months = uniqueN(yearmonth)
), by = .(order, family)][order(-total_occurrences)]

cat("**Number of families in GBIF data:**", nrow(family_summary), "\n\n")

family_summary |>
 head(20) |>
 mutate(total_occurrences = comma(total_occurrences)) |>
 kable(caption = "Top 20 Families by Occurrence Count")
```

## 4.5 Species-Level Statistics

```{r species-stats}
# Read and combine species summary files
if (length(species_files) > 0) {
 species_counts <- map_dfr(species_files, ~{
   dt <- fread(.x)
   dt[basisofrecord == "all", .(
     order = str_extract(basename(.x), "species_summary_(.+)_10km", group = 1),
     n_species = uniqueN(specieskey),
     total_occurrences = sum(occurrences)
   )]
 })
 
 species_totals <- species_counts[, .(
   n_species = sum(n_species),
   total_occurrences = sum(total_occurrences)
 )]
 
 cat("**Total unique species in GBIF data:**", comma(species_totals$n_species), "\n")
 cat("**Total occurrences:**", comma(species_totals$total_occurrences), "\n\n")
 
 species_counts[order(-n_species)] |>
   head(15) |>
   mutate(total_occurrences = comma(total_occurrences)) |>
   kable(caption = "Species Count by Order (from species_summary files)")
}
```

## 4.6 Threat Status Coverage

```{r threat-status}
if (!is.null(taxa_ref)) {
 # Backbone threat status
 dyntaxa_threat <- taxa_ref |>
   filter(!is.na(threatStatus_dyntaxa)) |>
   count(threatStatus_dyntaxa, sort = TRUE) |>
   rename(status = threatStatus_dyntaxa, n_dyntaxa = n)
 
 # Red List threat status
 redlist_threat <- taxa_ref |>
   filter(!is.na(threatStatus_redlist)) |>
   count(threatStatus_redlist, sort = TRUE) |>
   rename(status = threatStatus_redlist, n_redlist = n)
 
 # Combine
 threat_comparison <- full_join(dyntaxa_threat, redlist_threat, by = "status") |>
   mutate(across(starts_with("n_"), ~replace_na(.x, 0))) |>
   arrange(desc(n_dyntaxa + n_redlist))
 
 threat_comparison |>
   kable(caption = "Threat Status Coverage: Backbone vs Red List")
 
 # Priority threatened taxa
 priority_statuses <- c("CR", "EN", "VU", "NT")
 
 n_priority_dyntaxa <- taxa_ref |> 
   filter(threatStatus_dyntaxa %in% priority_statuses) |> nrow()
 n_priority_redlist <- taxa_ref |> 
   filter(threatStatus_redlist %in% priority_statuses) |> nrow()
 
 cat("\n**Priority threatened taxa (CR/EN/VU/NT):**\n")
 cat("- From backbone:", n_priority_dyntaxa, "\n")
 cat("- From Red List:", n_priority_redlist, "\n")
}
```

---

# 5. Bias Assessment {.tabset}

## 5.1 Basis of Record Bias

```{r basis-bias}
totals_file <- here(p_data_proc, "cube_totals_by_basisOfRecord.csv")

if (file.exists(totals_file)) {
 totals <- fread(totals_file)
 
 totals_summary <- totals[, .(total_occurrences = sum(total_occurrences)), by = basisOfRecord]
 totals_summary[, pct := round(100 * total_occurrences / sum(total_occurrences), 1)]
 
 totals_summary[order(-total_occurrences)] |>
   mutate(total_occurrences = comma(total_occurrences)) |>
   kable(caption = "Occurrence Share by Basis of Record")
 
 # Dominance check
 max_pct <- max(totals_summary$pct)
 dominant_basis <- totals_summary[pct == max_pct, basisOfRecord]
 
 if (max_pct > 80) {
   cat("\n⚠️ **Warning:** ", dominant_basis, " accounts for ", max_pct, 
       "% of all occurrences. This may bias gap analysis.\n", sep = "")
 }
}
```

## 5.2 Spatial Sampling Bias

```{r spatial-bias, fig.height=5}
# Check for hotspots (cells with disproportionately high occurrences)
# Note: cell10_all should already be defined from section 2.3
# If not, recreate it here
if (!exists("cell10_all") || nrow(cell10_all) == 0) {
  cell10_all <- cell10[basisofrecord == "all"]
}

# Debug info
cat("**Debug info:**\n")
cat("- Rows in cell10_all:", nrow(cell10_all), "\n")
cat("- Columns:", paste(names(cell10_all), collapse = ", "), "\n")
cat("- Non-zero occurrences:", sum(cell10_all$occurrences > 0, na.rm = TRUE), "\n\n")

# Calculate Gini coefficient for spatial inequality
gini <- function(x) {
  x <- as.numeric(x)  # Ensure numeric
  x <- x[!is.na(x) & x > 0]  # Remove NA and zero values
  if (length(x) < 2) {
    cat("Gini debug: Only", length(x), "non-zero values\n")
    return(NA_real_)
  }
  if (sum(x) == 0) {
    cat("Gini debug: Sum is zero\n")
    return(NA_real_)
  }
  x <- sort(x)
  n <- length(x)
  result <- sum((2 * seq_len(n) - n - 1) * x) / (n * sum(x))
  return(result)
}

spatial_gini <- gini(cell10_all$occurrences)

if (!is.na(spatial_gini)) {
  cat("**Spatial Gini coefficient:**", round(spatial_gini, 3), "\n")
  cat("(0 = perfectly even, 1 = all data in one cell)\n\n")
  
  if (spatial_gini > 0.7) {
    cat("⚠️ **Warning:** High spatial inequality detected. Data is concentrated in few cells.\n")
  }
} else {
  cat("**Spatial Gini coefficient:** Could not calculate\n\n")
}

# Top 10 cells vs rest
occ_values <- as.numeric(cell10_all$occurrences)
occ_values <- occ_values[!is.na(occ_values)]

if (length(occ_values) >= 10 && sum(occ_values) > 0) {
  top10_occ <- sum(head(sort(occ_values, decreasing = TRUE), 10))
  total_occ <- sum(occ_values)
  top10_pct <- round(100 * top10_occ / total_occ, 1)
  cat("**Top 10 cells account for:**", top10_pct, "% of all occurrences\n")
} else {
  top10_pct <- NA_real_
  cat("**Top 10 cells:** Insufficient data to calculate\n")
}

# Show the top 10 cells
top10_cells <- cell10_all[order(-occurrences)][1:10, .(eeacellcode, occurrences, n_species)]
top10_cells[, pct_of_total := round(100 * occurrences / sum(cell10_all$occurrences), 2)]

cat("\n**Top 10 most-sampled cells:**\n\n")
print(top10_cells)

# Map top 10 cells (if grid available)
if (!is.null(grid10)) {
  grid_names <- names(grid10)
  cellcode_col <- grid_names[str_detect(tolower(grid_names), "cell")][1]
  
  if (!is.na(cellcode_col)) {
    top10_codes <- top10_cells$eeacellcode
    
    grid10_top10 <- grid10 |>
      mutate(is_top10 = .data[[cellcode_col]] %in% top10_codes) |>
      left_join(
        top10_cells |> mutate(eeacellcode = as.character(eeacellcode)),
        by = setNames("eeacellcode", cellcode_col)
      )
    
    p_top10 <- ggplot() +
      geom_sf(data = grid10, fill = "grey90", color = "white", size = 0.1) +
      geom_sf(data = grid10_top10 |> filter(is_top10), 
              aes(fill = occurrences), color = "red", size = 0.5) +
      scale_fill_viridis(option = "plasma", labels = comma, name = "Occurrences") +
      labs(title = "Top 10 Most-Sampled Cells",
           subtitle = "These cells may represent sampling hotspots (cities, research stations)") +
      theme_void() +
      theme(legend.position = "right")
    
    print(p_top10)
  }
}

```

## 5.3 Temporal Sampling Bias

```{r temporal-bias}
# Weekend effect (if we had day-of-week data, we'd check this)
# For now, check year-over-year growth

annual_growth <- annual[, .(
 year, 
 occurrences,
 yoy_growth = round(100 * (occurrences - shift(occurrences)) / shift(occurrences), 1)
)]

# Recent acceleration?
recent_5yr <- annual_growth[year >= current_year - 5]
avg_recent_growth <- mean(recent_5yr$yoy_growth, na.rm = TRUE)

cat("**Average year-over-year growth (last 5 years):**", round(avg_recent_growth, 1), "%\n")

if (avg_recent_growth > 20) {
 cat("⚠️ **Note:** Rapid recent growth may indicate citizen science surge.\n")
 cat("Consider temporal detrending for trend analysis.\n")
}
```

## 5.4 Taxonomic Sampling Bias

```{r taxonomic-bias}
# Check if certain orders are over/under-represented relative to expected diversity
order_summary <- order_time[basisofrecord == "all", .(
 total_occurrences = sum(occurrences)
), by = order][order(-total_occurrences)]

# Top 5 orders share
top5_orders <- head(order_summary, 5)
top5_pct <- round(100 * sum(top5_orders$total_occurrences) / sum(order_summary$total_occurrences), 1)

cat("**Top 5 orders account for:**", top5_pct, "% of all occurrences\n")
cat("Top 5 orders:", paste(top5_orders$order, collapse = ", "), "\n\n")

if (top5_pct > 70) {
 cat("⚠️ **Warning:** Strong taxonomic bias toward few orders.\n")
 cat("Gap analysis should consider order-specific expectations.\n")
}
```

---

# 6. Gap Analysis Readiness {.tabset}

## 6.1 Spatial Gap Analysis Readiness

```{r spatial-readiness}
spatial_ready <- tribble(
 ~requirement, ~status, ~details,
 "Cell summary available",
   ifelse(file.exists(here(p_derived, "cell_summary_10km.csv")), "✅", "❌"),
   "Required for zero-coverage detection",
 "Cell × time summary available",
   ifelse(file.exists(here(p_derived, "cell_time_summary_10km.csv")), "✅", "❌"),
   "Required for data recency analysis",
 "Grid geometries available",
   ifelse(!is.null(grid10), "✅", "❌"),
   "Required for mapping",
 "Order × cell available",
   ifelse(file.exists(here(p_derived, "order_cell_summary_10km.csv")), "✅", "❌"),
   "Required for taxonomic spatial gaps",
 "Zero cells identified",
   glue("✅ {nrow(zero_cells_10km)} cells"),
   "Candidates for spatial gaps",
 "Low coverage cells identified",
   glue("✅ {nrow(low_cells)} cells"),
   "Below 10th percentile"
)

spatial_ready |> kable(caption = "Spatial Gap Analysis Readiness")
```

## 6.2 Temporal Gap Analysis Readiness

```{r temporal-readiness}
temporal_ready <- tribble(
 ~requirement, ~status, ~details,
 "Time summary available",
   ifelse(file.exists(here(p_derived, "time_summary_10km.csv")), "✅", "❌"),
   "Required for temporal trends",
 "Cell × time summary available",
   ifelse(file.exists(here(p_derived, "cell_time_summary_10km.csv")), "✅", "❌"),
   "Required for cell recency",
 "Order × time available",
   ifelse(file.exists(here(p_derived, "order_time_summary_10km.csv")), "✅", "❌"),
   "Required for taxonomic temporal patterns",
 "Family × time available",
   ifelse(file.exists(here(p_derived, "family_time_summary_10km.csv")), "✅", "❌"),
   "Required for family-level trends",
 "Year range adequate",
   ifelse(temporal_range$n_years >= 10, "✅", "⚠️"),
   glue("{temporal_range$n_years} years of data"),
 "Recent data available",
   ifelse(max(time10_all$year, na.rm = TRUE) >= current_year - 2, "✅", "⚠️"),
   glue("Most recent: {max(time10_all$year, na.rm = TRUE)}")
)

temporal_ready |> kable(caption = "Temporal Gap Analysis Readiness")
```

## 6.3 Taxonomic Gap Analysis Readiness

```{r taxonomic-readiness}
taxonomic_ready <- tribble(
 ~requirement, ~status, ~details,
 "Taxa backbone available",
   ifelse(!is.null(taxa_ref), "✅", "❌"),
   "National taxonomic backbone reference",
 "Threat status (backbone)",
   ifelse(!is.null(taxa_ref) && sum(!is.na(taxa_ref$threatStatus_dyntaxa)) > 0, "✅", "⚠️"),
   ifelse(!is.null(taxa_ref), glue("{sum(!is.na(taxa_ref$threatStatus_dyntaxa))} taxa"), "N/A"),
 "Threat status (Red List)",
   ifelse(!is.null(taxa_ref) && sum(!is.na(taxa_ref$threatStatus_redlist)) > 0, "✅", "⚠️"),
   ifelse(!is.null(taxa_ref), glue("{sum(!is.na(taxa_ref$threatStatus_redlist))} taxa"), "N/A"),
 "Species summaries available",
   ifelse(length(species_files) > 0, "✅", "❌"),
   glue("{length(species_files)} order files"),
 "Bias correction columns",
   "✅",
   "familycount, ordercount, classcount in species tables"
)

taxonomic_ready |> kable(caption = "Taxonomic Gap Analysis Readiness")
```

## 6.4 Overall Readiness Score

```{r overall-readiness}
# Count checks
all_checks <- bind_rows(
 spatial_ready |> mutate(category = "Spatial"),
 temporal_ready |> mutate(category = "Temporal"),
 taxonomic_ready |> mutate(category = "Taxonomic")
)

passed <- sum(str_detect(all_checks$status, "✅"))
total <- nrow(all_checks)
pct_passed <- round(100 * passed / total)

cat("## Overall Gap Analysis Readiness\n\n")
cat("**Checks passed:**", passed, "/", total, "(", pct_passed, "%)\n\n")

if (pct_passed >= 90) {
 cat("✅ **READY** for comprehensive gap analysis\n")
} else if (pct_passed >= 70) {
 cat("⚠️ **MOSTLY READY** - some components missing, review warnings\n")
} else {
 cat("❌ **NOT READY** - critical components missing\n")
}
```

---

# 7. Recommendations

```{r recommendations}
recommendations <- character()

# Spatial recommendations
if (nrow(zero_cells_10km) > 0) {
 recommendations <- c(recommendations, 
   glue("- Investigate {nrow(zero_cells_10km)} zero-coverage cells for spatial gap analysis"))
}

if (!is.na(spatial_gini) && spatial_gini > 0.7) {
 recommendations <- c(recommendations,
   "- Consider spatial bias correction due to high sampling inequality")
}

# Temporal recommendations
if (length(missing_years) > 0) {
 recommendations <- c(recommendations,
   glue("- {length(missing_years)} years have no data - may affect trend analysis"))
}

if (avg_recent_growth > 20) {
 recommendations <- c(recommendations,
   "- Apply temporal detrending to account for citizen science growth")
}

# Taxonomic recommendations
if (top5_pct > 70) {
 recommendations <- c(recommendations,
   "- Use relative occurrence (bias correction) for cross-taxa comparisons")
}

if (!is.null(taxa_ref)) {
 n_unmatched_estimate <- nrow(taxa_ref) - species_totals$n_species
 if (n_unmatched_estimate > 1000) {
   recommendations <- c(recommendations,
     glue("- ~{comma(n_unmatched_estimate)} Backbone taxa may be missing from GBIF - run taxonomy matching"))
 }
}

if (length(recommendations) > 0) {
 cat("Based on this sanity check, consider the following:\n\n")
 cat(paste(recommendations, collapse = "\n"))
} else {
 cat("No major issues detected. Data appears ready for gap analysis.\n")
}
```

---

# 8. Data Quality Summary

```{r final-summary}
quality_summary <- tribble(
 ~dimension, ~metric, ~value, ~status,
 "Files", "Core files present", paste0(sum(core_files$exists), "/", nrow(core_files)),
   ifelse(all(core_files$exists[core_files$required]), "✅", "❌"),
 "Files", "Derived files present", paste0(sum(derived_core$exists), "/", nrow(derived_core)),
   ifelse(all(derived_core$exists[derived_core$required]), "✅", "❌"),
 "Spatial", "Cell coverage", paste0(spatial_stats$pct_coverage[1], "%"),
   ifelse(spatial_stats$pct_coverage[1] > 50, "✅", "⚠️"),
 "Spatial", "Zero-coverage cells", comma(nrow(zero_cells_10km)),
   ifelse(nrow(zero_cells_10km) < nrow(cell10[basisofrecord == "all"]) * 0.3, "✅", "⚠️"),
 "Temporal", "Year range", paste0(temporal_range$min_year, "-", temporal_range$max_year),
   ifelse(temporal_range$n_years >= 20, "✅", "⚠️"),
 "Temporal", "Data recency", as.character(max(time10_all$year, na.rm = TRUE)),
   ifelse(max(time10_all$year, na.rm = TRUE) >= current_year - 2, "✅", "⚠️"),
 "Taxonomic", "Orders covered", comma(nrow(order_summary)),
   "✅",
 "Taxonomic", "Families covered", comma(nrow(family_summary)),
   "✅",
 "Bias", "Spatial Gini", ifelse(!is.na(spatial_gini), as.character(round(spatial_gini, 2)), "N/A"),
   ifelse(!is.na(spatial_gini) && spatial_gini < 0.7, "✅", "⚠️"),
 "Bias", "Top 5 orders share", paste0(top5_pct, "%"),
   ifelse(top5_pct < 70, "✅", "⚠️")
)

quality_summary |> kable(caption = "Data Quality Summary")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
