---
title: "GBIF-mediated data gaps in Sweden: Spatial, temporal and taxonomic gaps"
author: "Lena Thöle"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,     # show code
  eval = TRUE,     # execute code
  error = TRUE,    # show errors (and keep knitting)
  warning = TRUE,  # show warnings
  message = TRUE,   # show messages (e.g., package startup)
  collapse = TRUE,  # show code + output together
  comment = "#>"    # prefix output like reprex
)

# Always load the project setup from project root
if (!requireNamespace("here", quietly = TRUE)) {
  stop("Package 'here' is required. Install it and run renv::snapshot().")
}
source(here::here("scripts", "00_setup.R"))

# Paths
data_proc_rel <- cfg_get("paths.data_proc", "data_proc")
out_rel       <- cfg_get("paths.output", "output")

p_data_proc <- here::here(data_proc_rel)
p_derived   <- file.path(p_data_proc, cfg_get("derived.dir", "derived"))
p_gaps      <- file.path(p_data_proc, "gaps")

p_out     <- here::here(out_rel)
p_figures <- file.path(p_out, "figures")
p_tables  <- file.path(p_out, "tables")

dir.create(p_figures, showWarnings = FALSE, recursive = TRUE)
dir.create(p_tables, showWarnings = FALSE, recursive = TRUE)

# Packages
library(dplyr)
library(readr)
library(ggplot2)
library(stringr)
library(sf)
library(viridis)
```

# 1. Scope and aim

This report assesses **data gaps in GBIF-mediated biodiversity occurrence data for Sweden**, focusing on:

- **Spatial gaps**: where are there no or low records?
- **Temporal gaps**: when are data missing, and how recent are records?
- **Taxonomic gaps**: which taxa (especially threatened taxa) are missing from GBIF-mediated data?

The analysis is based on:
- **GBIF Occurrence Cube outputs**
- **EEA reference grids** (10 km and 50 km)
- **Swedish Red List taxonomy export** as reference for taxonomic gaps

# 2. Input data products used in this report

This report uses the following intermediate outputs:

## Phase 2 (derived summaries)
- `data_proc/derived/cell_summary_10km.csv`
- `data_proc/derived/cell_summary_50km.csv`
- `data_proc/derived/time_summary_10km.csv`
- `data_proc/derived/time_summary_50km.csv`
- `data_proc/derived/species_summary_10km.csv`
- `data_proc/derived/species_summary_50km.csv`

## Phase 3 (gap definitions)
- `data_proc/gaps/spatial_gaps_10km.csv`
- `data_proc/gaps/spatial_gaps_50km.csv`
- `data_proc/gaps/cell_recency_10km.csv`
- `data_proc/gaps/cell_recency_50km.csv`
- `data_proc/gaps/taxonomic_gap_summary.csv`
- `data_proc/gaps/missing_threatened_taxa.csv` (if available)

## Phase 4 (overview tables)
These can optionally be created via:

```r
source("scripts/10_make_phase4_overview.R")
```

# 3. Gap definitions (summary)

Gap metric definitions are documented in:

- `docs/metrics.md`

In short, this report uses:

## Spatial gaps
- **Zero coverage gap:** `occurrences == 0`
- **Low coverage gap:** among cells with `occurrences > 0`, the lowest 5% (`q05`) and 10% (`q10`)

## Temporal gaps
- **Recency per cell:** last observation month per cell
- **Staleness thresholds:** no records in last 12 months / last 5 years

## Taxonomic gaps
- Taxa from the Swedish reference list are considered **missing** if no match exists in cube taxa (string-matching approach)

# 4. Spatial gaps

## 4.1 Spatial gap summary (10 km)

```{r}
sp10 <- read_csv(file.path(p_gaps, "spatial_gaps_10km.csv"), show_col_types = FALSE)

sp10_all <- sp10 %>% filter(basisofrecord == "all")

sp10_all %>%
  summarise(
    n_cells = n(),
    pct_zero = round(100 * mean(gap_zero, na.rm = TRUE), 2),
    pct_low_q05 = round(100 * mean(gap_low_q05, na.rm = TRUE), 2),
    pct_low_q10 = round(100 * mean(gap_low_q10, na.rm = TRUE), 2)
  )
```

## 4.2 Spatial gap summary (50 km)

```{r}
sp50 <- read_csv(file.path(p_gaps, "spatial_gaps_50km.csv"), show_col_types = FALSE)

sp50_all <- sp50 %>% filter(basisofrecord == "all")

sp50_all %>%
  summarise(
    n_cells = n(),
    pct_zero = round(100 * mean(gap_zero, na.rm = TRUE), 2),
    pct_low_q05 = round(100 * mean(gap_low_q05, na.rm = TRUE), 2),
    pct_low_q10 = round(100 * mean(gap_low_q10, na.rm = TRUE), 2)
  )
```

## 4.3 Spatial coverage intensity maps (quick check)

This section joins spatial gap outputs to the processed grid geometries and plots a basic intensity map.

### 10 km grid intensity

```{r}
grid10_path <- file.path(p_data_proc, "grids_10km.gpkg")
stopifnot(file.exists(grid10_path))

g10 <- st_read(grid10_path, quiet = TRUE)

# Guess join field containing the cell code
guess_code_field <- function(nms) {
  n <- tolower(nms)
  cand <- nms[grepl("eea", n) & grepl("code", n)]
  if (length(cand) > 0) return(cand[1])
  cand <- nms[grepl("cell", n) & grepl("code", n)]
  if (length(cand) > 0) return(cand[1])
  cand <- nms[grepl("eea|cell|code|grid", n)]
  if (length(cand) > 0) return(cand[1])
  NA_character_
}

code_field_10 <- guess_code_field(names(g10))
stopifnot(!is.na(code_field_10))

g10$eeacellcode <- as.character(g10[[code_field_10]])

map10 <- g10 %>%
  left_join(sp10_all %>% select(eeacellcode, occurrences, log_occ, gap_zero), by = "eeacellcode") %>%
  mutate(
    occurrences = coalesce(occurrences, 0),
    log_occ = coalesce(log_occ, 0),
    gap_zero = coalesce(gap_zero, TRUE)
  )


ggplot(map10) +
  geom_sf(aes(fill = occurrences), linewidth = 0) +
  scale_fill_viridis(option = "B") +
  labs(
    title = "Spatial coverage intensity (10 km)",
    subtitle = "Fill = occurrences, basisOfRecord = all",
    fill = "occurrences"
  )

ggplot(map10) +
  geom_sf(aes(fill = log_occ), linewidth = 0) +
  scale_fill_viridis(option = "B") +
  labs(
    title = "Spatial coverage intensity (10 km)",
    subtitle = "Fill = log10(occurrences + 1), basisOfRecord = all",
    fill = "log10(occ+1)"
  )
```

### 50 km grid intensity

```{r}
# Load 50 km grid geometry (Europe-wide)
grid50_path <- file.path(p_data_proc, "grids_50km.gpkg")
stopifnot(file.exists(grid50_path))
g50 <- st_read(grid50_path, quiet = TRUE)


code_field_50 <- guess_code_field(names(g50))
stopifnot(!is.na(code_field_50))

g50$eeacellcode <- as.character(g50[[code_field_50]])

# Reuse guess_code_field() from earlier in the Rmd
code_field_50 <- guess_code_field(names(g50))
stopifnot(!is.na(code_field_50))

# Load spatial gaps (should already be Sweden-domain after Script 07 fix)
sp50_all <- read_csv(file.path(p_gaps, "spatial_gaps_50km.csv"), show_col_types = FALSE) %>%
  filter(basisofrecord == "all")

stopifnot("eeacellcode" %in% names(sp50_all))

# Filter geometry to Sweden-domain codes (for plotting only)
sweden_codes_50 <- unique(sp50_all$eeacellcode)

g50_se <- g50 %>%
  mutate(eeacellcode = as.character(.data[[code_field_50]])) %>%
  filter(eeacellcode %in% sweden_codes_50)

# Join + safe NAs
map50 <- g50_se %>%
  left_join(
    sp50_all %>% select(eeacellcode, occurrences, log_occ, gap_zero, gap_low_q05, gap_low_q10),
    by = "eeacellcode"
  ) %>%
  mutate(
    occurrences = coalesce(occurrences, 0),
    log_occ = coalesce(log_occ, 0),
    gap_zero = coalesce(gap_zero, TRUE),
    gap_low_q05 = coalesce(gap_low_q05, FALSE),
    gap_low_q10 = coalesce(gap_low_q10, FALSE)
  )

# Auto-zoom to Sweden-domain extent
bb50 <- st_bbox(map50)


ggplot(map50) +
  geom_sf(aes(fill = occurrences), linewidth = 0) +
  scale_fill_viridis(option = "B") +
  coord_sf(xlim = c(bb50["xmin"], bb50["xmax"]),
           ylim = c(bb50["ymin"], bb50["ymax"])) +
  labs(
    title = "Spatial coverage intensity (50 km)",
    subtitle = "Fill = occurrences, basisOfRecord = all",
    fill = "occurrences"
  )

ggplot(map50) +
  geom_sf(aes(fill = log_occ), linewidth = 0) +
  scale_fill_viridis(option = "B") +
  coord_sf(xlim = c(bb50["xmin"], bb50["xmax"]),
           ylim = c(bb50["ymin"], bb50["ymax"])) +
  labs(
    title = "Spatial coverage intensity (50 km)",
    subtitle = "Fill = log10(occurrences + 1), basisOfRecord = all",
    fill = "log10(occ+1)"
  )
```

## 4.4 Zero-coverage gap map

### 10 km grid intensity
```{r}
# Zero coverage (10 km): TRUE = no occurrences in cell
ggplot(map10) +
  geom_sf(aes(fill = gap_zero), linewidth = 0) +
  scale_fill_viridis_d(option = "B") +
  labs(
    title = "Spatial gap map (10 km): zero coverage",
    subtitle = "Cells with no GBIF-mediated occurrences (basisOfRecord = all)",
    fill = "Gap (zero)"
  )
```


### 50 km grid intensity
```{r}
# Zero coverage (50 km): TRUE = no occurrences in cell

ggplot(map50) +
  geom_sf(aes(fill = gap_zero), linewidth = 0) +
  scale_fill_viridis_d(option = "B") +
  coord_sf(xlim = c(bb50["xmin"], bb50["xmax"]),
           ylim = c(bb50["ymin"], bb50["ymax"])) +
  labs(
    title = "Spatial gap map (50 km): zero coverage",
    subtitle = "Cells with no GBIF-mediated occurrences (basisOfRecord = all)",
    fill = "Gap (zero)"
  )
```

## 4.5 Low-coverage gap maps: q05 and q10

### 10 km grid intensity

```{r}
# Attach low-coverage flags from spatial_gaps_10km to the geometry
map10_gaps <- g10 %>%
  left_join(
    sp10_all %>% select(eeacellcode, gap_low_q05, gap_low_q10),
    by = "eeacellcode"
  ) %>%
  mutate(
    gap_low_q05 = coalesce(gap_low_q05, FALSE),
    gap_low_q10 = coalesce(gap_low_q10, FALSE)
  )

# Low coverage = lowest 5% among non-zero cells
ggplot(map10_gaps) +
  geom_sf(aes(fill = gap_low_q05), linewidth = 0) +
  scale_fill_viridis_d(option = "B") +
  labs(
    title = "Spatial gap map (10 km): low coverage (q05)",
    subtitle = "Cells in the lowest 5% of non-zero occurrence counts (basisOfRecord = all)",
    fill = "Low (q05)"
  )

# Low coverage = lowest 10% among non-zero cells
ggplot(map10_gaps) +
  geom_sf(aes(fill = gap_low_q10), linewidth = 0) +
  scale_fill_viridis_d(option = "B") +
  
  labs(
    title = "Spatial gap map (10 km): low coverage (q10)",
    subtitle = "Cells in the lowest 10% of non-zero occurrence counts (basisOfRecord = all)",
    fill = "Low (q10)"
  )
```


### 50 km grid intensity

```{r}
# Attach low-coverage flags from spatial_gaps_10km to the geometry
map50_gaps <- g50 %>%
  left_join(
    sp50_all %>% select(eeacellcode, gap_low_q05, gap_low_q10),
    by = "eeacellcode"
  ) %>%
  mutate(
    gap_low_q05 = coalesce(gap_low_q05, FALSE),
    gap_low_q10 = coalesce(gap_low_q10, FALSE)
  )

# Low coverage = lowest 5% among non-zero cells
ggplot(map50_gaps) +
  geom_sf(aes(fill = gap_low_q05), linewidth = 0) +
  scale_fill_viridis_d(option = "B") +
  coord_sf(xlim = c(bb50["xmin"], bb50["xmax"]),
           ylim = c(bb50["ymin"], bb50["ymax"])) +
  labs(
    title = "Spatial gap map (50 km): low coverage (q05)",
    subtitle = "Cells in the lowest 5% of non-zero occurrence counts (basisOfRecord = all)",
    fill = "Low (q05)"
  )

# Low coverage = lowest 50% among non-zero cells
ggplot(map50_gaps) +
  geom_sf(aes(fill = gap_low_q10), linewidth = 0) +
  scale_fill_viridis_d(option = "B") +
  coord_sf(xlim = c(bb50["xmin"], bb50["xmax"]),
           ylim = c(bb50["ymin"], bb50["ymax"])) +
  labs(
    title = "Spatial gap map (10 km): low coverage (q10)",
    subtitle = "Cells in the lowest 10% of non-zero occurrence counts (basisOfRecord = all)",
    fill = "Low (q10)"
  )
```
# 5. Temporal gaps

## 5.1 National temporal trend (occurrences per year)

```{r}
ty10 <- read_csv(file.path(p_gaps, "temporal_overview_year_10km.csv"), show_col_types = FALSE)

ggplot(ty10, aes(x = year, y = occurrences)) +
  geom_line() +
  labs(
    title = "Temporal coverage (10 km): occurrences per year",
    x = "Year",
    y = "Occurrences"
  )
```

## 5.2 Seasonal bias (occurrences per month)

```{r}
tm10 <- read_csv(file.path(p_gaps, "temporal_overview_month_10km.csv"), show_col_types = FALSE)

ggplot(tm10, aes(x = month, y = occurrences)) +
  geom_col() +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Seasonality (10 km): occurrences per month",
    x = "Month",
    y = "Occurrences"
  )
```

## 5.3 Recency / staleness per grid cell (10 km)

```{r}
rec10 <- read_csv(file.path(p_gaps, "cell_recency_10km.csv"), show_col_types = FALSE) %>%
  filter(basisofrecord == "all")

rec10 %>%
  summarise(
    n_cells = n(),
    pct_stale_12m = round(100 * mean(gap_stale_12m, na.rm = TRUE), 2),
    pct_stale_5y = round(100 * mean(gap_stale_5y, na.rm = TRUE), 2),
    median_staleness = median(staleness_months, na.rm = TRUE)
  )
```

# 6. Taxonomic gaps

## 6.1 Summary by rank and threat status

```{r}
tax_sum <- read_csv(file.path(p_gaps, "taxonomic_gap_summary.csv"), show_col_types = FALSE)

tax_sum %>% arrange(taxonRank, threatStatus) %>% head(30)
```

## 6.2 Missing threatened taxa (CR/EN/VU)

```{r}
miss_thr_path <- file.path(p_gaps, "missing_threatened_taxa.csv")

if (file.exists(miss_thr_path)) {
  miss_thr <- read_csv(miss_thr_path, show_col_types = FALSE)
  miss_thr %>% select(scientificName, taxonRank, threatStatus) %>% head(50)
} else {
  cat("No missing_threatened_taxa.csv found.
")
}
```

# 7. Notes and limitations

- GBIF Occurrence Cube data represent **aggregated occurrence counts**, not true sampling effort.
- Spatial, temporal, and taxonomic coverage can be influenced by observer bias, accessibility, mobilization effort, and data publishing practices.
- Taxonomic matching in this phase uses **scientific name string matching**, which may undercount matches when synonyms or alternate spellings occur.

# 8. Next steps

Recommended next steps include:
- basisOfRecord comparisons (humanObservation vs preservedSpecimen)
- enhanced taxonomic mapping via GBIF backbone resolution
- interpretation of gap patterns in relation to land cover, protected areas, or population density (optional)
