---
title: "GBIF Sweden Gap Analysis - Exploratory Analysis"
subtitle: "Diagnostic plots and multi-dimensional exploration"
author: "GBIF Sweden Data Gap Analysis"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    fig_width: 12
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 300
)

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(viridis)
library(patchwork)
library(sf)
library(knitr)
library(data.table)

theme_set(theme_minimal(base_size = 13))

p_gaps <- here("data_proc", "gaps")
p_tables <- here("output", "tables")
p_integrated <- here("output", "tables", "integrated")
p_data_proc <- here("data_proc")
```

# Overview

Exploratory analysis of GBIF Sweden data gaps across spatial, temporal, and taxonomic dimensions.

---

# 1. Dashboard Summary

```{r dashboard}
dashboard <- read_csv(here(p_tables, "dashboard_summary.csv"), show_col_types = FALSE)

cat("## Overall Coverage Metrics\n\n")
cat("**Spatial Coverage:**\n")
cat("- 10km grid:", comma(dashboard$cells_10km_with_data), "/", comma(dashboard$cells_10km_total), 
    "cells (", dashboard$cells_10km_pct_coverage, "%)\n")
cat("- 50km grid:", comma(dashboard$cells_50km_with_data), "/", comma(dashboard$cells_50km_total), 
    "cells (", dashboard$cells_50km_pct_coverage, "%)\n\n")

cat("**Temporal Coverage:**\n")
cat("- Year range:", dashboard$year_min, "-", dashboard$year_max, 
    "(", dashboard$year_range, "years)\n")
cat("- Median staleness:", dashboard$median_staleness_months_10km, "months\n")
cat("- Cells >5 years stale:", dashboard$pct_stale_5y_10km, "%\n\n")

cat("**Taxonomic Coverage:**\n")
cat("- Taxa in reference:", comma(dashboard$taxa_in_reference), "\n")
cat("- Taxa in GBIF:", comma(dashboard$taxa_in_gbif), 
    "(", dashboard$taxa_pct_coverage, "%)\n\n")

cat("**Priority Taxa:**", comma(dashboard$n_priority_taxa), "\n\n")
cat("*Analysis date:", dashboard$analysis_date, "*\n\n")

# Simple bar chart
tibble(
  dimension = c("10km Spatial", "50km Spatial", "Taxonomic"),
  coverage = c(dashboard$cells_10km_pct_coverage, 
               dashboard$cells_50km_pct_coverage, 
               dashboard$taxa_pct_coverage)
) %>%
  ggplot(aes(x = dimension, y = coverage, fill = dimension)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(coverage, 1), "%")), 
            vjust = -0.5, size = 6, fontface = "bold") +
  scale_fill_viridis_d() +
  ylim(0, 110) +
  labs(title = "Coverage Across Dimensions", x = NULL, y = "Coverage (%)") +
  theme(legend.position = "none")
```

---

# 2. Spatial Coverage

## 2.1 Grid Resolution Comparison

```{r spatial-comparison}
comp_grids <- read_csv(here(p_tables, "comparison_grid_resolutions.csv"), show_col_types = FALSE)

kable(comp_grids, caption = "Grid Resolution Comparison", digits = 1, format.args = list(big.mark = ","))

comp_grids %>%
  select(grid_resolution, n_cells_total, n_cells_with_data) %>%
  pivot_longer(cols = c(n_cells_total, n_cells_with_data), names_to = "type", values_to = "count") %>%
  mutate(type = recode(type, n_cells_total = "Total", n_cells_with_data = "With Data")) %>%
  ggplot(aes(x = grid_resolution, y = count, fill = type)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = comma(count)), position = position_dodge(width = 0.6), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("Total" = "grey70", "With Data" = "#1976D2")) +
  scale_y_continuous(labels = comma) +
  labs(title = "Cell Counts by Grid Resolution", x = NULL, y = "Number of Cells", fill = NULL) +
  theme(legend.position = "bottom")
```

## 2.2 Coverage by Basis of Record

```{r spatial-basis}
spatial_overview <- read_csv(here(p_tables, "overview_spatial_gap_rates.csv"), show_col_types = FALSE)

# 10km grid, exclude "all"
spatial_10km <- spatial_overview %>%
  filter(grid == "grid10km", basisofrecord != "all") %>%
  arrange(desc(total_occurrences))

spatial_10km %>%
  ggplot(aes(x = fct_reorder(basisofrecord, total_occurrences), y = pct_zero)) +
  geom_col(fill = "#FF6F00") +
  geom_text(aes(label = paste0(round(pct_zero, 1), "%")), hjust = -0.2, size = 4) +
  coord_flip() +
  labs(title = "Zero Coverage Rates by Basis of Record (10km Grid)",
       subtitle = "Percentage of cells with zero occurrences",
       x = "Basis of Record", y = "Zero Coverage (%)",
       caption = "Lower is better")
```

## 2.3 Geographic Coverage Map

```{r spatial-map, fig.height=10, fig.width=12}
# Load grid
grid_10km <- st_read(here(p_data_proc, "grids_10km.gpkg"), quiet = TRUE)

# Load gaps data
spatial_gaps <- fread(here(p_gaps, "spatial_gaps_10km.csv"))

# Find cellcode field in grid
cellcode_candidates <- names(grid_10km)[grepl("cellcode|eeacell", tolower(names(grid_10km)))]
if (length(cellcode_candidates) > 0) {
  cellcode_field <- cellcode_candidates[1]
} else {
  # Try other patterns
  cellcode_field <- names(grid_10km)[grepl("code", tolower(names(grid_10km)))][1]
}

# Join data
grid_10km <- grid_10km %>%
  mutate(eeacellcode = as.character(.data[[cellcode_field]]))

# Get "all" basis data
gaps_all <- spatial_gaps %>%
  filter(basisofrecord == "all") %>%
  select(eeacellcode, occurrences, gap_zero, log_occ)

# Join and categorize
grid_map <- grid_10km %>%
  left_join(gaps_all, by = "eeacellcode") %>%
  mutate(
    coverage_cat = case_when(
      is.na(occurrences) | gap_zero ~ "Zero",
      occurrences <= 10 ~ "Very Low (1-10)",
      occurrences <= 100 ~ "Low (11-100)",
      occurrences <= 1000 ~ "Medium (101-1k)",
      TRUE ~ "High (>1k)"
    ),
    coverage_cat = factor(coverage_cat, levels = c(
      "Zero", "Very Low (1-10)", "Low (11-100)", "Medium (101-1k)", "High (>1k)"
    ))
  )

ggplot(grid_map) +
  geom_sf(aes(fill = coverage_cat), color = "white", linewidth = 0.05) +
  scale_fill_manual(
    values = c("Zero" = "#D32F2F", "Very Low (1-10)" = "#FF9800", 
               "Low (11-100)" = "#FDD835", "Medium (101-1k)" = "#66BB6A", 
               "High (>1k)" = "#1976D2"),
    na.value = "grey90", drop = FALSE
  ) +
  labs(title = "Geographic Distribution of Occurrence Data (10km Grid)", 
       fill = "Coverage Category") +
  theme_void() +
  theme(
    legend.position = "right",
    legend.key.height = unit(1, "cm"),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold", margin = margin(b = 10))
  )
```

---

# 3. Temporal Coverage

## 3.1 Annual Trends

```{r temporal-year}
temporal_year <- read_csv(here(p_tables, "overview_temporal_year.csv"), show_col_types = FALSE)

temporal_year %>%
  ggplot(aes(x = year, y = total_occurrences, color = grid)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2, alpha = 0.6) +
  scale_y_continuous(labels = comma) +
  scale_color_viridis_d(option = "C", labels = c("10km Grid", "50km Grid")) +
  labs(title = "Temporal Trend in Occurrence Records",
       subtitle = "Annual totals across all basis of record",
       x = "Year", y = "Total Occurrences", color = NULL) +
  theme(legend.position = "bottom")
```

## 3.2 Seasonal Patterns

```{r temporal-month}
temporal_month <- read_csv(here(p_tables, "overview_temporal_month.csv"), show_col_types = FALSE)

temporal_month %>%
  ggplot(aes(x = month, y = total_occurrences, fill = grid)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = comma) +
  scale_fill_viridis_d(option = "D", labels = c("10km Grid", "50km Grid")) +
  labs(title = "Seasonal Sampling Pattern",
       subtitle = "Total occurrences by month (all years combined)",
       x = "Month", y = "Total Occurrences", fill = NULL) +
  theme(legend.position = "bottom")
```

## 3.3 Data Recency by Basis of Record

```{r recency-all}
recency_overview <- read_csv(here(p_tables, "overview_temporal_recency_rates.csv"), show_col_types = FALSE)

# Show ALL basis types for 10km (including "all")
recency_10km <- recency_overview %>%
  filter(grid == "grid10km") %>%
  arrange(desc(pct_stale_5y))

recency_10km %>%
  ggplot(aes(x = fct_reorder(basisofrecord, pct_stale_5y), y = pct_stale_5y)) +
  geom_col(fill = "#D32F2F", alpha = 0.8) +
  geom_text(aes(label = paste0(round(pct_stale_5y, 1), "%")), hjust = -0.2, size = 4) +
  coord_flip() +
  labs(title = "Data Staleness by Basis of Record (10km Grid)",
       subtitle = "Percentage of cells not sampled in the last 5 years",
       x = "Basis of Record", y = "Percentage Stale (%)",
       caption = "Lower values indicate more recent data")
```

---

# 4. Taxonomic Coverage

## 4.1 Coverage by Taxonomic Rank

```{r taxonomic-rank}
tax_summary <- read_csv(here(p_tables, "overview_taxonomic_summary.csv"), show_col_types = FALSE)

tax_summary %>%
  filter(!is.na(taxonRank), n_ref_total >= 10) %>%
  arrange(desc(n_ref_total)) %>%
  slice_head(n = 15) %>%
  kable(caption = "Taxonomic Coverage by Rank (Top 15 by reference count)", digits = 1)

tax_summary %>%
  filter(!is.na(taxonRank), n_ref_total >= 10) %>%
  arrange(desc(pct_coverage)) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = fct_reorder(taxonRank, pct_coverage), y = pct_coverage)) +
  geom_col(aes(fill = pct_coverage)) +
  geom_text(aes(label = paste0(round(pct_coverage, 1), "%")), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_fill_viridis_c(option = "plasma") +
  labs(title = "Taxonomic Coverage by Rank (Top 15 by coverage)",
       x = "Taxonomic Rank", y = "Coverage (%)") +
  theme(legend.position = "none")
```

## 4.2 Basis of Record Comparison

```{r basis-comparison}
comp_basis <- read_csv(here(p_tables, "comparison_basis_types.csv"), show_col_types = FALSE)

comp_basis %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = fct_reorder(basisofrecord, total_occurrences), y = total_occurrences)) +
  geom_col(fill = "#1976D2") +
  geom_text(aes(label = comma(total_occurrences)), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Total Occurrences by Basis of Record (Top 10)",
       x = "Basis of Record", y = "Total Occurrences")
```

---

# 5. Priority Areas for Action

## 5.1 Zero Coverage Cells

```{r priority-zero}
priority_zero <- fread(here(p_integrated, "priority_zero_coverage_cells.csv"))

if (nrow(priority_zero) > 0) {
  cat("**", comma(nrow(priority_zero)), "cells with zero coverage across all basis types**\n\n")
  
  priority_zero %>%
    arrange(grid, eeacellcode) %>%
    slice_head(n = 30) %>%
    kable(caption = "Top 30 Zero Coverage Cells (Priority for Sampling)")
} else {
  cat("**No cells with zero coverage found** - excellent spatial coverage!\n")
}
```

## 5.2 Stale Cells (Not Sampled in 5+ Years)

```{r priority-stale}
priority_stale <- fread(here(p_integrated, "priority_stale_cells.csv"))

cat("**", comma(nrow(priority_stale)), "cells not sampled in 5+ years**\n\n")

# Distribution of staleness
priority_stale %>%
  ggplot(aes(x = staleness_months / 12)) +
  geom_histogram(bins = 40, fill = "#D32F2F", alpha = 0.7) +
  scale_x_continuous(labels = comma) +
  labs(title = "Distribution of Data Staleness",
       subtitle = paste("Median:", round(median(priority_stale$staleness_months / 12), 1), "years"),
       x = "Years Since Last Sample", y = "Number of Cells")

# Top stale cells
priority_stale %>%
  arrange(desc(staleness_months)) %>%
  mutate(years_stale = round(staleness_months / 12, 1)) %>%
  select(grid, eeacellcode, last_ym, years_stale, total_occurrences) %>%
  slice_head(n = 20) %>%
  kable(caption = "Top 20 Stalest Cells")
```

## 5.3 Priority Threatened Taxa

```{r priority-taxa}
priority_taxa <- fread(here(p_integrated, "priority_undersampled_taxa.csv"))

cat("**", comma(nrow(priority_taxa)), "priority taxa requiring attention**\n\n")

# Summary by threat status
priority_taxa %>%
  filter(!is.na(threatStatus)) %>%
  count(threatStatus) %>%
  arrange(match(threatStatus, c("CR", "EN", "VU", "NT"))) %>%
  kable(caption = "Priority Taxa by Threat Status", col.names = c("Threat Status", "Count"))

# Top priority species (CR and EN)
priority_taxa %>%
  filter(threatStatus %in% c("CR", "EN")) %>%
  arrange(threatStatus, scientificName) %>%
  select(scientificName, threatStatus, taxonRank, family, priority) %>%
  slice_head(n = 25) %>%
  kable(caption = "Top 25 Critically Endangered and Endangered Priority Taxa")
```

---

# 6. Multi-Dimensional Patterns

## 6.1 Spatial Coverage Over Time

```{r space-time}
space_time_10 <- fread(here(p_integrated, "space_time_basis_10km.csv"))

# Get top 6 basis types by total occurrences
top_basis <- space_time_10 %>%
  group_by(basisofrecord) %>%
  summarise(total = sum(total_occurrences, na.rm = TRUE)) %>%
  arrange(desc(total)) %>%
  slice_head(n = 6) %>%
  pull(basisofrecord)

# Plot recent decades
space_time_10 %>%
  filter(year >= 1990, basisofrecord %in% top_basis) %>%
  ggplot(aes(x = year, y = n_cells_sampled, color = basisofrecord)) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d() +
  scale_y_continuous(labels = comma) +
  labs(title = "Spatial Coverage Over Time by Basis of Record (10km Grid)",
       subtitle = "Number of cells sampled per year (since 1990, top 6 basis types)",
       x = "Year", y = "Number of Cells Sampled", color = "Basis of Record") +
  theme(legend.position = "bottom")
```

---

# 7. Key Findings

```{r summary-stats}
cat("## Summary Statistics\n\n")

cat("**Spatial:**\n")
cat("- Coverage:", dashboard$cells_10km_pct_coverage, "% (10km),", 
    dashboard$cells_50km_pct_coverage, "% (50km)\n")
cat("- Zero coverage cells:", comma(nrow(priority_zero)), "\n\n")

cat("**Temporal:**\n")
cat("- Year range:", dashboard$year_range, "years (", dashboard$year_min, "-", dashboard$year_max, ")\n")
cat("- Stale cells (>5 years):", comma(nrow(priority_stale)), 
    "(", dashboard$pct_stale_5y_10km, "%)\n\n")

cat("**Taxonomic:**\n")
cat("- Coverage:", dashboard$taxa_pct_coverage, "% (", comma(dashboard$taxa_in_gbif), "/", 
    comma(dashboard$taxa_in_reference), ")\n")
cat("- Priority taxa:", comma(dashboard$n_priority_taxa), "\n\n")

cat("**Data Quality:**\n")
cat("- Basis of record types:", length(unique(spatial_overview$basisofrecord)) - 1, "\n")
cat("- Files analyzed:", length(list.files(p_gaps, pattern = "\\.csv$")), "gap analysis files\n")
```

---

# Session Info

```{r session}
sessionInfo()
```
