---
title: "Data Mobilisation Priorities"
subtitle: "Actionable Targets for Closing Biodiversity Data Gaps"
author: "GBIF Gap Finder"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
  pdf_document:
    toc: true
    latex_engine: xelatex
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)

library(here); library(dplyr); library(tidyr); library(ggplot2)
library(scales); library(sf); library(knitr); library(kableExtra)

data_path <- here("shiny_app", "gap_finder", "data", "shiny_data.rds")
app_data <- readRDS(data_path)
safe_get <- function(name) if (name %in% names(app_data)) app_data[[name]] else NULL

grid_10km        <- safe_get("grid_10km")
admin_level1     <- safe_get("admin_level1")
dashboard        <- safe_get("dashboard")
match_summary    <- safe_get("taxonomic_match_summary")
species_scope    <- safe_get("species_scope_lookup")
spatial_gaps     <- safe_get("spatial_gaps_10km")
# Zero-coverage cells derived the same way as the app's Priorities tab — from
# the 10 km spatial gaps (basisofrecord == "all", no data) — so the count ties
# out to the Overview and Spatial reports (no 10/50 km mixing).
priority_zero    <- if (!is.null(spatial_gaps)) {
  spatial_gaps |> filter(basisofrecord == "all", !has_data) |>
    transmute(eeacellcode, n_species = 0L, occurrences = 0,
              priority_reason = "No GBIF records", priority_level = "Critical")
} else safe_get("priority_zero_cells")
priority_stale   <- safe_get("priority_stale_cells")
overview_ly      <- safe_get("overview_last_year")
publisher_cell_dep <- safe_get("publisher_cell_dependency")
troudet_order    <- safe_get("troudet_bias_order")
tax_by_family    <- safe_get("tax_by_family")
last_year_label  <- safe_get("recent_label") %||% "last 12 months"

pal <- list(sage = "#2A7F62", slate = "#4477AA", sand = "#CCBB44",
            coral = "#EE6677", plum = "#AA3377")
theme_gaps <- theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "#666", size = 11),
        panel.grid.minor = element_blank(), legend.position = "bottom")
theme_set(theme_gaps)

get_threat_col <- function(ms) {
  tc <- intersect(c("threatStatus", "threatStatus_redlist", "threatStatus_backbone"), names(ms))
  if (length(tc) > 0) tc[1] else NA
}
```

# Overview

This report synthesises the gap analysis into actionable priorities for data mobilisation. It identifies specific spatial cells, taxonomic groups, and institutional gaps that should be targeted to improve GBIF data completeness.

# Recommended Actions

```{r actions}
n_zero  <- if (!is.null(priority_zero)) nrow(priority_zero) else 0
n_stale <- if (!is.null(priority_stale)) nrow(priority_stale) else 0
n_taxa_missing <- if (!is.null(match_summary)) sum(!match_summary$matched_any, na.rm = TRUE) else 0

n_cr_en <- 0
if (!is.null(match_summary)) {
  tc <- get_threat_col(match_summary)
  if (!is.na(tc)) n_cr_en <- match_summary |> filter(.data[[tc]] %in% c("CR", "EN"), !matched_any) |> nrow()
}

n_sensitive <- if (!is.null(match_summary) && "is_sensitive" %in% names(match_summary))
  sum(match_summary$is_sensitive, na.rm = TRUE) else 0

n_single_pub <- if (!is.null(publisher_cell_dep)) sum(publisher_cell_dep$n_publishers == 1) else 0
n_cells <- if (!is.null(grid_10km)) nrow(grid_10km) else 0

actions <- tibble(
  Priority = c(
    "Spatial: Fill zero-coverage cells",
    "Temporal: Resurvey stale cells",
    "Taxonomic: Close species coverage gaps",
    "Threatened: Monitor CR and EN species",
    "Sensitive: Assess data availability",
    "Infrastructure: Diversify data sources"
  ),
  Detail = c(
    paste0(comma(n_zero), " grid cells have never been surveyed."),
    paste0(comma(n_stale), " cells have not been surveyed in over 5 years."),
    paste0(comma(n_taxa_missing), " backbone species have no GBIF records."),
    paste0(comma(n_cr_en), " CR/EN species lack any GBIF occurrence data."),
    paste0(comma(n_sensitive), " species have restricted coordinates (5–50 km generalisation)."),
    paste0(comma(n_single_pub), " of ", comma(n_cells), " grid cells depend on a single publisher.")
  )
)

kable(actions, align = c("l", "l")) |>
  kable_styling(full_width = FALSE) |>
  column_spec(1, bold = TRUE, width = "15em")
```

# Next 12 Months

Based on performance in the last 12 months (`r last_year_label`), projected targets at 1.5× the recent rate:

```{r next-12}
ly_occ <- if (!is.null(overview_ly)) overview_ly$occ_last_year else 0
ly_cells <- if (!is.null(overview_ly)) overview_ly$cells_active_last_year else 0
ly_new <- if (!is.null(overview_ly)) overview_ly$cells_newly_covered else 0
mult <- 1.5

targets <- tibble(
  Metric = c("New occurrence records", "Cells with active recording",
             "Newly covered cells", "CR/EN species with new records",
             "Single-publisher cells diversified"),
  `Achieved` = c(comma(ly_occ), comma(ly_cells), comma(ly_new), "—", "—"),
  `Target (1.5×)` = c(
    paste0(comma(ceiling(ly_occ * mult)), "+"),
    paste0(comma(ceiling(ly_cells * mult)), "+"),
    paste0(comma(ceiling(ly_new * mult)), " / ", comma(n_zero), " remaining"),
    paste0("Target: ", comma(min(n_cr_en, 20)), " of ", comma(n_cr_en)),
    paste0("Target: ", comma(min(ceiling(n_single_pub * 0.1), 50)), " of ", comma(n_single_pub))
  )
)

kable(targets, align = c("l", "r", "r")) |> kable_styling(full_width = FALSE)
```

# Zero Coverage Cells

```{r zero-map, fig.height=9}
if (!is.null(grid_10km) && !is.null(priority_zero) && nrow(priority_zero) > 0) {
  zero_sf <- grid_10km |> filter(eeacellcode %in% priority_zero$eeacellcode)

  p <- ggplot() +
    geom_sf(data = grid_10km, fill = "#f0f0f0", color = NA) +
    geom_sf(data = zero_sf, fill = pal$coral, color = pal$coral, linewidth = 0.3) +
    labs(title = paste0("Zero Coverage Cells (", nrow(zero_sf), " cells)"),
         subtitle = "Grid cells with no GBIF occurrence records") +
    theme_void()
  if (!is.null(admin_level1)) {
    p <- p + geom_sf(data = st_transform(admin_level1, st_crs(grid_10km)),
                     fill = NA, color = "#888", linewidth = 0.4)
  }
  print(p)
}
```

```{r zero-table}
if (!is.null(priority_zero) && nrow(priority_zero) > 0) {
  kable(priority_zero |> head(20), align = "l") |> kable_styling(full_width = FALSE, font_size = 11)
}
```

# Stale Cells

```{r stale-map, fig.height=9}
if (!is.null(grid_10km) && !is.null(priority_stale) && nrow(priority_stale) > 0) {
  stale_sf <- grid_10km |>
    inner_join(priority_stale |>
      mutate(years_stale = staleness_months / 12) |>
      select(eeacellcode, years_stale), by = "eeacellcode")

  p <- ggplot() +
    geom_sf(data = grid_10km, fill = "#f0f0f0", color = NA) +
    geom_sf(data = stale_sf, aes(fill = years_stale), color = NA) +
    scale_fill_gradient(low = pal$sand, high = "#114477", name = "Years stale") +
    labs(title = paste0("Stale Cells (", nrow(stale_sf), " cells >5 years old)"),
         subtitle = "Grid cells not surveyed in the last 5 years") +
    theme_void() + theme(legend.position = "right")
  if (!is.null(admin_level1)) {
    p <- p + geom_sf(data = st_transform(admin_level1, st_crs(grid_10km)),
                     fill = NA, color = "#888", linewidth = 0.4)
  }
  print(p)
}
```

# Taxonomic Mobilisation Targets

## Under-Sampled Orders

```{r undersampled-orders, fig.height=7}
if (!is.null(troudet_order)) {
  df <- troudet_order |>
    filter(bias < 0) |> arrange(bias) |> head(15) |>
    mutate(gap = n_known_species - n_in_gbif)

  ggplot(df, aes(x = reorder(order, -bias), y = gap)) +
    geom_col(fill = pal$coral) +
    geom_text(aes(label = gap), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(title = "Most Under-Sampled Orders",
         subtitle = "Species missing from GBIF (negative Troudet bias)",
         x = "", y = "Species missing")
}
```

## Under-Sampled Families

```{r undersampled-families, fig.height=7}
if (!is.null(tax_by_family)) {
  df <- tax_by_family |>
    filter(n_missing > 0) |> arrange(desc(n_missing)) |> head(15)

  ggplot(df, aes(x = reorder(family, n_missing), y = n_missing)) +
    geom_col(fill = pal$sand) +
    geom_text(aes(label = n_missing), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(title = "Most Under-Sampled Families",
         x = "", y = "Species missing from GBIF")
}
```

# Missing Threatened Species

```{r missing-threatened}
if (!is.null(match_summary)) {
  tc <- get_threat_col(match_summary)
  if (!is.na(tc)) {
    missing <- match_summary |>
      filter(.data[[tc]] %in% c("CR", "EN"), !matched_any) |>
      select(any_of(c("scientificName", tc, "kingdom", "order", "family"))) |>
      arrange(.data[[tc]], order, family)

    if (nrow(missing) > 0) {
      kable(missing |> head(30),
        caption = paste0("CR and EN species missing from GBIF (showing first 30 of ", nrow(missing), ")"),
        align = "l") |>
        kable_styling(full_width = FALSE, font_size = 11)
    }
  }
}
```

---

*Report generated by `gbif_gap_finder` on `r Sys.time()`.*
