---
title: "Spatial Gap Analysis"
subtitle: "GBIF Sweden Data Gap Analysis - Geographic Coverage Assessment"
author: "GBIF Sweden"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    fig_width: 12
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 300
)

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(viridis)
library(patchwork)
library(sf)
library(knitr)
library(data.table)
library(glue)

# Theme
theme_set(
 theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12, color = "grey40"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
)

# Paths
p_gaps <- here("data_proc", "gaps")
p_tables <- here("output", "tables")
p_data_proc <- here("data_proc")
p_derived <- here("data_proc", "derived")
p_integrated <- here("output", "tables", "integrated")
```

# Overview

This report analyzes **spatial gaps** in GBIF Sweden's biodiversity occurrence data, examining geographic coverage patterns across EEA reference grids at 10km and 50km resolutions.

**Key questions addressed:**

- What proportion of Sweden's territory has biodiversity occurrence data?
- Where are the geographic gaps in sampling coverage?
- How does coverage vary by evidence type (basis of record)?
- Which areas should be prioritized for future sampling efforts?

---

# 1. Spatial Coverage Summary

## 1.1 Grid Resolution Comparison

```{r load-comparison}
comp_grids <- read_csv(here(p_tables, "comparison_grid_resolutions.csv"), show_col_types = FALSE)

comp_grids |>
  kable(
    caption = "Spatial Coverage by Grid Resolution",
    digits = 1,
    format.args = list(big.mark = ",")
  )
```

```{r coverage-bars, fig.height=5}
p1 <- comp_grids |>
  ggplot(aes(x = grid_resolution, y = pct_coverage, fill = grid_resolution)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(pct_coverage, 1), "%")),
            vjust = -0.5, size = 6, fontface = "bold") +
  scale_fill_viridis_d(option = "viridis", begin = 0.3, end = 0.7) +
  ylim(0, 110) +
  labs(title = "Coverage Percentage", y = "Coverage (%)", x = NULL) +
 theme(legend.position = "none")

p2 <- comp_grids |>
  select(grid_resolution, n_cells_total, n_cells_with_data) |>
  pivot_longer(cols = c(n_cells_total, n_cells_with_data), 
               names_to = "type", values_to = "count") |>
  mutate(type = recode(type, 
                       n_cells_total = "Total Cells", 
                       n_cells_with_data = "Cells with Data")) |>
  ggplot(aes(x = grid_resolution, y = count, fill = type)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = comma(count)), 
            position = position_dodge(width = 0.6), vjust = -0.5, size = 4) +
  scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.7) +
  scale_y_continuous(labels = comma) +
  labs(title = "Cell Counts", y = "Number of Cells", x = NULL, fill = NULL)

p1 + p2 +
  plot_annotation(
    title = "Spatial Coverage Overview",
    subtitle = "Comparing 10km and 50km EEA reference grids"
  )
```

## 1.2 Occurrence Distribution

```{r occurrence-distribution, fig.height=5}
cell10 <- read_csv(here(p_derived, "cell_summary_10km.csv"), show_col_types = FALSE)

cell10 |>
  filter(basisofrecord == "all", occurrences > 0) |>
  ggplot(aes(x = occurrences)) +
  geom_histogram(bins = 50, fill = viridis(1, option = "viridis", begin = 0.4), 
                 color = "white", alpha = 0.8) +
  scale_x_log10(labels = comma) +
  labs(
    title = "Distribution of Occurrences per Cell",
    subtitle = "10km grid, cells with data only (log scale)",
    x = "Occurrences per Cell",
    y = "Number of Cells"
  )
```

---

# 2. Geographic Distribution Maps

## 2.1 Overall Coverage Map (10km)
 
```{r load-spatial-data}
# Load grid
grid_10km <- st_read(here(p_data_proc, "grids_10km.gpkg"), quiet = TRUE)

# Load gaps data
spatial_gaps <- fread(here(p_gaps, "spatial_gaps_10km.csv"))

# Find cellcode field
cellcode_candidates <- names(grid_10km)[grepl("cellcode|eeacell", tolower(names(grid_10km)))]
cellcode_field <- if(length(cellcode_candidates) > 0) {
  cellcode_candidates[1]
} else {
 names(grid_10km)[grepl("code", tolower(names(grid_10km)))][1]
}

# Prepare grid for joining
grid_10km <- grid_10km |>
  mutate(eeacellcode = as.character(.data[[cellcode_field]]))
```

```{r coverage-map, fig.height=12, fig.width=10}
# Get "all" basis data
gaps_all <- spatial_gaps |>
  filter(basisofrecord == "all") |>
  select(eeacellcode, occurrences, gap_zero, log_occ)

# Join and create map data
grid_map <- grid_10km |>
  left_join(gaps_all, by = "eeacellcode") |>
  mutate(
    log_occ = ifelse(is.na(occurrences) | occurrences == 0, NA, log10(occurrences + 1))
  )

ggplot(grid_map) +
  geom_sf(aes(fill = log_occ), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "log10(occurrences)",
    na.value = "grey85",
    guide = guide_colorbar(barwidth = 20, barheight = 0.8)
  ) +
  labs(
    title = "Geographic Distribution of Occurrence Data",
    subtitle = "10km EEA Grid - Sweden",
    caption = "Grey cells indicate zero coverage"
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40"),
    plot.caption = element_text(size = 10, hjust = 1, color = "grey50")
  )
```

## 2.2 Coverage Categories Map

```{r category-map, fig.height=12, fig.width=10}
# Categorize coverage
grid_cat <- grid_map |>
  mutate(
    coverage_cat = case_when(
      is.na(occurrences) | occurrences == 0 ~ "Zero",
      occurrences <= 10 ~ "Very Low (1-10)",
      occurrences <= 100 ~ "Low (11-100)",
      occurrences <= 1000 ~ "Medium (101-1k)",
      occurrences <= 10000 ~ "High (1k-10k)",
      TRUE ~ "Very High (>10k)"
    ),
    coverage_cat = factor(coverage_cat, levels = c(
      "Zero", "Very Low (1-10)", "Low (11-100)", 
      "Medium (101-1k)", "High (1k-10k)", "Very High (>10k)"
    ))
  )

ggplot(grid_cat) +
  geom_sf(aes(fill = coverage_cat), color = "white", linewidth = 0.02) +
  scale_fill_viridis_d(
    option = "plasma",
    direction = -1,
    drop = FALSE
  ) +
  labs(
    title = "Coverage Categories",
    subtitle = "10km grid cells classified by occurrence count",
    fill = "Coverage"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.key.height = unit(1, "cm"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40")
  )
```

---

# 3. Coverage by Basis of Record

## 3.1 Zero Coverage Rates

```{r basis-zero-rates}
spatial_overview <- read_csv(here(p_tables, "overview_spatial_gap_rates.csv"), show_col_types = FALSE)

spatial_10km <- spatial_overview |>
  filter(grid == "grid10km", basisofrecord != "all") |>
  arrange(pct_zero)

# Select columns that exist
available_cols <- intersect(
  c("basisofrecord", "total_cells", "n_cells", "cells_zero", "pct_zero", "total_occurrences"),
  names(spatial_10km)
)

spatial_10km |>
  select(all_of(available_cols)) |>
  mutate(across(where(is.numeric) & !matches("pct"), comma)) |>
  kable(caption = "Zero Coverage by Basis of Record (10km Grid)", digits = 1)
```

```{r basis-zero-plot, fig.height=6}
spatial_10km |>
  ggplot(aes(x = fct_reorder(basisofrecord, -pct_zero), y = pct_zero, fill = pct_zero)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct_zero, 1), "%")), 
            vjust = -0.3, size = 4) +
  scale_fill_viridis_c(option = "rocket", direction = -1, begin = 0.2, end = 0.8) +
  labs(
    title = "Zero Coverage Rate by Evidence Type",
    subtitle = "Percentage of 10km cells with no occurrences",
    x = "Basis of Record",
    y = "Zero Coverage (%)",
    caption = "Lower values indicate better spatial coverage"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

## 3.2 Coverage Comparison Across Basis Types

```{r basis-facet-maps, fig.height=14, fig.width=12}
# Get top 4 basis types by occurrence count
top_basis <- spatial_overview |>
  filter(grid == "grid10km", basisofrecord != "all") |>
  arrange(desc(total_occurrences)) |>
  slice_head(n = 4) |>
  pull(basisofrecord)

# Prepare data for faceted map
gaps_by_basis <- spatial_gaps |>
  filter(basisofrecord %in% top_basis) |>
  select(eeacellcode, basisofrecord, occurrences)

grid_facet <- grid_10km |>
  select(eeacellcode) |>
  crossing(basisofrecord = top_basis) |>
  left_join(gaps_by_basis, by = c("eeacellcode", "basisofrecord")) |>
  mutate(
    log_occ = ifelse(is.na(occurrences) | occurrences == 0, NA, log10(occurrences + 1)),
    basisofrecord = factor(basisofrecord, levels = top_basis)
  )

# Rejoin geometry
grid_facet_sf <- grid_10km |>
  select(eeacellcode) |>
  right_join(grid_facet |> st_drop_geometry(), by = "eeacellcode")

ggplot(grid_facet_sf) +
  geom_sf(aes(fill = log_occ), color = NA) +
  facet_wrap(~basisofrecord, ncol = 2) +
  scale_fill_viridis(
    option = "viridis",
    name = "log10(occ)",
    na.value = "grey85"
  ) +
  labs(
    title = "Coverage by Top 4 Evidence Types",
    subtitle = "Comparing geographic distribution across basis of record"
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "grey40")
  )
```

---

# 4. Priority Areas

## 4.1 Zero Coverage Cells

```{r priority-zero}
if (file.exists(here(p_integrated, "priority_zero_coverage_cells.csv"))) {
 priority_zero <- fread(here(p_integrated, "priority_zero_coverage_cells.csv"))
  
  cat("**", comma(nrow(priority_zero)), "cells** have zero coverage across all basis types.\n\n")
  
  if (nrow(priority_zero) > 0) {
    priority_zero |>
      slice_head(n = 20) |>
      kable(caption = "Sample of Zero Coverage Cells (Top 20)")
  }
} else {
  cat("Priority zero coverage file not found. Run script 11 first.\n")
  priority_zero <- data.frame()
}
```

## 4.2 Priority Cells Map

```{r priority-map, fig.height=12, fig.width=10}
if (nrow(priority_zero) > 0) {
  grid_priority <- grid_10km |>
    mutate(
      is_zero = eeacellcode %in% priority_zero$eeacellcode,
      priority = ifelse(is_zero, "Zero Coverage", "Has Data")
    )
  
  ggplot(grid_priority) +
    geom_sf(aes(fill = priority), color = "white", linewidth = 0.02) +
    scale_fill_manual(
      values = c("Zero Coverage" = "#E63946", "Has Data" = "#2A9D8F"),
      name = "Status"
    ) +
    labs(
      title = "Priority Areas: Zero Coverage Cells",
      subtitle = glue("{comma(nrow(priority_zero))} cells requiring sampling attention"),
      caption = "Red cells have no occurrence records across any evidence type"
    ) +
    theme_void() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40")
    )
} else {
  cat("No zero coverage cells to map.\n")
}
```

## 4.3 Low Coverage Hotspots

```{r low-coverage-analysis, fig.height=6}
# Identify cells with very low coverage (bottom 10%)
threshold <- quantile(cell10$occurrences[cell10$basisofrecord == "all" & cell10$occurrences > 0], 0.1)

low_coverage <- cell10 |>
  filter(basisofrecord == "all", occurrences > 0, occurrences <= threshold)

cat("**Cells with very low coverage (bottom 10%):**\n")
cat("- Threshold:", comma(threshold), "occurrences\n")
cat("- Number of cells:", comma(nrow(low_coverage)), "\n\n")

# Distribution of low coverage cells
cell10 |>
  filter(basisofrecord == "all", occurrences > 0) |>
  mutate(
    coverage_level = case_when(
      occurrences <= threshold ~ "Very Low (bottom 10%)",
      occurrences <= quantile(occurrences, 0.25) ~ "Low (10-25%)",
      occurrences <= quantile(occurrences, 0.50) ~ "Below Median",
      TRUE ~ "Above Median"
    ),
    coverage_level = factor(coverage_level, levels = c(
      "Very Low (bottom 10%)", "Low (10-25%)", "Below Median", "Above Median"
    ))
  ) |>
  count(coverage_level) |>
  ggplot(aes(x = coverage_level, y = n, fill = coverage_level)) +
  geom_col() +
  geom_text(aes(label = comma(n)), vjust = -0.3, size = 4) +
  scale_fill_viridis_d(option = "turbo", direction = -1) +
  labs(
    title = "Distribution of Cells by Coverage Level",
    subtitle = "10km grid, cells with data only",
    x = NULL,
    y = "Number of Cells"
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )
```

---

# 5. Key Findings

```{r summary-stats}
cat("## Spatial Coverage Summary\n\n")

cat("**Grid Coverage:**\n")
for (i in 1:nrow(comp_grids)) {
  cat("- ", comp_grids$grid_resolution[i], ": ", 
      comp_grids$pct_coverage[i], "% coverage (",
      comma(comp_grids$n_cells_with_data[i]), " / ",
      comma(comp_grids$n_cells_total[i]), " cells)\n", sep = "")
}

cat("\n**Priority Areas:**\n")
if (exists("priority_zero") && nrow(priority_zero) > 0) {
  cat("- Zero coverage cells:", comma(nrow(priority_zero)), "\n")
}
cat("- Very low coverage cells (bottom 10%):", comma(nrow(low_coverage)), "\n")

cat("\n**Best Coverage by Evidence Type:**\n")
best_basis <- spatial_10km |> 
  arrange(pct_zero) |> 
  slice_head(n = 3)
for (i in 1:nrow(best_basis)) {
  cat("- ", best_basis$basisofrecord[i], ": ", 
      round(100 - best_basis$pct_zero[i], 1), "% coverage\n", sep = "")
}

cat("\n**Recommendations:**\n")
cat("1. Target zero-coverage cells for systematic biodiversity surveys\n")
cat("2. Prioritize mountain and northern regions if they show lower coverage\n")
cat("3. Leverage citizen science programs to expand geographic reach\n")
```

---

# Session Info

```{r session}
sessionInfo()
```
