---
title: "Spatial Gap Analysis"
subtitle: "GBIF Gap Analysis - Geographic Coverage Assessment"
author: "gbifgaps"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    fig_width: 12
    fig_height: 8
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 3
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 150
)

# Hide code in PDF output, show (with folding) in HTML
if (!knitr::is_html_output()) {
  knitr::opts_chunk$set(echo = FALSE)
}

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(viridis)
library(patchwork)
library(sf)
library(knitr)
library(data.table)
library(glue)
library(lubridate)

# Consistent theme for all plots
theme_gaps <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "grey40"),
    plot.caption = element_text(size = 9, color = "grey50"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 11, face = "bold")
  )

theme_set(theme_gaps)

# Paths
p_gaps <- here("data_proc", "gaps")
p_tables <- here("output", "tables")
p_data_proc <- here("data_proc")
p_derived <- here("data_proc", "derived")
p_integrated <- here("output", "tables", "integrated")

# Admin boundaries (loaded once, used across map chunks)
admin_sf <- tryCatch(
  st_read(here("data_raw", "admin", "admin_level1.gpkg"), quiet = TRUE),
  error = function(e) NULL
)

# Helper function for safe file reading
safe_read <- function(path) {
  if (!file.exists(path)) {
    return(NULL)
  }

  fread(path) |> as_tibble()
}

# Helper for safe sf reading
safe_read_sf <- function(path) {
  if (!file.exists(path)) {
    return(NULL)
  }

  tryCatch(
    st_read(path, quiet = TRUE),
    error = function(e) NULL
  )
}
```

# Overview

This report analyzes **spatial gaps** in the national biodiversity occurrence data, examining geographic coverage patterns across EEA reference grids at 10km and 50km resolutions.

**Key questions addressed:**
 
- What proportion of the study area has biodiversity occurrence data?
- Where are the geographic gaps in sampling coverage?
- How does coverage vary by evidence type (basis of record)?
- Which areas should be prioritized for future sampling efforts?

---

# 1. Spatial Coverage Summary

## 1.1 Grid Resolution Comparison

```{r load-comparison}
comp_grids <- safe_read(here(p_tables, "comparison_grid_resolutions.csv"))

if (!is.null(comp_grids)) {
  comp_grids |>
    kable(
      caption = "Spatial Coverage by Grid Resolution",
      digits = 1,
      format.args = list(big.mark = ",")
    )
}
```

```{r coverage-bars, fig.height=5}
if (!is.null(comp_grids)) {
  # Detect column names dynamically
  grid_col <- if ("grid_resolution" %in% names(comp_grids)) "grid_resolution" else "grid"
  total_col <- if ("n_cells_total" %in% names(comp_grids)) "n_cells_total" else names(comp_grids)[grepl("total", names(comp_grids))][1]
  data_col <- if ("n_cells_with_data" %in% names(comp_grids)) "n_cells_with_data" else names(comp_grids)[grepl("with_data", names(comp_grids))][1]
  
  p1 <- comp_grids |>
    ggplot(aes(x = .data[[grid_col]], y = pct_coverage, fill = .data[[grid_col]])) +
    geom_col(width = 0.6) +
    geom_text(
      aes(label = glue("{round(pct_coverage, 1)}%")),
      vjust = -0.5, size = 6, fontface = "bold"
    ) +
    scale_fill_viridis_d(option = "viridis", begin = 0.3, end = 0.7) +
    ylim(0, 110) +
    labs(title = "Coverage Percentage", y = "Coverage (%)", x = NULL) +
    theme(legend.position = "none")
  
  comp_long <- comp_grids |>
    select(all_of(c(grid_col, total_col, data_col))) |>
    pivot_longer(
      cols = all_of(c(total_col, data_col)),
      names_to = "type",
      values_to = "count"
    ) |>
    mutate(
      type = case_when(
        grepl("total", type, ignore.case = TRUE) ~ "Total Cells",
        grepl("with_data", type, ignore.case = TRUE) ~ "Cells with Data",
        TRUE ~ type
      )
    )
  
  p2 <- comp_long |>
    ggplot(aes(x = .data[[grid_col]], y = count, fill = type)) +
    geom_col(position = "dodge", width = 0.6) +
    geom_text(
      aes(label = comma(count)),
      position = position_dodge(width = 0.6), vjust = -0.5, size = 4
    ) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.7) +
    scale_y_continuous(labels = comma) +
    labs(title = "Cell Counts", y = "Number of Cells", x = NULL, fill = NULL)
  
  p1 + p2 +
    plot_annotation(
      title = "Spatial Coverage Overview",
      subtitle = "Comparing 10km and 50km EEA reference grids"
    )
}
```

## 1.2 Occurrence Distribution

```{r occurrence-distribution, fig.height=5}
cell_summary <- safe_read(here(p_derived, "cell_summary_10km.csv"))

if (!is.null(cell_summary)) {
  cell_all <- cell_summary |>
    filter(basisofrecord == "all", occurrences > 0)
  
  ggplot(cell_all, aes(x = occurrences)) +
    geom_histogram(
      bins = 50,
      fill = viridis(1, option = "viridis", begin = 0.4),
      color = "white",
      alpha = 0.8
    ) +
    scale_x_log10(labels = comma) +
    labs(
      title = "Distribution of Occurrences per Cell",
      subtitle = "10km grid, cells with data only (log scale)",
      x = "Occurrences per Cell",
      y = "Number of Cells"
    )
}
```

---

# 2. Geographic Distribution Maps

## 2.1 Overall Coverage Map (10km)

```{r load-spatial-data}
grid_10km <- safe_read_sf(here(p_data_proc, "grids_10km.gpkg"))
spatial_gaps <- safe_read(here(p_gaps, "spatial_gaps_10km.csv"))

if (!is.null(grid_10km)) {
  # Find cellcode column
 cellcode_candidates <- names(grid_10km)[grepl("cellcode|eeacell", tolower(names(grid_10km)))]
  cellcode_field <- if (length(cellcode_candidates) > 0) {
    cellcode_candidates[1]
  } else {
    names(grid_10km)[grepl("code", tolower(names(grid_10km)))][1]
  }
  
  grid_10km$eeacellcode <- as.character(grid_10km[[cellcode_field]])
}
```

```{r coverage-map, fig.height=12, fig.width=10}
if (!is.null(grid_10km) && !is.null(spatial_gaps)) {
  gaps_all <- spatial_gaps |>
    filter(basisofrecord == "all") |>
    select(eeacellcode, occurrences)
  
  grid_map <- grid_10km |>
    left_join(gaps_all, by = "eeacellcode") |>
    mutate(
      log_occ = case_when(
        is.na(occurrences) ~ NA_real_,
        occurrences == 0 ~ NA_real_,
        TRUE ~ log10(occurrences + 1)
      )
    )
  
  # Admin boundaries
  
  p <- ggplot(grid_map) +
    geom_sf(aes(fill = log_occ), color = NA) +
    scale_fill_viridis(
      option = "viridis",
      name = "log10(occurrences)",
      na.value = "grey85",
      guide = guide_colorbar(barwidth = 20, barheight = 0.8)
    )
  
  if (!is.null(admin_sf)) {
    p <- p + geom_sf(data = admin_sf, fill = NA, color = "#333", linewidth = 0.4)
  }
  
  p + labs(
      title = "Geographic Distribution of Occurrence Data",
      subtitle = "10km EEA Grid — continuous scale",
      caption = "Grey cells indicate zero coverage"
    ) +
    theme_void() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40"),
      plot.caption = element_text(size = 10, hjust = 1, color = "grey50")
    )
}
```

## 2.2 Coverage Categories Map

```{r category-map, fig.height=12, fig.width=10}
if (!is.null(grid_10km) && !is.null(spatial_gaps)) {
  gaps_all <- spatial_gaps |>
    filter(basisofrecord == "all") |>
    select(eeacellcode, occurrences)
  
  # Match app's categorical breaks
  grid_cat <- grid_10km |>
    left_join(gaps_all, by = "eeacellcode") |>
    mutate(
      occ_cat = case_when(
        is.na(occurrences) | occurrences == 0 ~ "No data",
        occurrences <= 100 ~ "1–100",
        occurrences <= 1000 ~ "101–1,000",
        occurrences <= 10000 ~ "1,001–10,000",
        occurrences <= 100000 ~ "10,001–100,000",
        TRUE ~ ">100,000"
      ),
      occ_cat = factor(occ_cat, levels = c(
        "1–100", "101–1,000", "1,001–10,000",
        "10,001–100,000", ">100,000", "No data"
      ))
    )
  
  # App palette: teal → sage → sand → coral → dark red → gray
  occ_colors <- c("1–100" = "#5c7a99", "101–1,000" = "#6b8f71",
                   "1,001–10,000" = "#c4a882", "10,001–100,000" = "#c47a6c",
                   ">100,000" = "#8b3a3a", "No data" = "grey85")
  
  # Load admin boundaries if available
  
  p <- ggplot(grid_cat) +
    geom_sf(aes(fill = occ_cat), color = "white", linewidth = 0.02) +
    scale_fill_manual(values = occ_colors, drop = FALSE, name = "Occurrences")
  
  if (!is.null(admin_sf)) {
    p <- p + geom_sf(data = admin_sf, fill = NA, color = "#333", linewidth = 0.4)
  }
  
  p + labs(title = "Occurrence Density by Grid Cell",
           subtitle = "10km EEA grid — categorical breaks matching app display") +
    theme_void() +
    theme(legend.position = "right", legend.key.height = unit(1, "cm"),
          plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40"))
}
```

## 2.3 Data Recency (Staleness) Map

```{r staleness-map, fig.height=12, fig.width=10}
cell_recency <- safe_read(here(p_gaps, "cell_recency_10km.csv"))

if (!is.null(grid_10km) && !is.null(cell_recency)) {
  grid_stale <- grid_10km |>
    left_join(cell_recency |> filter(basisofrecord == "all") |>
                select(eeacellcode, staleness_months),
              by = "eeacellcode") |>
    mutate(
      stale_cat = case_when(
        is.na(staleness_months) ~ "No data",
        staleness_months <= 12 ~ "< 1 year",
        staleness_months <= 36 ~ "1–3 years",
        staleness_months <= 60 ~ "3–5 years",
        staleness_months <= 120 ~ "5–10 years",
        TRUE ~ "> 10 years"
      ),
      stale_cat = factor(stale_cat, levels = c(
        "< 1 year", "1–3 years", "3–5 years",
        "5–10 years", "> 10 years", "No data"
      ))
    )
  
  stale_colors <- c("< 1 year" = "#5c7a99", "1–3 years" = "#6b8f71",
                     "3–5 years" = "#c4a882", "5–10 years" = "#c47a6c",
                     "> 10 years" = "#8b3a3a", "No data" = "grey85")
  
  p <- ggplot(grid_stale) +
    geom_sf(aes(fill = stale_cat), color = "white", linewidth = 0.02) +
    scale_fill_manual(values = stale_colors, drop = FALSE, name = "Data recency")
  
  if (!is.null(admin_sf)) {
    p <- p + geom_sf(data = admin_sf, fill = NA, color = "#333", linewidth = 0.4)
  }
  
  p + labs(title = "Data Recency by Grid Cell",
           subtitle = "Time since most recent occurrence record",
           caption = "Grey cells have no data; red cells have stale data (>10 years)") +
    theme_void() +
    theme(legend.position = "right", legend.key.height = unit(1, "cm"),
          plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40"))
}
```

## 2.4 Species Richness Map

```{r species-richness-map, fig.height=12, fig.width=10}
if (!is.null(grid_10km) && !is.null(spatial_gaps) && "n_species" %in% names(spatial_gaps)) {
  gaps_species <- spatial_gaps |>
    filter(basisofrecord == "all") |>
    select(eeacellcode, n_species)
  
  grid_species <- grid_10km |>
    left_join(gaps_species, by = "eeacellcode") |>
    mutate(
      rich_cat = case_when(
        is.na(n_species) | n_species == 0 ~ "No data",
        n_species <= 10 ~ "1–10",
        n_species <= 100 ~ "11–100",
        n_species <= 1000 ~ "101–1,000",
        TRUE ~ ">1,000"
      ),
      rich_cat = factor(rich_cat, levels = c(
        "1–10", "11–100", "101–1,000", ">1,000", "No data"
      ))
    )
  
  rich_colors <- c("1–10" = "#5c7a99", "11–100" = "#6b8f71",
                    "101–1,000" = "#c4a882", ">1,000" = "#c47a6c",
                    "No data" = "grey85")
  
  p <- ggplot(grid_species) +
    geom_sf(aes(fill = rich_cat), color = "white", linewidth = 0.02) +
    scale_fill_manual(values = rich_colors, drop = FALSE, name = "Species")
  
  if (!is.null(admin_sf)) {
    p <- p + geom_sf(data = admin_sf, fill = NA, color = "#333", linewidth = 0.4)
  }
  
  p + labs(title = "Species Richness Distribution",
           subtitle = "Number of unique species per 10km cell") +
    theme_void() +
    theme(legend.position = "right", legend.key.height = unit(1, "cm"),
          plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40"))
}
```

---

# 3. Coverage by Basis of Record

## 3.1 Zero Coverage Rates

```{r basis-overview}
spatial_overview <- safe_read(here(p_tables, "overview_spatial_by_basis.csv"))

if (!is.null(spatial_overview)) {
  spatial_10km <- spatial_overview |>
    filter(grid == "grid10km", basisofrecord != "all") |>
    arrange(pct_zero) |>
    select(basisofrecord, n_with_data, pct_coverage, n_zero, pct_zero, total_occurrences)
  
  spatial_10km |>
    kable(caption = "Coverage by Basis of Record (10km Grid)", digits = 1)
}
```

```{r basis-zero-plot, fig.height=6}
if (!is.null(spatial_overview)) {
  spatial_10km <- spatial_overview |>
    filter(grid == "grid10km", basisofrecord != "all")
  
  ggplot(spatial_10km, aes(x = fct_reorder(basisofrecord, -pct_zero), y = pct_zero, fill = pct_zero)) +
    geom_col() +
    geom_text(
      aes(label = glue("{round(pct_zero, 1)}%")),
      vjust = -0.3, size = 4
    ) +
    scale_fill_viridis_c(option = "rocket", direction = -1, begin = 0.2, end = 0.8) +
    labs(
      title = "Zero Coverage Rate by Evidence Type",
      subtitle = "Percentage of 10km cells with no occurrences",
      x = "Basis of Record",
      y = "Zero Coverage (%)",
      caption = "Lower values indicate better spatial coverage"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )
}
```

## 3.2 Coverage Comparison Across Basis Types

```{r basis-facet-maps, fig.height=14, fig.width=12}
if (!is.null(grid_10km) && !is.null(spatial_gaps) && !is.null(spatial_overview)) {
  # Get top 4 basis types by occurrence count
  top_basis <- spatial_overview |>
    filter(grid == "grid10km", basisofrecord != "all") |>
    arrange(desc(total_occurrences)) |>
    slice_head(n = 4) |>
    pull(basisofrecord)
  
  # Prepare data for faceted map
  gaps_by_basis <- spatial_gaps |>
    filter(basisofrecord %in% top_basis) |>
    select(eeacellcode, basisofrecord, occurrences)
  
  # Create faceted data by binding rows for each basis
  grid_facet_list <- lapply(top_basis, function(b) {
    basis_data <- gaps_by_basis |> filter(basisofrecord == b)
    grid_10km |>
      select(eeacellcode) |>
      left_join(basis_data, by = "eeacellcode") |>
      mutate(basisofrecord = b)
  })
  
  grid_facet <- do.call(rbind, grid_facet_list) |>
    mutate(
      log_occ = case_when(
        is.na(occurrences) ~ NA_real_,
        occurrences == 0 ~ NA_real_,
        TRUE ~ log10(occurrences + 1)
      ),
      basisofrecord = factor(basisofrecord, levels = top_basis)
    )
  
  ggplot(grid_facet) +
    geom_sf(aes(fill = log_occ), color = NA) +
    facet_wrap(~basisofrecord, ncol = 2) +
    scale_fill_viridis(
      option = "viridis",
      name = "log10(occ)",
      na.value = "grey85"
    ) +
    labs(
      title = "Coverage by Top 4 Evidence Types",
      subtitle = "Comparing geographic distribution across basis of record"
    ) +
    theme_void() +
    theme(
      legend.position = "bottom",
      strip.text = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "grey40")
    )
}
```

---

# 4. Priority Areas

## 4.1 Zero Coverage Cells

```{r priority-zero}
priority_zero <- safe_read(here(p_integrated, "priority_cells_zero_coverage.csv"))

if (!is.null(priority_zero) && nrow(priority_zero) > 0) {
  n_zero <- nrow(priority_zero)
  cat(glue("**{comma(n_zero)} cells** have zero coverage across all basis types.\n\n"))
  
  priority_zero |>
    select(any_of(c("grid", "eeacellcode", "priority_reason", "priority_level"))) |>
    slice_head(n = 20) |>
    kable(caption = "Sample of Zero Coverage Cells (Top 20)")
} else {
  cat("No zero coverage cells found or file not available.\n")
}
```

## 4.2 Priority Cells Map

```{r priority-map, fig.height=12, fig.width=10}
if (!is.null(grid_10km) && !is.null(priority_zero) && nrow(priority_zero) > 0) {
  grid_priority <- grid_10km |>
    mutate(
      is_zero = eeacellcode %in% priority_zero$eeacellcode,
      priority = if_else(is_zero, "Zero Coverage", "Has Data")
    )
  
  n_zero <- sum(grid_priority$is_zero)
  
  p <- ggplot(grid_priority) +
    geom_sf(aes(fill = priority), color = "white", linewidth = 0.02) +
    scale_fill_manual(
      values = c("Zero Coverage" = "#E63946", "Has Data" = "#2A9D8F"),
      name = "Status"
    )
  
  if (!is.null(admin_sf)) {
    p <- p + geom_sf(data = admin_sf, fill = NA, color = "#333", linewidth = 0.4)
  }
  
  p + labs(
      title = "Priority Areas: Zero Coverage Cells",
      subtitle = glue("{comma(n_zero)} cells requiring sampling attention"),
      caption = "Red cells have no occurrence records across any evidence type"
    ) +
    theme_void() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40")
    )
}
```

## 4.3 Low Coverage Analysis

```{r low-coverage-analysis, fig.height=6}
if (!is.null(cell_summary)) {
  cell_all <- cell_summary |>
    filter(basisofrecord == "all", occurrences > 0)
  
  threshold <- quantile(cell_all$occurrences, 0.1)
  q25 <- quantile(cell_all$occurrences, 0.25)
  q50 <- quantile(cell_all$occurrences, 0.50)
  
  low_coverage <- cell_all |>
    filter(occurrences <= threshold)
  
  cat("**Cells with very low coverage (bottom 10%):**\n\n")
  cat(glue("- Threshold: {comma(threshold)} occurrences\n"))
  cat(glue("- Number of cells: {comma(nrow(low_coverage))}\n\n"))
  
  level_counts <- cell_all |>
    mutate(
      coverage_level = case_when(
        occurrences <= threshold ~ "Very Low (bottom 10%)",
        occurrences <= q25 ~ "Low (10-25%)",
        occurrences <= q50 ~ "Below Median",
        TRUE ~ "Above Median"
      ),
      coverage_level = factor(coverage_level, levels = c(
        "Very Low (bottom 10%)", "Low (10-25%)", "Below Median", "Above Median"
      ))
    ) |>
    count(coverage_level)
  
  ggplot(level_counts, aes(x = coverage_level, y = n, fill = coverage_level)) +
    geom_col() +
    geom_text(aes(label = comma(n)), vjust = -0.3, size = 4) +
    scale_fill_viridis_d(option = "turbo", direction = -1) +
    labs(
      title = "Distribution of Cells by Coverage Level",
      subtitle = "10km grid, cells with data only",
      x = NULL,
      y = "Number of Cells"
    ) +
    theme(
      axis.text.x = element_text(angle = 30, hjust = 1),
      legend.position = "none"
    )
}
```

---

# 5. Last 12 Months Highlights

```{r ly-spatial-load}
cell_time <- safe_read(here(p_derived, "cell_time_summary_10km.csv"))
time_summary_ly <- safe_read(here(p_derived, "time_summary_10km.csv"))

recent_cutoff_ym <- NULL
recent_label <- NULL

if (!is.null(time_summary_ly)) {
  ts_ly <- time_summary_ly |> filter(basisofrecord == "all") |>
    mutate(yearmonth_int = as.integer(gsub("-", "", yearmonth)))

  all_ym <- sort(unique(ts_ly$yearmonth_int))
  if (length(all_ym) > 0) {
    latest_ym <- max(all_ym)
    latest_year <- as.integer(substr(as.character(latest_ym), 1, 4))
    latest_month <- as.integer(substr(as.character(latest_ym), 5, 6))
    cutoff_date <- as.Date(paste0(latest_year, "-", sprintf("%02d", latest_month), "-01")) %m-% months(11)
    recent_cutoff_ym <- as.integer(format(cutoff_date, "%Y%m"))
    recent_label <- paste0(format(cutoff_date, "%b %Y"), " – ",
      format(as.Date(paste0(latest_year, "-", sprintf("%02d", latest_month), "-01")), "%b %Y"))
  }
}
if (is.null(recent_cutoff_ym)) {
  recent_cutoff_ym <- as.integer(paste0(year(Sys.Date()) - 1, "01"))
  recent_label <- paste0(year(Sys.Date()) - 1)
}
ly_label <- recent_label
```

```{r ly-spatial-summary}
if (!is.null(cell_time)) {
  ct <- cell_time |> filter(basisofrecord == "all") |>
    mutate(yearmonth_int = as.integer(gsub("-", "", yearmonth)))

  cell_era <- ct |>
    mutate(era = ifelse(yearmonth_int >= recent_cutoff_ym, "last_year", "prior")) |>
    group_by(eeacellcode, era) |>
    summarise(occ = sum(as.numeric(occurrences), na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = era, values_from = occ, values_fill = 0)

  if (!"prior" %in% names(cell_era)) cell_era$prior <- 0
  if (!"last_year" %in% names(cell_era)) cell_era$last_year <- 0

  cell_era <- cell_era |>
    mutate(newly_covered = prior == 0 & last_year > 0,
           has_ly_data = last_year > 0)

  tibble(
    Metric = c(
      paste0("Cells active (", ly_label, ")"),
      "Newly covered cells",
      "Total recent occurrences"
    ),
    Value = c(
      comma(sum(cell_era$has_ly_data)),
      comma(sum(cell_era$newly_covered)),
      comma(sum(cell_era$last_year))
    )
  ) |> kable(caption = paste0("Spatial Activity: ", ly_label))
}
```

```{r ly-spatial-map, fig.height=12, fig.width=10}
if (!is.null(grid_10km) && exists("cell_era")) {
  grid_ly <- grid_10km |>
    left_join(cell_era |> select(eeacellcode, last_year, newly_covered, has_ly_data),
              by = "eeacellcode") |>
    mutate(status = case_when(
      newly_covered ~ "Newly covered",
      has_ly_data ~ "Active (recent)",
      TRUE ~ "No recent data"
    ))

  p <- ggplot(grid_ly) +
    geom_sf(aes(fill = status), color = "white", linewidth = 0.02) +
    scale_fill_manual(values = c("Newly covered" = "#E63946",
                                  "Active (recent)" = "#2A9D8F",
                                  "No recent data" = "grey85"))
  
  if (!is.null(admin_sf)) {
    p <- p + geom_sf(data = admin_sf, fill = NA, color = "#333", linewidth = 0.4)
  }
  
  p + labs(title = paste0("Spatial Activity: ", ly_label),
         subtitle = paste0(comma(sum(cell_era$has_ly_data)), " cells active, ",
                           comma(sum(cell_era$newly_covered)), " newly covered")) +
    theme_void() +
    theme(legend.position = "bottom",
          plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40"))
}
```

---

# 6. Key Findings

```{r summary-stats, results='asis'}
cat("## Spatial Coverage Summary\n\n")

if (!is.null(comp_grids)) {
  # Detect column names
  grid_col <- if ("grid_resolution" %in% names(comp_grids)) "grid_resolution" else "grid"
  total_col <- if ("n_cells_total" %in% names(comp_grids)) "n_cells_total" else names(comp_grids)[grepl("total", names(comp_grids))][1]
  data_col <- if ("n_cells_with_data" %in% names(comp_grids)) "n_cells_with_data" else names(comp_grids)[grepl("with_data", names(comp_grids))][1]
  
  cat("**Grid Coverage:**\n\n")
  for (i in 1:nrow(comp_grids)) {
    grid_name <- comp_grids[[grid_col]][i]
    pct_cov <- comp_grids$pct_coverage[i]
    n_data <- comp_grids[[data_col]][i]
    n_total <- comp_grids[[total_col]][i]
    cat(glue("- {grid_name}: {pct_cov}% coverage ({comma(n_data)} / {comma(n_total)} cells)\n\n"))
  }
}

cat("\n**Priority Areas:**\n\n")
if (!is.null(priority_zero) && nrow(priority_zero) > 0) {
  cat(glue("- Zero coverage cells: {comma(nrow(priority_zero))}\n\n"))
}
if (exists("low_coverage")) {
  cat(glue("- Very low coverage cells (bottom 10%): {comma(nrow(low_coverage))}\n\n"))
}

if (!is.null(spatial_overview)) {
  cat("\n**Best Coverage by Evidence Type:**\n\n")
  best_basis <- spatial_overview |>
    filter(grid == "grid10km", basisofrecord != "all") |>
    arrange(pct_zero) |>
    slice_head(n = 3)
  
  for (i in 1:nrow(best_basis)) {
    cat(glue("- {best_basis$basisofrecord[i]}: {round(100 - best_basis$pct_zero[i], 1)}% coverage\n\n"))
  }
}

cat("\n**Recommendations:**\n\n")
cat("1. Target zero-coverage cells for systematic biodiversity surveys\n")
cat("2. Prioritize mountain and northern regions if they show lower coverage\n")
cat("3. Leverage citizen science programs to expand geographic reach\n")
```

---

# Session Info

```{r session}
sessionInfo()
```
