---
title: "Spatial Gap Analysis"
subtitle: "Geographic Coverage Assessment"
author: "GBIF Gap Finder"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
  pdf_document:
    toc: true
    latex_engine: xelatex
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)

library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(sf)
library(knitr)
library(kableExtra)

data_path <- here("shiny_app", "gap_finder", "data", "shiny_data.rds")
app_data <- readRDS(data_path)
safe_get <- function(name) if (name %in% names(app_data)) app_data[[name]] else NULL

grid_10km       <- safe_get("grid_10km")
spatial_gaps    <- safe_get("spatial_gaps_10km")
cell_summary    <- safe_get("cell_summary_10km")
cell_recency    <- safe_get("cell_recency_10km")
dashboard       <- safe_get("dashboard")
admin_level1    <- safe_get("admin_level1")
kingdom_recency <- safe_get("kingdom_cell_recency")
cell_last_year  <- safe_get("cell_last_year")
last_year_label <- safe_get("recent_label") %||% "last 12 months"

# Source of truth: spatial coverage computed exactly as the app (truth_spatial),
# from the 10 km grid with basisofrecord == "all"; staleness from cell_recency.
truth_spatial <- if (!is.null(spatial_gaps)) {
  sf <- spatial_gaps |> filter(basisofrecord == "all")
  total <- nrow(sf); with_data <- sum(sf$has_data, na.rm = TRUE)
  list(total_cells = total, cells_data = with_data, cells_zero = total - with_data,
       pct_coverage = if (total > 0) round(100 * mean(sf$has_data, na.rm = TRUE), 1) else 0)
} else NULL
median_staleness <- if (!is.null(cell_recency)) {
  cr0 <- cell_recency |> filter(basisofrecord == "all")
  round(median(cr0$staleness_months, na.rm = TRUE), 1)
} else NA

pal <- list(sage = "#2A7F62", slate = "#4477AA", sand = "#CCBB44",
            coral = "#EE6677", plum = "#AA3377")

theme_gaps <- theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "#666", size = 11),
        panel.grid.minor = element_blank(), legend.position = "bottom")
theme_set(theme_gaps)
```

# Overview

This report assesses the geographic coverage of GBIF occurrence data across `r if (!is.null(grid_10km)) comma(nrow(grid_10km)) else "?"` grid cells at 10 km resolution. Spatial gaps are areas where biodiversity data is missing or insufficient, potentially due to inaccessibility, lack of survey effort, or institutional bias.

```{r spatial-stats}
if (!is.null(truth_spatial)) {
  stats <- tibble(
    Metric = c("Total grid cells", "Cells with data", "Spatial coverage",
               "Zero-coverage cells", "Median staleness"),
    Value = c(comma(truth_spatial$total_cells),
              comma(truth_spatial$cells_data),
              paste0(truth_spatial$pct_coverage, "%"),
              comma(truth_spatial$cells_zero),
              paste0(median_staleness, " months"))
  )
  kable(stats, align = c("l", "r")) |> kable_styling(full_width = FALSE)
}
```

# Occurrence Density

```{r occ-map, fig.height=9}
if (!is.null(grid_10km) && !is.null(spatial_gaps)) {
  sg <- spatial_gaps |> filter(basisofrecord == "all")
  map_sf <- grid_10km |>
    left_join(sg |> select(eeacellcode, occurrences, n_species), by = "eeacellcode") |>
    mutate(occ_cat = case_when(
      is.na(occurrences) | occurrences == 0 ~ "No data",
      occurrences <= 100 ~ "1–100",
      occurrences <= 1000 ~ "101–1,000",
      occurrences <= 10000 ~ "1,001–10,000",
      occurrences <= 100000 ~ "10,001–100,000",
      TRUE ~ "> 100,000"
    ) |> factor(levels = c("1–100", "101–1,000", "1,001–10,000",
                           "10,001–100,000", "> 100,000", "No data")))

  occ_colors <- c("1–100" = "#c6dbef", "101–1,000" = "#9ecae1",
                  "1,001–10,000" = "#6baed6", "10,001–100,000" = "#3182bd",
                  "> 100,000" = "#08519c", "No data" = "#e0e0e0")

  p <- ggplot(map_sf) +
    geom_sf(aes(fill = occ_cat), color = NA) +
    scale_fill_manual(values = occ_colors, name = "Occurrences") +
    labs(title = "Occurrence Density per Grid Cell",
         subtitle = "All basis of record types combined") +
    theme_void() + theme(legend.position = "right")
  if (!is.null(admin_level1)) {
    admin_t <- st_transform(admin_level1, st_crs(map_sf))
    p <- p + geom_sf(data = admin_t, fill = NA, color = "#888", linewidth = 0.4)
  }
  print(p)
}
```

# Species Richness

```{r richness-map, fig.height=9}
if (!is.null(grid_10km) && !is.null(spatial_gaps)) {
  sg <- spatial_gaps |> filter(basisofrecord == "all")
  map_sf <- grid_10km |>
    left_join(sg |> select(eeacellcode, n_species), by = "eeacellcode") |>
    mutate(sp_cat = case_when(
      is.na(n_species) | n_species == 0 ~ "No data",
      n_species <= 10 ~ "1–10",
      n_species <= 100 ~ "11–100",
      n_species <= 500 ~ "101–500",
      n_species <= 2000 ~ "501–2,000",
      TRUE ~ "> 2,000"
    ) |> factor(levels = c("1–10", "11–100", "101–500", "501–2,000", "> 2,000", "No data")))

  rich_colors <- c("1–10" = "#c6dbef", "11–100" = "#9ecae1",
                   "101–500" = "#6baed6", "501–2,000" = "#3182bd",
                   "> 2,000" = "#08519c", "No data" = "#e0e0e0")

  p <- ggplot(map_sf) +
    geom_sf(aes(fill = sp_cat), color = NA) +
    scale_fill_manual(values = rich_colors, name = "Species") +
    labs(title = "Species Richness per Grid Cell") +
    theme_void() + theme(legend.position = "right")
  if (!is.null(admin_level1)) {
    admin_t <- st_transform(admin_level1, st_crs(map_sf))
    p <- p + geom_sf(data = admin_t, fill = NA, color = "#888", linewidth = 0.4)
  }
  print(p)
}
```

# Coverage by Basis of Record

```{r bor-spatial, fig.height=5}
if (!is.null(spatial_gaps)) {
  bor_cov <- spatial_gaps |>
    filter(basisofrecord != "all") |>
    group_by(basisofrecord) |>
    summarise(cells_with_data = sum(has_data, na.rm = TRUE), .groups = "drop") |>
    mutate(
      total_cells = if (!is.null(grid_10km)) nrow(grid_10km) else max(cells_with_data),
      pct = round(100 * cells_with_data / total_cells, 1),
      bor_label = gsub("_", " ", basisofrecord) |> tools::toTitleCase()
    ) |> arrange(desc(pct))

  ggplot(bor_cov, aes(x = reorder(bor_label, pct), y = pct)) +
    geom_col(fill = pal$sage) +
    geom_text(aes(label = paste0(pct, "%")), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_y_continuous(limits = c(0, 105)) +
    labs(title = "Spatial Coverage by Basis of Record",
         subtitle = paste0("% of ", comma(nrow(grid_10km)), " grid cells with ≥1 occurrence"),
         x = "", y = "Coverage (%)")
}
```

# Data Recency

```{r recency-map, fig.height=9}
if (!is.null(grid_10km) && !is.null(cell_recency)) {
  cr <- cell_recency |> filter(basisofrecord == "all")
  map_sf <- grid_10km |>
    left_join(cr |> select(eeacellcode, staleness_months), by = "eeacellcode") |>
    mutate(stale_cat = case_when(
      is.na(staleness_months) ~ "No data",
      staleness_months <= 12 ~ "< 1 year",
      staleness_months <= 36 ~ "1–3 years",
      staleness_months <= 60 ~ "3–5 years",
      TRUE ~ "> 5 years"
    ) |> factor(levels = c("< 1 year", "1–3 years", "3–5 years", "> 5 years", "No data")))

  stale_colors <- c("< 1 year" = pal$sage, "1–3 years" = pal$slate,
                    "3–5 years" = pal$sand, "> 5 years" = pal$coral,
                    "No data" = "#e0e0e0")

  p <- ggplot(map_sf) +
    geom_sf(aes(fill = stale_cat), color = NA) +
    scale_fill_manual(values = stale_colors, name = "Last record") +
    labs(title = "Data Recency per Grid Cell",
         subtitle = "Time since the most recent occurrence record") +
    theme_void() + theme(legend.position = "right")
  if (!is.null(admin_level1)) {
    admin_t <- st_transform(admin_level1, st_crs(map_sf))
    p <- p + geom_sf(data = admin_t, fill = NA, color = "#888", linewidth = 0.4)
  }
  print(p)
}
```

# Coverage by Kingdom

```{r kingdom-recency, fig.height=8}
if (!is.null(kingdom_recency)) {
  king_cov <- kingdom_recency |>
    group_by(kingdom) |>
    summarise(
      n_cells = n(),
      total_occ = sum(total_occ, na.rm = TRUE),
      .groups = "drop"
    ) |> arrange(desc(total_occ))

  ggplot(king_cov, aes(x = reorder(kingdom, n_cells), y = n_cells)) +
    geom_col(fill = pal$sage) +
    geom_text(aes(label = comma(n_cells)), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(title = "Grid Cells with Data by Kingdom",
         x = "", y = "Number of cells")
}
```

# Last 12 Months — Spatial Activity

```{r last-year-map, fig.height=9}
if (!is.null(grid_10km) && !is.null(cell_last_year)) {
  map_sf <- grid_10km |>
    left_join(cell_last_year |> select(eeacellcode, occ_last_year, occ_prior, newly_covered),
              by = "eeacellcode") |>
    mutate(activity = case_when(
      !is.na(newly_covered) & newly_covered ~ "Newly covered",
      !is.na(occ_last_year) & occ_last_year > 0 ~ "Active",
      !is.na(occ_prior) & occ_prior > 0 ~ "Prior only",
      TRUE ~ "No data"
    ) |> factor(levels = c("Newly covered", "Active", "Prior only", "No data")))

  act_colors <- c("Newly covered" = pal$coral, "Active" = pal$sage,
                  "Prior only" = pal$sand, "No data" = "#e0e0e0")

  p <- ggplot(map_sf) +
    geom_sf(aes(fill = activity), color = NA) +
    scale_fill_manual(values = act_colors, name = "Activity") +
    labs(title = paste0("Spatial Activity: ", last_year_label),
         subtitle = "Cells with observations dated in the last 12 months") +
    theme_void() + theme(legend.position = "right")
  if (!is.null(admin_level1)) {
    admin_t <- st_transform(admin_level1, st_crs(map_sf))
    p <- p + geom_sf(data = admin_t, fill = NA, color = "#888", linewidth = 0.4)
  }
  print(p)
}
```

---

*Report generated by `gbif_gap_finder` on `r Sys.time()`.*
