---
title: "Temporal Gap Analysis"
subtitle: "GBIF Gap Analysis - Temporal Patterns & Data Recency"
author: "gbifgaps"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    fig_width: 12
    fig_height: 8
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 3
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 300
)

# Hide code in PDF output, show (with folding) in HTML
if (!knitr::is_html_output()) {
  knitr::opts_chunk$set(echo = FALSE)
}

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(viridis)
library(patchwork)
library(sf)
library(knitr)
library(data.table)
library(glue)
library(lubridate)

# Theme
theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12, color = "grey40"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
)

# Paths
p_gaps <- here("data_proc", "gaps")
p_tables <- here("output", "tables")
p_data_proc <- here("data_proc")
p_derived <- here("data_proc", "derived")
p_integrated <- here("output", "tables", "integrated")

# Current year for calculations
current_year <- year(Sys.Date())
```

# Overview

This report analyzes **temporal gaps** in the national biodiversity occurrence data, examining historical trends, seasonal patterns, data recency, and how data mobilization has changed over time for different taxonomic groups.

**Key questions addressed:**

- What are the long-term trends in biodiversity data collection?
- Are there seasonal biases in sampling?
- How recent is the data in different geographic areas?
- How has data mobilization changed for different taxonomic orders over 5-year periods?

---

# 1. Historical Trends

## 1.1 Annual Occurrence Trends

```{r load-temporal}
# Load time summaries
time10 <- read_csv(here(p_derived, "time_summary_10km.csv"), show_col_types = FALSE) |>
  filter(basisofrecord == "all") |>
  mutate(
    yearmonth_chr = str_trim(as.character(yearmonth)),
    year = as.integer(str_sub(yearmonth_chr, 1, 4)),
    month = as.integer(str_sub(yearmonth_chr, 6, 7))
  )

# Annual summary
annual <- time10 |>
  group_by(year) |>
  summarise(
    occurrences = sum(occurrences, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(year >= 1900, year <= current_year)
```

```{r annual-trend, fig.height=6}
ggplot(annual, aes(x = year, y = occurrences)) +
  geom_area(fill = viridis(1, option = "viridis", begin = 0.4), alpha = 0.3) +
  geom_line(color = viridis(1, option = "viridis", begin = 0.4), linewidth = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Historical Trend in Occurrence Records",
    subtitle = "Annual totals across all evidence types",
    x = "Year",
    y = "Total Occurrences"
  )
```

## 1.2 Recent Decades (1970-Present)

```{r recent-trend, fig.height=6}
annual |>
  filter(year >= 1970) |>
  ggplot(aes(x = year, y = occurrences)) +
  geom_col(fill = viridis(1, option = "mako", begin = 0.3), alpha = 0.8) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Recent Occurrence Trends (1970-Present)",
    subtitle = "Annual occurrence records",
    x = "Year",
    y = "Total Occurrences"
  )
```

## 1.3 Year-over-Year Change

```{r yoy-change, fig.height=6}
annual |>
  filter(year >= 1990) |>
  mutate(
    prev_year = lag(occurrences),
    pct_change = 100 * (occurrences - prev_year) / prev_year,
    direction = ifelse(pct_change >= 0, "Increase", "Decrease")
  ) |>
  filter(!is.na(pct_change)) |>
  ggplot(aes(x = year, y = pct_change, fill = direction)) +
  geom_col() +
  scale_fill_viridis_d(option = "turbo", begin = 0.2, end = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title = "Year-over-Year Change in Data Collection",
    subtitle = "Percentage change from previous year (1990-present)",
    x = "Year",
    y = "% Change",
    fill = NULL
  )
```

## 1.4 Five-Year Period Change

```{r 5yr-change, fig.height=6}
# Aggregate by 5-year periods
annual_5yr <- annual |>
  filter(year >= 1970) |>
  mutate(period = floor(year / 5) * 5) |>
  group_by(period) |>
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
  mutate(
    prev_period = lag(occurrences),
    pct_change = 100 * (occurrences - prev_period) / prev_period,
    direction = ifelse(pct_change >= 0, "Increase", "Decrease"),
    period_label = paste0(period, "-", period + 4)
  ) |>
  filter(!is.na(pct_change))

ggplot(annual_5yr, aes(x = period_label, y = pct_change, fill = direction)) +
  geom_col() +
  geom_text(aes(label = paste0(ifelse(pct_change >= 0, "+", ""), round(pct_change, 0), "%")),
            vjust = ifelse(annual_5yr$pct_change >= 0, -0.3, 1.3), size = 3.5) +
  scale_fill_viridis_d(option = "turbo", begin = 0.2, end = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title = "Five-Year Period Change in Data Collection",
    subtitle = "Percentage change from previous 5-year period (1970-present)",
    x = "Period",
    y = "% Change",
    fill = NULL
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 2. Seasonal Patterns

## 2.1 Monthly Distribution

```{r monthly-pattern, fig.height=5}
monthly <- time10 |>
  group_by(month) |>
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")

ggplot(monthly, aes(x = factor(month), y = occurrences, fill = occurrences)) +
  geom_col() +
  scale_fill_viridis_c(option = "plasma", labels = comma) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Seasonal Sampling Pattern",
    subtitle = "Total occurrences by month (all years combined)",
    x = "Month",
    y = "Total Occurrences",
    fill = "Occurrences"
  ) +
  theme(legend.position = "none")
```

## 2.2 Year × Month Heatmap

```{r heatmap, fig.height=10}
time10 |>
  filter(year >= 1950, year <= current_year) |>
  group_by(year, month) |>
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = factor(month), y = year, fill = log10(occurrences + 1))) +
  geom_tile() +
  scale_fill_viridis(option = "inferno", name = "log10(occ+1)") +
  scale_x_discrete(labels = month.abb) +
  labs(
    title = "Sampling Intensity Over Time",
    subtitle = "Year × Month heatmap (1950-present)",
    x = "Month",
    y = "Year"
  )
```

## 2.3 Seasonal Bias by Decade

```{r seasonal-decade, fig.height=7}
time10 |>
  filter(year >= 1970) |>
  mutate(decade = paste0(floor(year / 10) * 10, "s")) |>
  group_by(decade, month) |>
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
  group_by(decade) |>
  mutate(pct = 100 * occurrences / sum(occurrences)) |>
  ungroup() |>
  ggplot(aes(x = factor(month), y = pct, fill = decade)) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d(option = "viridis") +
  scale_x_discrete(labels = month.abb) +
  labs(
    title = "Seasonal Distribution by Decade",
    subtitle = "Has seasonal sampling bias changed over time?",
    x = "Month",
    y = "% of Decade's Total",
    fill = "Decade"
  )
```

---

# 3. Data Recency

## 3.1 Staleness by Basis of Record

```{r recency-overview}
if (file.exists(here(p_tables, "overview_temporal_recency_rates.csv"))) {
  recency <- read_csv(here(p_tables, "overview_temporal_recency_rates.csv"), show_col_types = FALSE)
  
  recency_10km <- recency |>
    filter(grid == "grid10km") |>
    arrange(pct_stale_5y)
  
  # Select only columns that exist
  display_cols <- intersect(
    c("basisofrecord", "median_staleness_months", "pct_stale_5y"),
    names(recency_10km)
  )
  
  recency_10km |>
    select(all_of(display_cols)) |>
    kable(caption = "Data Staleness by Evidence Type (10km Grid)", digits = 1)
}
```

```{r recency-plot, fig.height=6}
if (exists("recency_10km")) {
  recency_10km |>
    filter(basisofrecord != "all") |>
    ggplot(aes(x = fct_reorder(basisofrecord, pct_stale_5y), 
               y = pct_stale_5y, fill = pct_stale_5y)) +
    geom_col() +
    geom_text(aes(label = paste0(round(pct_stale_5y, 1), "%")), 
              hjust = -0.1, size = 4) +
    coord_flip() +
    scale_fill_viridis_c(option = "rocket", direction = -1, begin = 0.2, end = 0.9) +
    labs(
      title = "Data Staleness by Evidence Type",
      subtitle = "Percentage of cells not sampled in 5+ years",
      x = NULL,
      y = "Stale Cells (%)",
      caption = "Lower values indicate more recent data"
    ) +
    theme(legend.position = "none")
}
```

## 3.2 Geographic Distribution of Staleness

```{r stale-map, fig.height=12, fig.width=10}
# Try multiple possible file names
recency_file <- NULL
possible_files <- c(
  here(p_gaps, "temporal_cell_recency_10km.csv"),
  here(p_gaps, "cell_recency_10km.csv"),
  here(p_derived, "cell_recency_10km.csv")
)

for (f in possible_files) {
  if (file.exists(f)) {
    recency_file <- f
    break
  }
}

if (!is.null(recency_file)) {
  cell_recency <- fread(recency_file)
  
  # Load grid
  grid_10km <- st_read(here(p_data_proc, "grids_10km.gpkg"), quiet = TRUE)
  
  # Find cellcode field
  cellcode_candidates <- names(grid_10km)[grepl("cellcode|eeacell", tolower(names(grid_10km)))]
  cellcode_field <- if(length(cellcode_candidates) > 0) cellcode_candidates[1] else names(grid_10km)[grepl("code", tolower(names(grid_10km)))][1]
  
  grid_10km <- grid_10km |>
    mutate(eeacellcode = as.character(.data[[cellcode_field]]))
  
  # Filter for "all" basis and join
  recency_all <- cell_recency |>
    filter(basisofrecord == "all") |>
    select(eeacellcode, staleness_months)
  
  grid_stale <- grid_10km |>
    left_join(recency_all, by = "eeacellcode") |>
    mutate(
      staleness_years = staleness_months / 12,
      staleness_cat = case_when(
        is.na(staleness_years) ~ "No Data",
        staleness_years <= 1 ~ "< 1 year",
        staleness_years <= 3 ~ "1-3 years",
        staleness_years <= 5 ~ "3-5 years",
        staleness_years <= 10 ~ "5-10 years",
        TRUE ~ "> 10 years"
      ),
      staleness_cat = factor(staleness_cat, levels = c(
        "< 1 year", "1-3 years", "3-5 years", "5-10 years", "> 10 years", "No Data"
      ))
    )
  
  ggplot(grid_stale) +
    geom_sf(aes(fill = staleness_cat), color = "white", linewidth = 0.02) +
    scale_fill_viridis_d(option = "turbo", direction = -1, na.value = "grey80") +
    labs(
      title = "Geographic Distribution of Data Recency",
      subtitle = "Time since last occurrence record per cell",
      fill = "Time Since\nLast Record"
    ) +
    theme_void() +
    theme(
      legend.position = "right",
      legend.key.height = unit(0.8, "cm"),
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40")
    )
} else {
  cat("Cell recency file not found. Checked:\n")
  cat(paste("-", possible_files, collapse = "\n"))
}
```

---

# 4. Data Mobilization by Taxonomic Order (5-Year Periods)

This section examines how data mobilization has changed over time for different taxonomic orders, using 5-year time bins.

```{r load-order-data}
# Check for order-level time summary
order_path <- here(p_derived, "order_time_summary_10km.csv")

if (file.exists(order_path)) {
  order_time <- fread(order_path) |>
    filter(basisofrecord == "all") |>
    mutate(
      yearmonth_chr = str_trim(as.character(yearmonth)),
      year = as.integer(str_sub(yearmonth_chr, 1, 4))
    ) |>
    filter(year >= 1970, year <= current_year, !is.na(order), order != "")
  
  # Create 5-year bins
  order_5yr <- order_time |>
    mutate(
      period_start = floor(year / 5) * 5,
      period = paste0(period_start, "-", period_start + 4)
    ) |>
    group_by(order, period, period_start) |>
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")
  
  # Get top 12 orders by total occurrences
  top_orders <- order_time |>
    group_by(order) |>
    summarise(total = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(total)) |>
    slice_head(n = 12) |>
    pull(order)
  
  order_5yr_top <- order_5yr |>
    filter(order %in% top_orders)
  
  has_order_data <- TRUE
} else {
  cat("Order-level time summary not found. Please ensure script 06 generates order_time_summary_10km.csv\n")
  has_order_data <- FALSE
}
```

## 4.1 Data Mobilization Trends by Order (Top 12)

```{r order-5yr-histogram, fig.height=10, fig.width=14}
if (has_order_data) {
  ggplot(order_5yr_top, aes(x = period, y = occurrences, fill = order)) +
    geom_col() +
    facet_wrap(~order, scales = "free_y", ncol = 3) +
    scale_fill_viridis_d(option = "turbo") +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Data Mobilization by Taxonomic Order (5-Year Periods)",
      subtitle = "Top 12 orders by total occurrence count",
      x = "5-Year Period",
      y = "Occurrences",
      caption = "Each panel shows a different taxonomic order"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      legend.position = "none",
      strip.text = element_text(face = "bold", size = 10)
    )
}
```

## 4.2 Data Mobilization Trends by Order (Bottom 12)

```{r order-5yr-bottom, fig.height=10, fig.width=14}
if (has_order_data) {
  # Get bottom 12 orders by total occurrences (excluding very rare)
  bottom_orders <- order_time |>
    group_by(order) |>
    summarise(total = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
    filter(total >= 100) |>  # Exclude extremely rare orders
    arrange(total) |>
    slice_head(n = 12) |>
    pull(order)
  
  order_5yr_bottom <- order_5yr |>
    filter(order %in% bottom_orders)
  
  ggplot(order_5yr_bottom, aes(x = period, y = occurrences, fill = order)) +
    geom_col() +
    facet_wrap(~order, scales = "free_y", ncol = 3) +
    scale_fill_viridis_d(option = "rocket") +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Data Mobilization by Taxonomic Order (5-Year Periods)",
      subtitle = "Bottom 12 orders by total occurrence count (min 100 records)",
      x = "5-Year Period",
      y = "Occurrences",
      caption = "Each panel shows a different taxonomic order"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      legend.position = "none",
      strip.text = element_text(face = "bold", size = 10)
    )
}
```

## 4.3 Comparative Mobilization Trends

```{r order-comparison, fig.height=8}
if (has_order_data) {
  # Normalize to show relative change
  order_5yr_norm <- order_5yr_top |>
    group_by(order) |>
    mutate(
      max_occ = max(occurrences),
      pct_of_max = 100 * occurrences / max_occ
    ) |>
    ungroup()
  
  ggplot(order_5yr_norm, aes(x = period_start, y = pct_of_max, color = order, group = order)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2) +
    scale_color_viridis_d(option = "turbo") +
    labs(
      title = "Relative Data Mobilization Trends by Order",
      subtitle = "Normalized to each order's peak period (= 100%)",
      x = "Period Start Year",
      y = "% of Peak",
      color = "Order"
    ) +
    theme(legend.position = "right")
}
```

## 4.4 Change in Mobilization: Recent vs Historical

```{r order-change, fig.height=7}
if (has_order_data) {
  # Compare last 10 years vs previous 10 years
  recent_cutoff <- current_year - 10
  historical_cutoff <- current_year - 20
  
  order_change <- order_time |>
    filter(order %in% top_orders) |>
    mutate(
      era = case_when(
        year >= recent_cutoff ~ "Recent (last 10 years)",
        year >= historical_cutoff ~ "Historical (10-20 years ago)",
        TRUE ~ NA_character_
      )
    ) |>
    filter(!is.na(era)) |>
    group_by(order, era) |>
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = era, values_from = occurrences) |>
    mutate(
      pct_change = 100 * (`Recent (last 10 years)` - `Historical (10-20 years ago)`) / 
                   `Historical (10-20 years ago)`,
      direction = ifelse(pct_change >= 0, "Increased", "Decreased")
    ) |>
    arrange(desc(pct_change))
  
  ggplot(order_change, aes(x = fct_reorder(order, pct_change), y = pct_change, fill = direction)) +
    geom_col() +
    geom_text(aes(label = paste0(round(pct_change, 0), "%")),
              hjust = ifelse(order_change$pct_change >= 0, -0.1, 1.1), size = 3.5) +
    coord_flip() +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.7) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    labs(
      title = "Change in Data Mobilization by Order",
      subtitle = paste0("Comparing ", recent_cutoff, "-", current_year, 
                       " vs ", historical_cutoff, "-", recent_cutoff - 1),
      x = NULL,
      y = "% Change in Occurrences",
      fill = NULL
    ) +
    theme(legend.position = "bottom")
}
```

## 4.5 5-Year Mobilization Summary Table

```{r order-table}
if (has_order_data) {
  # Create summary table
  order_summary <- order_5yr_top |>
    pivot_wider(
      id_cols = order,
      names_from = period,
      values_from = occurrences,
      values_fill = 0
    )
  
  # Add total column
  order_summary <- order_summary |>
    mutate(Total = rowSums(across(where(is.numeric)))) |>
    arrange(desc(Total))
  
  order_summary |>
    mutate(across(where(is.numeric), comma)) |>
    kable(caption = "Occurrences by Order and 5-Year Period")
}
```

---

# 5. Stale Cells Analysis

## 5.1 Distribution of Staleness

```{r stale-distribution, fig.height=5}
if (file.exists(here(p_integrated, "priority_stale_cells.csv"))) {
  priority_stale <- fread(here(p_integrated, "priority_stale_cells.csv"))
  
  cat("**", comma(nrow(priority_stale)), "cells** have not been sampled in 5+ years.\n\n")
  
  ggplot(priority_stale, aes(x = staleness_months / 12)) +
    geom_histogram(bins = 40, fill = viridis(1, option = "rocket", begin = 0.3), 
                   color = "white", alpha = 0.8) +
    labs(
      title = "Distribution of Data Staleness",
      subtitle = paste0("Median: ", round(median(priority_stale$staleness_months) / 12, 1), " years"),
      x = "Years Since Last Sample",
      y = "Number of Cells"
    )
} else {
  cat("Priority stale cells file not found.\n")
}
```

## 5.2 Most Stale Cells

```{r most-stale-map, fig.height=12, fig.width=10}
if (exists("priority_stale") && exists("grid_10km")) {
  # Join staleness data to grid
  stale_data <- priority_stale |>
    select(eeacellcode, staleness_months) |>
    mutate(staleness_years = staleness_months / 12)
  
  grid_stale_priority <- grid_10km |>
    left_join(stale_data, by = "eeacellcode") |>
    mutate(
      staleness_cat = case_when(
        is.na(staleness_years) ~ "Recent (<5 years)",
        staleness_years <= 7 ~ "5-7 years",
        staleness_years <= 10 ~ "7-10 years",
        staleness_years <= 15 ~ "10-15 years",
        staleness_years <= 20 ~ "15-20 years",
        TRUE ~ ">20 years"
      ),
      staleness_cat = factor(staleness_cat, levels = c(
        "Recent (<5 years)", "5-7 years", "7-10 years", 
        "10-15 years", "15-20 years", ">20 years"
      ))
    )
  
  ggplot(grid_stale_priority) +
    geom_sf(aes(fill = staleness_cat), color = "white", linewidth = 0.02) +
    scale_fill_viridis_d(
      option = "rocket", 
      direction = -1,
      na.value = "#2A9D8F",
      drop = FALSE
    ) +
    labs(
      title = "Priority Stale Cells by Staleness",
      subtitle = glue("{comma(nrow(priority_stale))} cells not sampled in 5+ years"),
      fill = "Years Since\nLast Sample",
      caption = "Darker colors indicate longer time since last observation"
    ) +
    theme_void() +
    theme(
      legend.position = "right",
      legend.key.height = unit(0.8, "cm"),
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40")
    )
}
```

```{r most-stale-table}
if (exists("priority_stale")) {
  priority_stale |>
    arrange(desc(staleness_months)) |>
    mutate(years_stale = round(staleness_months / 12, 1)) |>
    select(grid, eeacellcode, last_ym, years_stale, total_occurrences) |>
    slice_head(n = 20) |>
    kable(caption = "Top 20 Most Stale Cells")
}
```

---

# 6. Key Findings

```{r summary, results='asis'}
cat("## Temporal Gap Summary\n\n")

# Year range
if (exists("annual") && nrow(annual) > 0) {
  cat("**Time Range:**\n\n")
  cat("- Data spans:", min(annual$year), "-", max(annual$year), "\n")
  cat("- Peak year:", annual$year[which.max(annual$occurrences)], 
      "(", comma(max(annual$occurrences)), " occurrences)\n\n")
}

# Seasonal patterns
if (exists("monthly") && nrow(monthly) > 0) {
  peak_month <- monthly$month[which.max(monthly$occurrences)]
  cat("**Seasonal Patterns:**\n\n")
  cat("- Peak month:", month.name[peak_month], "\n")
  cat("- Summer (Jun-Aug) accounts for:", 
      round(100 * sum(monthly$occurrences[monthly$month %in% 6:8]) / sum(monthly$occurrences), 1),
      "% of records\n\n")
}

# Recency
if (exists("recency_10km") && nrow(recency_10km) > 0) {
  all_recency <- recency_10km |> filter(basisofrecord == "all")
  if (nrow(all_recency) > 0) {
    cat("**Data Recency:**\n\n")
    cat("- Median staleness:", round(all_recency$median_staleness_months[1], 1), "months\n")
    cat("- Cells stale >5 years:", round(all_recency$pct_stale_5y[1], 1), "%\n\n")
  }
}

# Order mobilization
if (exists("has_order_data") && has_order_data && exists("order_change")) {
  cat("**Mobilization Trends by Order:**\n\n")
  increasing <- order_change |> filter(pct_change > 0) |> nrow()
  decreasing <- order_change |> filter(pct_change < 0) |> nrow()
  cat("- Orders with increased mobilization:", increasing, "\n")
  cat("- Orders with decreased mobilization:", decreasing, "\n")
  top_increase <- order_change |> slice_head(n = 1)
  if (nrow(top_increase) > 0) {
    cat("- Largest increase:", top_increase$order[1], "(+", round(top_increase$pct_change[1], 0), "%)\n\n")
  }
}

cat("**Recommendations:**\n\n")
cat("1. Re-survey cells with >5 years since last observation\n")
cat("2. Address seasonal sampling bias with winter surveys\n")
cat("3. Maintain momentum for orders showing increased mobilization\n")
cat("4. Investigate causes for orders showing decreased mobilization\n")
```

---

# Session Info

```{r session}
sessionInfo()
```
