---
title: "Temporal Gap Analysis"
subtitle: "Temporal Patterns, Trends, and Data Recency"
author: "GBIF Gap Finder"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
  pdf_document:
    toc: true
    latex_engine: xelatex
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6, dpi = 150
)

library(here); library(dplyr); library(tidyr); library(ggplot2)
library(scales); library(knitr); library(kableExtra)

data_path <- here("shiny_app", "gap_finder", "data", "shiny_data.rds")
app_data <- readRDS(data_path)
safe_get <- function(name) if (name %in% names(app_data)) app_data[[name]] else NULL

time_summary   <- safe_get("time_summary_10km")
cell_recency   <- safe_get("cell_recency_10km")
dashboard      <- safe_get("dashboard")
overview_ly    <- safe_get("overview_last_year")
last_year_label <- safe_get("recent_label") %||% "last 12 months"

pal <- list(sage = "#6b8f71", slate = "#5c7a99", sand = "#c4a882",
            coral = "#c47a6c", plum = "#8b6d8f")
theme_gaps <- theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "#666", size = 11),
        panel.grid.minor = element_blank(), legend.position = "bottom")
theme_set(theme_gaps)
```

# Overview

This report examines the temporal distribution of GBIF occurrence data — when records were observed, how data collection has evolved over decades, and which areas have the most recent or most outdated records.

# Occurrence Volume Over Time

```{r annual-trend, fig.height=5}
if (!is.null(time_summary)) {
  df <- time_summary |>
    filter(basisofrecord == "all", !is.na(year), year >= 1900) |>
    group_by(year) |>
    summarise(occurrences = sum(as.numeric(occurrences), na.rm = TRUE), .groups = "drop")

  ggplot(df, aes(x = year, y = occurrences)) +
    geom_area(fill = pal$sage, alpha = 0.7) +
    geom_line(color = pal$sage, linewidth = 0.5) +
    scale_y_continuous(labels = label_comma()) +
    labs(title = "Annual Occurrence Volume", subtitle = "All record types, 1900–present",
         x = "Year", y = "Number of occurrences")
}
```

# By Basis of Record

```{r bor-temporal, fig.height=6}
if (!is.null(time_summary)) {
  df <- time_summary |>
    filter(basisofrecord != "all", !is.na(year), year >= 1970) |>
    mutate(bor_label = gsub("_", " ", basisofrecord) |> tools::toTitleCase()) |>
    group_by(year, bor_label) |>
    summarise(occurrences = sum(as.numeric(occurrences), na.rm = TRUE), .groups = "drop")

  # Keep top 4 BOR types
  top_bor <- df |> group_by(bor_label) |>
    summarise(total = sum(occurrences)) |> arrange(desc(total)) |> head(4)
  df <- df |> filter(bor_label %in% top_bor$bor_label)

  bor_colors <- c("Human Observation" = pal$sage, "Preserved Specimen" = pal$slate,
                  "Material Sample" = pal$sand, "Machine Observation" = pal$coral)

  ggplot(df, aes(x = year, y = occurrences, fill = bor_label)) +
    geom_area(alpha = 0.7, position = "stack") +
    scale_fill_manual(values = bor_colors) +
    scale_y_continuous(labels = label_comma()) +
    labs(title = "Occurrence Volume by Record Type",
         subtitle = "Top 4 basis of record types, 1970–present",
         x = "Year", y = "Occurrences", fill = "")
}
```

# Decade Summary

```{r decade, fig.height=5}
if (!is.null(time_summary)) {
  df <- time_summary |>
    filter(basisofrecord == "all", !is.na(year), year >= 1900) |>
    mutate(decade = floor(year / 10) * 10) |>
    group_by(decade) |>
    summarise(occ = sum(as.numeric(occurrences), na.rm = TRUE),
              spp = sum(as.numeric(n_species), na.rm = TRUE), .groups = "drop")

  ggplot(df, aes(x = decade, y = occ)) +
    geom_col(fill = pal$slate) +
    scale_y_continuous(labels = label_comma()) +
    scale_x_continuous(breaks = seq(1900, 2020, 20)) +
    labs(title = "Occurrences per Decade", x = "Decade", y = "Total occurrences")
}
```

# Data Recency Distribution

```{r recency-hist, fig.height=5}
if (!is.null(cell_recency)) {
  cr <- cell_recency |> filter(basisofrecord == "all", !is.na(staleness_months))

  ggplot(cr, aes(x = staleness_months / 12)) +
    geom_histogram(binwidth = 1, fill = pal$sage, color = "white") +
    geom_vline(xintercept = 5, linetype = "dashed", color = pal$coral) +
    annotate("text", x = 5.5, y = Inf, label = "5-year threshold", vjust = 2,
             color = pal$coral, size = 3.5) +
    labs(title = "Distribution of Cell Staleness",
         subtitle = "Years since the most recent record per grid cell",
         x = "Years since last record", y = "Number of cells")
}
```

# Staleness Summary

```{r staleness-table}
if (!is.null(cell_recency)) {
  cr <- cell_recency |> filter(basisofrecord == "all")
  stale_df <- tibble(
    Category = c("< 1 year", "1–3 years", "3–5 years", "> 5 years", "No data"),
    Cells = c(
      sum(cr$staleness_months <= 12, na.rm = TRUE),
      sum(cr$staleness_months > 12 & cr$staleness_months <= 36, na.rm = TRUE),
      sum(cr$staleness_months > 36 & cr$staleness_months <= 60, na.rm = TRUE),
      sum(cr$staleness_months > 60, na.rm = TRUE),
      sum(is.na(cr$staleness_months))
    )
  ) |> mutate(`%` = round(100 * Cells / sum(Cells), 1))

  kable(stale_df, align = c("l", "r", "r")) |> kable_styling(full_width = FALSE)
}
```

# The Last 12 Months

```{r last-12}
if (!is.null(overview_ly)) {
  ly <- tibble(
    Metric = c("Occurrences dated", "Cells active", "Newly covered cells"),
    Value = c(comma(overview_ly$occ_last_year),
              comma(overview_ly$cells_active_last_year),
              comma(overview_ly$cells_newly_covered))
  )
  kable(ly, caption = last_year_label, align = c("l", "r")) |>
    kable_styling(full_width = FALSE)
}
```

---

*Report generated by `gbif_gap_finder` on `r Sys.time()`.*
