---
title: "Taxonomic Gap Analysis"
subtitle: "GBIF Gap Analysis - Species Coverage & Conservation Priorities"
author: "gbifgaps"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    fig_width: 12
    fig_height: 8
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 3
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 150
)

# Hide code in PDF output, show (with folding) in HTML
if (!knitr::is_html_output()) {
  knitr::opts_chunk$set(echo = FALSE)
}

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(viridis)
library(patchwork)
library(sf)
library(knitr)
library(data.table)
library(glue)
library(lubridate)

theme_gaps <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "grey40"),
    plot.caption = element_text(size = 9, color = "grey50"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 11, face = "bold")
  )

theme_set(theme_gaps)

p_gaps <- here("data_proc", "gaps")
p_tables <- here("output", "tables")
p_data_proc <- here("data_proc")
p_derived <- here("data_proc", "derived")
p_integrated <- here("output", "tables", "integrated")

current_year <- year(Sys.Date())

threat_levels <- c("CR", "EN", "VU", "NT", "LC", "DD", "NE")
threat_colors <- c(
  "CR" = "#D32F2F", "EN" = "#E64A19", "VU" = "#F57C00",
  "NT" = "#FBC02D", "LC" = "#8BC34A", "DD" = "#78909C", "NE" = "#B0BEC5"
)

safe_read <- function(path) {
  if (!file.exists(path)) return(NULL)
  fread(path)
}
```

# Overview

This report analyzes **taxonomic gaps** in the national biodiversity data, comparing coverage against the national backbone and examining threatened species representation.

**Key Questions:**

- What proportion of the country's known species are in GBIF?
- Which taxonomic groups have the best/worst coverage?
- How well are threatened species represented?
- Which species should be prioritized for data collection?

---

# 1. Overall Taxonomic Coverage

## 1.1 Summary Statistics

```{r load-data}
dashboard <- safe_read(here(p_tables, "dashboard_summary.csv"))
tax_ref <- safe_read(here(p_data_proc, "taxonomy_backbone.csv"))
tax_match <- safe_read(here(p_gaps, "taxonomic_match_summary.csv"))

if (!is.null(dashboard)) {
  cat(glue("- **Taxa in backbone reference**: {comma(dashboard$taxa_in_reference)}\n\n"))
  cat(glue("- **Taxa found in GBIF**: {comma(dashboard$taxa_in_gbif)}\n\n"))
  cat(glue("- **Overall coverage**: {dashboard$taxa_pct_coverage}%\n\n"))
  cat(glue("- **Priority taxa identified**: {comma(dashboard$n_priority_taxa)}\n\n"))
}
```

## 1.2 Coverage by Rank

```{r coverage-by-rank, fig.height=8}
tax_by_rank <- safe_read(here(p_tables, "overview_taxonomic_by_rank.csv"))

if (!is.null(tax_by_rank) && nrow(tax_by_rank) > 0) {
  rank_plot <- tax_by_rank |>
    filter(n_ref_total >= 10) |>
    arrange(desc(n_ref_total)) |>
    slice_head(n = 15)
  
  if (nrow(rank_plot) > 0) {
    ggplot(rank_plot, aes(x = fct_reorder(taxonRank, pct_coverage), 
                           y = pct_coverage, fill = pct_coverage)) +
      geom_col(alpha = 0.9) +
      geom_text(aes(label = paste0(round(pct_coverage, 1), "%")), hjust = -0.1, size = 4) +
      coord_flip() +
      scale_fill_viridis_c(option = "viridis", begin = 0.2, end = 0.9) +
      scale_y_continuous(limits = c(0, 110), expand = expansion(mult = c(0, 0))) +
      labs(title = "GBIF Coverage by Taxonomic Rank", x = NULL, y = "Coverage (%)") +
      theme(legend.position = "none")
  }
}
```

## 1.3 Orders by Kingdom: Backbone vs GBIF Coverage

This section shows all orders in the backbone grouped by class, where:

- **Bar length** = total number of species in the backbone for that order
- **Green portion** = species found in GBIF
- **Blue portion** = species missing from GBIF

```{r order-coverage-data}
# Load the match summary which has species-level matching info
match_summary <- safe_read(here(p_gaps, "taxonomic_match_summary.csv"))

order_coverage <- NULL

if (!is.null(match_summary) && nrow(match_summary) > 0 && "order" %in% names(match_summary)) {
  # Summarize by order: count species in backbone and in GBIF

  order_coverage <- match_summary |>
    filter(!is.na(order), order != "") |>
    group_by(across(any_of(c("kingdom", "phylum", "class", "order")))) |>
    summarise(
      n_species_backbone = n(),
      n_species_gbif = sum(matched_any, na.rm = TRUE),
      n_species_missing = n_species_backbone - n_species_gbif,
      pct_coverage = round(100 * n_species_gbif / n_species_backbone, 1),
      .groups = "drop"
    )
  
  cat(glue("**Total orders in backbone**: {nrow(order_coverage)}\n\n"))
  cat(glue("**Total species in backbone**: {comma(sum(order_coverage$n_species_backbone))}\n\n"))
  cat(glue("**Species found in GBIF**: {comma(sum(order_coverage$n_species_gbif))} ({round(100 * sum(order_coverage$n_species_gbif) / sum(order_coverage$n_species_backbone), 1)}%)\n\n"))
}
```

```{r order-plot-function}
# Function to create the stacked bar plot for a kingdom (absolute numbers)
plot_kingdom_orders <- function(data, kingdom_name, n_top_classes = 12) {
  if (is.null(data) || nrow(data) == 0) {
    return(NULL)
  }
  
  # Reshape to long format for stacking
  plot_data <- data |>
    select(class, order, n_species_gbif, n_species_missing) |>
    pivot_longer(
      cols = c(n_species_gbif, n_species_missing),
      names_to = "status",
      values_to = "n_species"
    ) |>
    mutate(
      status = factor(
        status,
        levels = c("n_species_missing", "n_species_gbif"),
        labels = c("Missing from GBIF", "In GBIF")
      )
    )
  
  # Get top classes by total species
  top_classes <- data |>
    group_by(class) |>
    summarise(total = sum(n_species_backbone), .groups = "drop") |>
    arrange(desc(total)) |>
    slice_head(n = n_top_classes) |>
    pull(class)
  
  plot_data <- plot_data |>
    filter(class %in% top_classes)
  
  # Calculate total species per order for sorting
  order_totals <- data |>
    filter(class %in% top_classes) |>
    group_by(class, order) |>
    summarise(total = sum(n_species_backbone), .groups = "drop")
  
  plot_data <- plot_data |>
    left_join(order_totals, by = c("class", "order"))
  
  # Create plot
  ggplot(plot_data, aes(x = reorder(order, total), y = n_species, fill = status)) +
    geom_col(alpha = 0.9, width = 0.8) +
    coord_flip() +
    facet_wrap(~class, scales = "free_y", ncol = 3) +
    scale_fill_manual(
      values = c("In GBIF" = "#2A9D8F", "Missing from GBIF" = "#457B9D"),
      name = NULL
    ) +
    scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.05))) +
    labs(
      title = glue("{kingdom_name}: Species Coverage by Order (Absolute)"),
      subtitle = "Bar length = species in backbone | Green = in GBIF | Blue = missing",
      x = NULL, 
      y = "Number of Species"
    ) +
    theme(
      legend.position = "bottom",
      strip.text = element_text(size = 9, face = "bold"),
      axis.text.y = element_text(size = 6),
      panel.spacing = unit(0.5, "lines")
    )
}

# Function to create percentage bar plot for a kingdom
plot_kingdom_orders_pct <- function(data, kingdom_name, n_top_classes = 12) {
  if (is.null(data) || nrow(data) == 0) {
    return(NULL)
  }
  
  # Get top classes by total species
  top_classes <- data |>
    group_by(class) |>
    summarise(total = sum(n_species_backbone), .groups = "drop") |>
    arrange(desc(total)) |>
    slice_head(n = n_top_classes) |>
    pull(class)
  
  # Calculate percentages
  plot_data <- data |>
    filter(class %in% top_classes) |>
    mutate(
      pct_gbif = 100 * n_species_gbif / n_species_backbone,
      pct_missing = 100 * n_species_missing / n_species_backbone
    ) |>
    select(class, order, n_species_backbone, pct_gbif, pct_missing) |>
    pivot_longer(
      cols = c(pct_gbif, pct_missing),
      names_to = "status",
      values_to = "pct"
    ) |>
    mutate(
      status = factor(
        status,
        levels = c("pct_missing", "pct_gbif"),
        labels = c("Missing from GBIF", "In GBIF")
      )
    )
  
  # Create plot - sort by coverage percentage (pct_gbif)
  order_pct <- data |>
    filter(class %in% top_classes) |>
    select(class, order, pct_coverage)
  
  plot_data <- plot_data |>
    left_join(order_pct, by = c("class", "order"))
  
  ggplot(plot_data, aes(x = reorder(order, pct_coverage), y = pct, fill = status)) +
    geom_col(alpha = 0.9, width = 0.8, position = "stack") +
    coord_flip() +
    facet_wrap(~class, scales = "free_y", ncol = 3) +
    scale_fill_manual(
      values = c("In GBIF" = "#2A9D8F", "Missing from GBIF" = "#457B9D"),
      name = NULL
    ) +
    scale_y_continuous(
      limits = c(0, 100), 
      breaks = c(0, 25, 50, 75, 100),
      expand = expansion(mult = c(0, 0))
    ) +
    labs(
      title = glue("{kingdom_name}: Species Coverage by Order (Percentage)"),
      subtitle = "All orders = 100% of backbone species | Green = % in GBIF | Blue = % missing",
      x = NULL, 
      y = "Coverage (%)"
    ) +
    theme(
      legend.position = "bottom",
      strip.text = element_text(size = 9, face = "bold"),
      axis.text.y = element_text(size = 6),
      panel.spacing = unit(0.5, "lines")
    )
}
```

### Animalia

```{r orders-animalia, fig.height=16, fig.width=14}
if (!is.null(order_coverage) && "kingdom" %in% names(order_coverage)) {
  animalia <- order_coverage |>
    filter(kingdom == "Animalia", !is.na(class), class != "")
  
  if (nrow(animalia) > 0) {
    print(plot_kingdom_orders(animalia, "Animalia", n_top_classes = 12))
  } else {
    cat("No Animalia data found.\n")
  }
}
```

```{r orders-animalia-pct, fig.height=16, fig.width=14}
if (exists("animalia") && nrow(animalia) > 0) {
  print(plot_kingdom_orders_pct(animalia, "Animalia", n_top_classes = 12))
}
```

```{r animalia-table}
if (exists("animalia") && nrow(animalia) > 0) {
  animalia |>
    group_by(class) |>
    summarise(
      n_orders = n(),
      species_backbone = sum(n_species_backbone),
      species_gbif = sum(n_species_gbif),
      species_missing = sum(n_species_missing),
      pct_coverage = round(100 * species_gbif / species_backbone, 1),
      .groups = "drop"
    ) |>
    arrange(desc(species_backbone)) |>
    kable(caption = "Animalia: Species Coverage by Class")
}
```

### Plantae

```{r orders-plantae, fig.height=12, fig.width=14}
if (!is.null(order_coverage) && "kingdom" %in% names(order_coverage)) {
  plantae <- order_coverage |>
    filter(kingdom == "Plantae", !is.na(class), class != "")
  
  if (nrow(plantae) > 0) {
    print(plot_kingdom_orders(plantae, "Plantae", n_top_classes = 9))
  } else {
    cat("No Plantae data found.\n")
  }
}
```

```{r orders-plantae-pct, fig.height=12, fig.width=14}
if (exists("plantae") && nrow(plantae) > 0) {
  print(plot_kingdom_orders_pct(plantae, "Plantae", n_top_classes = 9))
}
```

```{r plantae-table}
if (exists("plantae") && nrow(plantae) > 0) {
  plantae |>
    group_by(class) |>
    summarise(
      n_orders = n(),
      species_backbone = sum(n_species_backbone),
      species_gbif = sum(n_species_gbif),
      species_missing = sum(n_species_missing),
      pct_coverage = round(100 * species_gbif / species_backbone, 1),
      .groups = "drop"
    ) |>
    arrange(desc(species_backbone)) |>
    kable(caption = "Plantae: Species Coverage by Class")
}
```

### Fungi

```{r orders-fungi, fig.height=12, fig.width=14}
if (!is.null(order_coverage) && "kingdom" %in% names(order_coverage)) {
  fungi <- order_coverage |>
    filter(kingdom == "Fungi", !is.na(class), class != "")
  
  if (nrow(fungi) > 0) {
    print(plot_kingdom_orders(fungi, "Fungi", n_top_classes = 9))
  } else {
    cat("No Fungi data found.\n")
  }
}
```

```{r orders-fungi-pct, fig.height=12, fig.width=14}
if (exists("fungi") && nrow(fungi) > 0) {
  print(plot_kingdom_orders_pct(fungi, "Fungi", n_top_classes = 9))
}
```

```{r fungi-table}
if (exists("fungi") && nrow(fungi) > 0) {
  fungi |>
    group_by(class) |>
    summarise(
      n_orders = n(),
      species_backbone = sum(n_species_backbone),
      species_gbif = sum(n_species_gbif),
      species_missing = sum(n_species_missing),
      pct_coverage = round(100 * species_gbif / species_backbone, 1),
      .groups = "drop"
    ) |>
    arrange(desc(species_backbone)) |>
    kable(caption = "Fungi: Species Coverage by Class")
}
```

### Other Kingdoms

```{r orders-other, fig.height=10, fig.width=14}
if (!is.null(order_coverage) && "kingdom" %in% names(order_coverage)) {
  other <- order_coverage |>
    filter(!kingdom %in% c("Animalia", "Plantae", "Fungi"),
           !is.na(kingdom), kingdom != "",
           !is.na(class), class != "")
  
  if (nrow(other) > 0) {
    # For other kingdoms, facet by kingdom instead of class
    plot_data <- other |>
      select(kingdom, class, order, n_species_gbif, n_species_missing, n_species_backbone) |>
      pivot_longer(
        cols = c(n_species_gbif, n_species_missing),
        names_to = "status",
        values_to = "n_species"
      ) |>
      mutate(
        status = factor(
          status,
          levels = c("n_species_missing", "n_species_gbif"),
          labels = c("Missing from GBIF", "In GBIF")
        ),
        label = paste0(class, ": ", order)
      )
    
    ggplot(plot_data, aes(x = reorder(label, n_species_backbone), y = n_species, fill = status)) +
      geom_col(alpha = 0.9, width = 0.8) +
      coord_flip() +
      facet_wrap(~kingdom, scales = "free_y", ncol = 2) +
      scale_fill_manual(
        values = c("In GBIF" = "#2A9D8F", "Missing from GBIF" = "#457B9D"),
        name = NULL
      ) +
      scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.05))) +
      labs(
        title = "Other Kingdoms: Species Coverage by Order",
        subtitle = "Bar length = species in backbone | Green = in GBIF | Blue = missing",
        x = NULL, y = "Number of Species"
      ) +
      theme(
        legend.position = "bottom",
        strip.text = element_text(size = 10, face = "bold"),
        axis.text.y = element_text(size = 7)
      )
  } else {
    cat("No other kingdoms found.\n")
  }
}
```

### Summary by Kingdom

```{r kingdom-summary-table}
if (!is.null(order_coverage) && "kingdom" %in% names(order_coverage)) {
  order_coverage |>
    group_by(kingdom) |>
    summarise(
      n_orders = n(),
      species_backbone = sum(n_species_backbone),
      species_gbif = sum(n_species_gbif),
      species_missing = sum(n_species_missing),
      pct_coverage = round(100 * species_gbif / species_backbone, 1),
      .groups = "drop"
    ) |>
    arrange(desc(species_backbone)) |>
    kable(caption = "Species Coverage Summary by Kingdom")
}
```

---

# 2. Threatened Species

This section analyzes coverage of threatened species from backbone in GBIF data.

```{r threatened-data}
# Use match_summary which we know has the data
# Check for threat status columns - could be named differently

threatened_summary <- NULL
has_threat_data <- FALSE

if (!is.null(match_summary) && nrow(match_summary) > 0) {
  
  # Look for threat status column - try different possible names
  threat_col <- NULL
  possible_threat_cols <- c("threatStatus", "redlistCategory", "redlist_category", 
                            "swedish_redlist_category", "dyntaxa_redlist_category",
                            "RedlistCategory", "IUCN_category")
  
  for (col in possible_threat_cols) {
    if (col %in% names(match_summary)) {
      threat_col <- col
      break
    }
  }
  
  if (!is.null(threat_col)) {
    # Check if there's actual data in the column
    threat_values <- match_summary[[threat_col]]
    non_empty <- sum(!is.na(threat_values) & threat_values != "" & str_trim(threat_values) != "")
    
    if (non_empty > 0) {
      has_threat_data <- TRUE
      
      # Rename to threatStatus for consistency
      threatened_summary <- match_summary |>
        mutate(threatStatus = .data[[threat_col]]) |>
        filter(!is.na(threatStatus), threatStatus != "", str_trim(threatStatus) != "")
      
      cat(glue("**Total taxa with threat status in backbone**: {comma(nrow(threatened_summary))}\n\n"))
      cat(glue("**Found in GBIF**: {comma(sum(threatened_summary$matched_any, na.rm = TRUE))}\n\n"))
      cat(glue("**Missing from GBIF**: {comma(sum(!threatened_summary$matched_any, na.rm = TRUE))}\n\n"))
    }
  }
  
  if (!has_threat_data) {
    cat("
<div style='background-color: #fff3cd; border: 1px solid #ffc107; padding: 15px; border-radius: 5px; margin: 20px 0;'>
<h4 style='color: #856404; margin-top: 0;'>⚠️ No Threat Status Data Available</h4>
<p>The national backbone does not contain Red List threat categories (CR, EN, VU, NT, LC, etc.).</p>
<p><strong>To enable threatened species analysis:</strong></p>
<ol>
<li>Obtain the National Red List data (Artdatabanken)</li>
<li>Join it to the backbone backbone in script 03 (taxa_reference preparation)</li>
<li>Ensure the joined data has a column named <code>threatStatus_redlist</code> or <code>threatStatus_backbone</code></li>
<li>Re-run script 09 to regenerate the taxonomic match summary</li>
</ol>
</div>
")
  }
} else {
  cat("match_summary not available. Please ensure section 1.3 ran successfully.\n")
}
```

```{r threat-sections-conditional, results='asis'}
if (!has_threat_data) {
  cat("
## 2.1 Coverage by Threat Status
*Skipped - no threat status data available*

## 2.2 Missing Threatened Taxa
*Skipped - no threat status data available*

## 2.3 Missing Taxa by Taxonomic Group
*Skipped - no threat status data available*

## 2.4 Threatened Species with Poor Spatial Coverage
*Skipped - no threat status data available*
")
}
```

```{r threat-coverage, fig.height=7, fig.width=12, eval=has_threat_data}
# Only runs if has_threat_data is TRUE
if (!is.null(threatened_summary) && nrow(threatened_summary) > 0) {
  # Summarize by threat status
  threat_stats <- threatened_summary |>
    group_by(threatStatus) |>
    summarise(
      n_backbone = n(),
      n_gbif = sum(matched_any, na.rm = TRUE),
      n_missing = n_backbone - n_gbif,
      pct_coverage = round(100 * n_gbif / n_backbone, 1),
      .groups = "drop"
    ) |>
    filter(threatStatus %in% c("CR", "EN", "VU", "NT", "LC")) |>
    mutate(threatStatus = factor(threatStatus, levels = c("CR", "EN", "VU", "NT", "LC")))
  
  if (nrow(threat_stats) > 0) {
    # Plot 1: Absolute numbers (stacked bar)
    threat_long <- threat_stats |>
      select(threatStatus, n_gbif, n_missing) |>
      pivot_longer(cols = c(n_gbif, n_missing), names_to = "status", values_to = "count") |>
      mutate(
        status = factor(status, 
                        levels = c("n_missing", "n_gbif"),
                        labels = c("Missing from GBIF", "In GBIF"))
      )
    
    p1 <- ggplot(threat_long, aes(x = threatStatus, y = count, fill = status)) +
      geom_col(alpha = 0.9, width = 0.7) +
      geom_text(data = threat_stats, 
                aes(x = threatStatus, y = n_backbone, label = comma(n_backbone), fill = NULL),
                vjust = -0.3, size = 4) +
      scale_fill_manual(
        values = c("In GBIF" = "#2A9D8F", "Missing from GBIF" = "#457B9D"),
        name = NULL
      ) +
      scale_y_continuous(expand = expansion(mult = c(0, 0.1)), labels = comma) +
      labs(
        title = "Taxa Count by Threat Status",
        subtitle = "Total bar height = backbone taxa",
        x = "Threat Status", y = "Number of Taxa"
      ) +
      theme(legend.position = "bottom")
    
    # Plot 2: Coverage percentage
    p2 <- ggplot(threat_stats, aes(x = threatStatus, y = pct_coverage, fill = threatStatus)) +
      geom_col(alpha = 0.9, width = 0.7) +
      geom_text(aes(label = paste0(pct_coverage, "%")), vjust = -0.3, size = 4) +
      scale_fill_manual(values = threat_colors) +
      scale_y_continuous(limits = c(0, 105), expand = expansion(mult = c(0, 0))) +
      labs(
        title = "GBIF Coverage by Threat Status",
        subtitle = "% of backbone taxa found in GBIF",
        x = "Threat Status", y = "Coverage (%)"
      ) +
      theme(legend.position = "none")
    
    print(p1 + p2 + plot_annotation(
      title = "Threatened Species Coverage in GBIF",
      subtitle = "CR=Critically Endangered, EN=Endangered, VU=Vulnerable, NT=Near Threatened, LC=Least Concern"
    ))
  }
}
```

```{r threat-table, eval=has_threat_data}
if (exists("threat_stats") && !is.null(threat_stats) && nrow(threat_stats) > 0) {
  threat_stats |>
    select(threatStatus, n_backbone, n_gbif, n_missing, pct_coverage) |>
    kable(
      caption = "Threatened Species Coverage Summary",
      col.names = c("Threat Status", "In Backbone", "In GBIF", "Missing", "Coverage %")
    )
}
```

```{r missing-threatened, fig.height=8, fig.width=12, eval=has_threat_data}
if (!is.null(threatened_summary) && nrow(threatened_summary) > 0) {
  # Get missing threatened taxa
  missing_threatened <- threatened_summary |>
    filter(!matched_any) |>
    filter(threatStatus %in% c("CR", "EN", "VU", "NT"))
  
  if (nrow(missing_threatened) > 0) {
    cat(glue("**{comma(nrow(missing_threatened))} threatened taxa (CR/EN/VU/NT)** are missing from GBIF.\n\n"))
    
    # Summary by threat status
    missing_by_threat <- missing_threatened |>
      group_by(threatStatus) |>
      summarise(n = n(), .groups = "drop") |>
      mutate(threatStatus = factor(threatStatus, levels = c("CR", "EN", "VU", "NT")))
    
    # Bar plot of missing taxa
    print(ggplot(missing_by_threat, aes(x = threatStatus, y = n, fill = threatStatus)) +
      geom_col(alpha = 0.9, width = 0.6) +
      geom_text(aes(label = comma(n)), vjust = -0.3, size = 5) +
      scale_fill_manual(values = threat_colors) +
      scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
      labs(
        title = "Threatened Taxa Missing from GBIF",
        subtitle = "Number of backbone species with no GBIF records",
        x = "Threat Status", y = "Number of Missing Taxa"
      ) +
      theme(legend.position = "none"))
  } else {
    cat("All threatened taxa are represented in GBIF!\n")
  }
}
```

```{r missing-threatened-table, eval=has_threat_data, results='asis'}
if (exists("missing_threatened") && !is.null(missing_threatened) && nrow(missing_threatened) > 0) {
  # Get CR and EN taxa
  cr_en <- missing_threatened |>
    filter(threatStatus %in% c("CR", "EN")) |>
    arrange(threatStatus, kingdom, phylum, class, order, family, scientificName)
  
  if (nrow(cr_en) > 0) {
    cat("\n### Complete List of Missing CR/EN Taxa\n\n")
    cat(glue("**Total: {nrow(cr_en)} taxa** ({sum(cr_en$threatStatus == 'CR')} CR, {sum(cr_en$threatStatus == 'EN')} EN)\n\n"))
    
    # Group by kingdom
    kingdoms <- unique(cr_en$kingdom)
    kingdoms <- kingdoms[!is.na(kingdoms) & kingdoms != ""]
    
    for (k in kingdoms) {
      k_data <- cr_en |> filter(kingdom == k)
      cat(glue("\n## {k} ({nrow(k_data)} taxa)\n\n"))
      
      # Group by phylum within kingdom
      phyla <- unique(k_data$phylum)
      phyla <- phyla[!is.na(phyla) & phyla != ""]
      
      for (p in phyla) {
        p_data <- k_data |> filter(phylum == p)
        cat(glue("\n### {p} ({nrow(p_data)} taxa)\n\n"))
        
        # Group by class within phylum
        classes <- unique(p_data$class)
        classes <- classes[!is.na(classes) & classes != ""]
        
        for (cl in classes) {
          cl_data <- p_data |> filter(class == cl)
          cat(glue("\n#### {cl} ({nrow(cl_data)} taxa)\n\n"))
          
          # Group by order within class
          orders <- unique(cl_data$order)
          orders <- orders[!is.na(orders) & orders != ""]
          
          for (o in orders) {
            o_data <- cl_data |> filter(order == o)
            cat(glue("\n**{o}** ({nrow(o_data)} taxa)\n\n"))
            
            # Group by family within order
            families <- unique(o_data$family)
            families <- families[!is.na(families) & families != ""]
            
            for (f in families) {
              f_data <- o_data |> filter(family == f)
              
              # List species
              species_list <- f_data |>
                mutate(entry = glue("- {scientificName} ({threatStatus})")) |>
                pull(entry)
              
              cat(glue("\n*{f}*\n\n"))
              cat(paste(species_list, collapse = "\n"), "\n\n")
            }
            
            # Handle species without family
            no_family <- o_data |> filter(is.na(family) | family == "")
            if (nrow(no_family) > 0) {
              species_list <- no_family |>
                mutate(entry = glue("- {scientificName} ({threatStatus})")) |>
                pull(entry)
              cat("\n*(Family unknown)*\n\n")
              cat(paste(species_list, collapse = "\n"), "\n\n")
            }
          }
          
          # Handle species without order
          no_order <- cl_data |> filter(is.na(order) | order == "")
          if (nrow(no_order) > 0) {
            cat("\n**Order unknown**\n\n")
            species_list <- no_order |>
              mutate(entry = glue("- {scientificName} ({threatStatus}) - Family: {family}")) |>
              pull(entry)
            cat(paste(species_list, collapse = "\n"), "\n\n")
          }
        }
        
        # Handle species without class
        no_class <- p_data |> filter(is.na(class) | class == "")
        if (nrow(no_class) > 0) {
          cat("\n#### Class unknown\n\n")
          species_list <- no_class |>
            mutate(entry = glue("- {scientificName} ({threatStatus})")) |>
            pull(entry)
          cat(paste(species_list, collapse = "\n"), "\n\n")
        }
      }
      
      # Handle species without phylum
      no_phylum <- k_data |> filter(is.na(phylum) | phylum == "")
      if (nrow(no_phylum) > 0) {
        cat("\n### Phylum unknown\n\n")
        species_list <- no_phylum |>
          mutate(entry = glue("- {scientificName} ({threatStatus})")) |>
          pull(entry)
        cat(paste(species_list, collapse = "\n"), "\n\n")
      }
    }
    
    # Handle species without kingdom
    no_kingdom <- cr_en |> filter(is.na(kingdom) | kingdom == "")
    if (nrow(no_kingdom) > 0) {
      cat("\n## Kingdom unknown\n\n")
      species_list <- no_kingdom |>
        mutate(entry = glue("- {scientificName} ({threatStatus})")) |>
        pull(entry)
      cat(paste(species_list, collapse = "\n"), "\n\n")
    }
    
  } else {
    cat("No CR or EN taxa are missing from GBIF.\n")
  }
}
```

```{r missing-by-order, fig.height=10, fig.width=12, eval=has_threat_data}
if (exists("missing_threatened") && !is.null(missing_threatened) && nrow(missing_threatened) > 0 && "order" %in% names(missing_threatened)) {
  # Summarize missing by order
  missing_by_order <- missing_threatened |>
    filter(!is.na(order), order != "") |>
    group_by(order, threatStatus) |>
    summarise(n = n(), .groups = "drop") |>
    mutate(threatStatus = factor(threatStatus, levels = c("CR", "EN", "VU", "NT")))
  
  # Get top 20 orders by total missing
  top_orders <- missing_by_order |>
    group_by(order) |>
    summarise(total = sum(n), .groups = "drop") |>
    arrange(desc(total)) |>
    slice_head(n = 20) |>
    pull(order)
  
  missing_by_order_top <- missing_by_order |>
    filter(order %in% top_orders)
  
  if (nrow(missing_by_order_top) > 0) {
    print(ggplot(missing_by_order_top, aes(x = fct_reorder(order, n, .fun = sum), y = n, fill = threatStatus)) +
      geom_col(alpha = 0.9) +
      coord_flip() +
      scale_fill_manual(values = threat_colors, name = "Threat Status") +
      labs(
        title = "Missing Threatened Taxa by Order (Top 20)",
        subtitle = "Orders with most threatened species not in GBIF",
        x = NULL, y = "Number of Missing Taxa"
      ) +
      theme(legend.position = "right"))
  }
}
```

```{r threatened-spatial, fig.height=8, fig.width=12, eval=has_threat_data}
# For threatened species that ARE in GBIF, how well are they covered spatially?
if (!is.null(threatened_summary) && nrow(threatened_summary) > 0) {
  # We need spatial data - check if it's in match_summary or load separately

  spatial_data <- safe_read(here(p_gaps, "taxonomic_spatial_coverage.csv"))
  
  if (!is.null(spatial_data) && nrow(spatial_data) > 0) {
    # Prepare threatened summary with only needed columns
    threatened_for_join <- threatened_summary |>
      filter(matched_any, threatStatus %in% c("CR", "EN", "VU", "NT")) |>
      select(any_of(c("scientificName", "scientific_std", "threatStatus")))
    
    if (nrow(threatened_for_join) > 0) {
      # Store threat status before join
      threat_lookup <- threatened_for_join |>
        select(any_of(c("scientificName", "scientific_std", "threatStatus"))) |>
        distinct()
      
      # Try to match on scientific name
      if ("scientificName" %in% names(spatial_data) && "scientificName" %in% names(threat_lookup)) {
        # Remove threatStatus from spatial_data if present to avoid conflicts
        spatial_cols <- setdiff(names(spatial_data), "threatStatus")
        threatened_spatial <- spatial_data |>
          select(all_of(spatial_cols)) |>
          inner_join(threat_lookup, by = "scientificName")
      } else if ("scientific_std" %in% names(spatial_data) && "scientific_std" %in% names(threat_lookup)) {
        spatial_cols <- setdiff(names(spatial_data), "threatStatus")
        threatened_spatial <- spatial_data |>
          select(all_of(spatial_cols)) |>
          inner_join(threat_lookup, by = "scientific_std")
      } else {
        threatened_spatial <- data.frame()
      }
      
      if (nrow(threatened_spatial) > 0 && "n_cells_10km" %in% names(threatened_spatial) && "threatStatus" %in% names(threatened_spatial)) {
        threatened_spatial <- threatened_spatial |>
          filter(n_cells_10km > 0) |>
          mutate(threatStatus = factor(threatStatus, levels = c("CR", "EN", "VU", "NT")))
        
        if (nrow(threatened_spatial) > 0) {
          print(ggplot(threatened_spatial, aes(x = n_cells_10km, fill = threatStatus)) +
            geom_histogram(bins = 30, alpha = 0.8) +
            facet_wrap(~threatStatus, scales = "free_y", ncol = 2) +
            scale_fill_manual(values = threat_colors) +
            scale_x_log10(labels = comma) +
            labs(
              title = "Spatial Coverage of Threatened Species in GBIF",
              subtitle = "Number of 10km grid cells with records",
              x = "Number of Cells (log scale)", y = "Number of Species"
            ) +
            theme(legend.position = "none"))
        } else {
          cat("No threatened species with spatial coverage data after filtering.\n")
        }
      } else {
        cat("No spatial coverage data available for threatened species.\n")
      }
    } else {
      cat("No threatened species (CR/EN/VU/NT) found in GBIF.\n")
    }
  } else {
    cat("Spatial coverage file not found.\n")
  }
}
```

---

# 3. Coverage by Higher Taxonomy

## 3.1 Coverage by Order

```{r order-coverage, fig.height=10}
tax_by_order <- safe_read(here(p_tables, "overview_taxonomic_gaps_by_order.csv"))

if (!is.null(tax_by_order) && nrow(tax_by_order) > 0) {
  order_plot <- tax_by_order |>
    filter(!is.na(order), order != "", n_taxa >= 10) |>
    arrange(desc(n_taxa)) |>
    slice_head(n = 20)
  
  if (nrow(order_plot) > 0 && "pct_coverage" %in% names(order_plot)) {
    ggplot(order_plot, aes(x = fct_reorder(order, pct_coverage), 
                            y = pct_coverage, fill = pct_coverage)) +
      geom_col(alpha = 0.9) +
      geom_text(aes(label = paste0(round(pct_coverage, 0), "%")), hjust = -0.1, size = 3.5) +
      coord_flip() +
      scale_fill_viridis_c(option = "plasma", begin = 0.1, end = 0.9) +
      scale_y_continuous(limits = c(0, 110), expand = expansion(mult = c(0, 0))) +
      labs(
        title = "GBIF Coverage by Order (Top 20 by taxa count)",
        subtitle = "Orders with at least 10 taxa in backbone",
        x = NULL, y = "Coverage (%)"
      ) +
      theme(legend.position = "none")
  } else {
    cat("Order coverage data available but missing pct_coverage column.\n")
    order_plot |>
      select(any_of(c("order", "n_taxa", "n_in_gbif", "n_missing"))) |>
      kable(caption = "Order Coverage Data")
  }
} else {
  cat("Order coverage file not found or empty.\n")
}
```

## 3.2 Coverage by Family

```{r family-coverage, fig.height=12}
tax_by_family <- safe_read(here(p_tables, "overview_taxonomic_gaps_by_family.csv"))

if (!is.null(tax_by_family) && nrow(tax_by_family) > 0) {
  family_plot <- tax_by_family |>
    filter(!is.na(family), family != "", n_taxa >= 10) |>
    arrange(desc(n_taxa)) |>
    slice_head(n = 25)
  
  if (nrow(family_plot) > 0 && "pct_coverage" %in% names(family_plot)) {
    ggplot(family_plot, aes(x = fct_reorder(family, pct_coverage), 
                             y = pct_coverage, fill = pct_coverage)) +
      geom_col(alpha = 0.9) +
      geom_text(aes(label = paste0(round(pct_coverage, 0), "%")), hjust = -0.1, size = 3) +
      coord_flip() +
      scale_fill_viridis_c(option = "mako", begin = 0.2, end = 0.9) +
      scale_y_continuous(limits = c(0, 110), expand = expansion(mult = c(0, 0))) +
      labs(
        title = "GBIF Coverage by Family (Top 25 by taxa count)",
        subtitle = "Families with at least 10 taxa in backbone",
        x = NULL, y = "Coverage (%)"
      ) +
      theme(legend.position = "none")
  } else {
    cat("Family coverage data available but missing pct_coverage column.\n")
    family_plot |>
      select(any_of(c("family", "n_taxa", "n_in_gbif", "n_missing"))) |>
      kable(caption = "Family Coverage Data")
  }
} else {
  cat("Family coverage file not found or empty.\n")
}
```

---

# 4. Data Mobilization Trends

## 4.1 Top 10 Orders by Phylum (5-Year Trends)

```{r order-by-phylum, fig.height=16, fig.width=14}
order_temporal <- safe_read(here(p_integrated, "order_temporal_trends.csv"))

if (!is.null(order_temporal) && nrow(order_temporal) > 0) {
  # Get phylum info - try to join with match data
  if (!is.null(tax_match)) {
    phylum_lookup <- tax_match |>
      filter(!is.na(order), !is.na(phylum)) |>
      select(order, phylum) |>
      distinct()
    
    order_with_phylum <- order_temporal |>
      filter(grid == "grid10km") |>
      left_join(phylum_lookup, by = "order")
  } else {
    # If no match data, try to get phylum from order_temporal if available
    order_with_phylum <- order_temporal |>
      filter(grid == "grid10km")
    if (!"phylum" %in% names(order_with_phylum)) {
      order_with_phylum$phylum <- "Unknown"
    }
  }
  
  # Aggregate to 5-year periods
  order_5yr <- order_with_phylum |>
    filter(year >= 1970) |>
    mutate(period = paste0(floor(year / 5) * 5, "-", floor(year / 5) * 5 + 4)) |>
    group_by(phylum, order, period) |>
    summarise(occurrences = sum(as.numeric(total_occurrences), na.rm = TRUE), .groups = "drop")
  
  # Get top 10 orders per phylum
  top_orders_by_phylum <- order_5yr |>
    group_by(phylum, order) |>
    summarise(total = sum(as.numeric(occurrences)), .groups = "drop") |>
    group_by(phylum) |>
    arrange(desc(total)) |>
    slice_head(n = 10) |>
    ungroup()
  
  # Filter to top orders and plot
  order_5yr_top <- order_5yr |>
    inner_join(top_orders_by_phylum |> select(phylum, order), by = c("phylum", "order"))
  
  # Get unique phyla with data
  phyla_with_data <- unique(order_5yr_top$phylum)
  phyla_with_data <- phyla_with_data[!is.na(phyla_with_data) & phyla_with_data != "Unknown"]
  
  if (length(phyla_with_data) > 0) {
    # Plot for each major phylum
    for (phy in phyla_with_data[1:min(4, length(phyla_with_data))]) {
      phy_data <- order_5yr_top |> filter(phylum == phy)
      
      if (nrow(phy_data) > 0) {
        p <- ggplot(phy_data, aes(x = period, y = occurrences, fill = order)) +
          geom_col(alpha = 0.9) +
          facet_wrap(~order, scales = "free_y", ncol = 2) +
          scale_fill_viridis_d(option = "turbo") +
          scale_y_continuous(labels = comma) +
          labs(
            title = glue("Data Mobilization: {phy}"),
            subtitle = "Top 10 orders by total occurrences (5-year periods)",
            x = "Period", y = "Occurrences"
          ) +
          theme(
            axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
            legend.position = "none",
            strip.text = element_text(size = 9)
          )
        print(p)
      }
    }
  } else {
    cat("No phylum data available for orders.\n")
  }
} else {
  cat("Order temporal trends file not found.\n")
}
```

## 4.2 Recent vs Historical Change

```{r order-change, fig.height=8, fig.width=12}
top_orders <- safe_read(here(p_integrated, "top_orders.csv"))

# If order_temporal exists from section 4.1, use it
if (!is.null(order_temporal) && nrow(order_temporal) > 0) {
  
  # If top_orders file doesn't exist, derive from order_temporal
  if (is.null(top_orders) || nrow(top_orders) == 0) {
    cat("Deriving top orders from temporal data...\n\n")
    top_orders <- order_temporal |>
      filter(grid == "grid10km") |>
      group_by(order) |>
      summarise(total = sum(as.numeric(total_occurrences), na.rm = TRUE), .groups = "drop") |>
      arrange(desc(total)) |>
      slice_head(n = 12)
  }
  
  if (nrow(top_orders) > 0) {
    recent_cutoff <- current_year - 10
    historical_cutoff <- current_year - 20
    
    top_order_names <- top_orders$order[1:min(12, nrow(top_orders))]
    
    order_change <- order_temporal |>
      filter(order %in% top_order_names, grid == "grid10km") |>
      mutate(
        era = case_when(
          year >= recent_cutoff ~ "Recent",
          year >= historical_cutoff ~ "Historical",
          TRUE ~ NA_character_
        )
      ) |>
      filter(!is.na(era))
    
    if (nrow(order_change) > 0) {
      order_change_agg <- order_change |>
        group_by(order, era) |>
        summarise(occ = sum(as.numeric(total_occurrences), na.rm = TRUE), .groups = "drop")
      
      order_change_wide <- order_change_agg |>
        pivot_wider(names_from = era, values_from = occ, values_fill = 0)
      
      # Check if both columns exist
      if ("Recent" %in% names(order_change_wide) && "Historical" %in% names(order_change_wide)) {
        order_change_wide <- order_change_wide |>
          filter(Historical > 0) |>  # Avoid division by zero
          mutate(
            pct_change = 100 * (Recent - Historical) / Historical,
            direction = if_else(pct_change >= 0, "Increased", "Decreased")
          ) |>
          arrange(desc(pct_change))
        
        if (nrow(order_change_wide) > 0) {
          # Create label positions
          order_change_wide <- order_change_wide |>
            mutate(
              label_hjust = if_else(pct_change >= 0, -0.1, 1.1),
              label = paste0(if_else(pct_change >= 0, "+", ""), round(pct_change, 0), "%")
            )
          
          ggplot(order_change_wide, aes(x = fct_reorder(order, pct_change), 
                                         y = pct_change, fill = direction)) +
            geom_col(alpha = 0.9) +
            geom_text(aes(label = label, hjust = label_hjust), size = 4) +
            coord_flip() +
            scale_fill_manual(values = c("Increased" = "#2A9D8F", "Decreased" = "#E76F51")) +
            geom_hline(yintercept = 0, linetype = "dashed") +
            labs(
              title = "Change in Data Mobilization by Order",
              subtitle = glue("Comparing {recent_cutoff}-{current_year} vs {historical_cutoff}-{recent_cutoff - 1}"),
              x = NULL, y = "% Change"
            ) +
            theme(legend.position = "none")
        } else {
          cat("No orders with historical data (before ", historical_cutoff, ") found.\n")
        }
      } else {
        cat("Missing Recent or Historical era data.\n")
        cat("Available eras: ", paste(names(order_change_wide), collapse = ", "), "\n")
      }
    } else {
      cat("No data within the specified time ranges.\n")
      cat("Recent cutoff: ", recent_cutoff, ", Historical cutoff: ", historical_cutoff, "\n")
      cat("Year range in data: ", min(order_temporal$year, na.rm = TRUE), " - ", max(order_temporal$year, na.rm = TRUE), "\n")
    }
  } else {
    cat("No top orders available.\n")
  }
} else {
  cat("Order temporal trends file not found or empty.\n")
  cat("Expected location: ", here(p_integrated, "order_temporal_trends.csv"), "\n")
}
```

---

# 5. Priority Taxa

Priority taxa are species that should be targeted for data collection efforts. These include:
- Threatened species (CR, EN, VU, NT) missing from GBIF
- Threatened species with poor spatial coverage in GBIF

## 5.1 Overview

```{r priority-overview}
# Try to load priority_taxa file first
priority_taxa <- safe_read(here(p_integrated, "priority_taxa_all.csv"))

# If not available, derive from threatened_summary (from section 2)
if ((is.null(priority_taxa) || nrow(priority_taxa) == 0) && exists("threatened_summary") && !is.null(threatened_summary)) {
  cat("Deriving priority taxa from threatened species analysis...\n\n")
  
  # Priority = threatened species missing from GBIF or with poor coverage
  priority_taxa <- threatened_summary |>
    filter(threatStatus %in% c("CR", "EN", "VU", "NT")) |>
    filter(!matched_any) |>  # Missing from GBIF
    mutate(priority_reason = "Missing from GBIF")
}

if (!is.null(priority_taxa) && nrow(priority_taxa) > 0) {
  cat(glue("**{comma(nrow(priority_taxa))} taxa** require priority attention.\n\n"))
  
  # Check for threatStatus column
 threat_col <- intersect(c("threatStatus", "threatStatus_redlist", "threatStatus_backbone"), names(priority_taxa))[1]
  
  if (!is.na(threat_col)) {
    priority_taxa <- priority_taxa |>
      mutate(threatStatus = .data[[threat_col]])
    
    priority_by_threat <- priority_taxa |>
      filter(!is.na(threatStatus), threatStatus != "") |>
      group_by(threatStatus) |>
      summarise(n = n(), .groups = "drop") |>
      filter(threatStatus %in% c("CR", "EN", "VU", "NT", "LC")) |>
      arrange(match(threatStatus, c("CR", "EN", "VU", "NT", "LC")))
    
    if (nrow(priority_by_threat) > 0) {
      for (i in 1:nrow(priority_by_threat)) {
        cat(glue("- **{priority_by_threat$threatStatus[i]}**: {priority_by_threat$n[i]} taxa\n\n"))
      }
    }
  } else {
    cat("No threat status column found in priority taxa.\n")
  }
} else {
  cat("No priority taxa identified.\n")
  cat("This could mean:\n")
  cat("- All threatened species are already in GBIF, or\n
")
  cat("- No threatened species data is available in the reference taxonomy\n")
}
```

## 5.2 Highest Priority (CR/EN)

```{r priority-cr-en}
if (!is.null(priority_taxa) && nrow(priority_taxa) > 0 && "threatStatus" %in% names(priority_taxa)) {
  cr_en_priority <- priority_taxa |>
    filter(threatStatus %in% c("CR", "EN"))
  
  if (nrow(cr_en_priority) > 0) {
    cat(glue("**{nrow(cr_en_priority)} Critically Endangered and Endangered taxa** are priority for data mobilization.\n\n"))
    
    cr_en_priority |>
      select(any_of(c("scientificName", "threatStatus", "taxonRank", "family", "order", "class", "priority_reason"))) |>
      arrange(threatStatus, order, family, scientificName) |>
      slice_head(n = 50) |>
      kable(caption = "Priority CR/EN Taxa (First 50)")
  } else {
    cat("No CR/EN taxa in priority list - all may already be in GBIF.\n")
  }
} else {
  cat("No priority taxa data available.\n")
}
```

## 5.3 Priority by Family

```{r priority-by-family, fig.height=8, fig.width=10}
if (!is.null(priority_taxa) && nrow(priority_taxa) > 0 && "family" %in% names(priority_taxa)) {
  priority_by_family <- priority_taxa |>
    filter(!is.na(family), family != "") |>
    group_by(family) |>
    summarise(n = n(), .groups = "drop") |>
    arrange(desc(n)) |>
    slice_head(n = 20)
  
  if (nrow(priority_by_family) > 0) {
    ggplot(priority_by_family, aes(x = fct_reorder(family, n), y = n, fill = n)) +
      geom_col(alpha = 0.9) +
      geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
      coord_flip() +
      scale_fill_viridis_c(option = "rocket", direction = -1) +
      scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
      labs(
        title = "Priority Taxa by Family (Top 20)", 
        subtitle = "Families with most threatened species missing from GBIF",
        x = NULL, y = "Number of Priority Taxa"
      ) +
      theme(legend.position = "none")
  } else {
    cat("No family data available for priority taxa.\n")
  }
} else {
  cat("No priority taxa with family information available.\n")
}
```

## 5.4 Priority by Order

```{r priority-by-order, fig.height=8, fig.width=10}
if (!is.null(priority_taxa) && nrow(priority_taxa) > 0 && "order" %in% names(priority_taxa)) {
  priority_by_order <- priority_taxa |>
    filter(!is.na(order), order != "") |>
    group_by(order) |>
    summarise(n = n(), .groups = "drop") |>
    arrange(desc(n)) |>
    slice_head(n = 20)
  
  if (nrow(priority_by_order) > 0) {
    ggplot(priority_by_order, aes(x = fct_reorder(order, n), y = n, fill = n)) +
      geom_col(alpha = 0.9) +
      geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
      coord_flip() +
      scale_fill_viridis_c(option = "mako", direction = -1) +
      scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
      labs(
        title = "Priority Taxa by Order (Top 20)", 
        subtitle = "Orders with most threatened species missing from GBIF",
        x = NULL, y = "Number of Priority Taxa"
      ) +
      theme(legend.position = "none")
  } else {
    cat("No order data available for priority taxa.\n")
  }
} else {
  cat("No priority taxa with order information available.\n")
}
```

---

# 6. Summary & Recommendations

```{r summary, results='asis'}
cat("## Key Findings\n\n")

if (!is.null(dashboard)) {
  cat(glue("- **Overall coverage**: {dashboard$taxa_pct_coverage}% ({comma(dashboard$taxa_in_gbif)} of {comma(dashboard$taxa_in_reference)} taxa)"), "\n\n")
}

if (!is.null(missing_threatened) && nrow(missing_threatened) > 0) {
  n_cr <- sum(missing_threatened$threatStatus == "CR", na.rm = TRUE)
  n_en <- sum(missing_threatened$threatStatus == "EN", na.rm = TRUE)
  cat(glue("- **Missing threatened taxa**: {comma(nrow(missing_threatened))}"), "\n")
  cat(glue("  - CR: {n_cr}, EN: {n_en}"), "\n\n")
}

if (!is.null(priority_taxa) && nrow(priority_taxa) > 0) {
  cat(glue("- **Priority taxa**: {comma(nrow(priority_taxa))}"), "\n\n")
}

cat("\n## Recommendations\n\n")
cat("1. **Highest priority**: Focus on CR and EN species missing from GBIF\n\n")
cat("2. **Target underrepresented families** for systematic surveys\n\n")
cat("3. **Digitize museum specimens** for poorly covered taxonomic groups\n\n")
cat("4. **Maintain mobilization momentum** for orders showing growth\n\n")
cat("5. **Investigate declining orders** and address data collection gaps\n\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
