---
title: "Taxonomic Gap Analysis"
subtitle: "GBIF Sweden Data Gap Analysis - Species Coverage & Mobilization"
author: "GBIF Sweden"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    fig_width: 12
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 300
)

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(viridis)
library(patchwork)
library(sf)
library(knitr)
library(data.table)
library(glue)
library(lubridate)

# Theme
theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12, color = "grey40"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
)

# Paths
p_gaps <- here("data_proc", "gaps")
p_tables <- here("output", "tables")
p_data_proc <- here("data_proc")
p_derived <- here("data_proc", "derived")
p_integrated <- here("output", "tables", "integrated")

# Current year
current_year <- year(Sys.Date())

# Threat status order and colors
threat_levels <- c("CR", "EN", "VU", "NT", "LC", "DD", "NE", NA)
threat_colors <- c("CR" = "#D32F2F", "EN" = "#E64A19", "VU" = "#F57C00", 
                   "NT" = "#FBC02D", "LC" = "#8BC34A", "DD" = "#78909C", 
                   "NE" = "#B0BEC5")
```

# Overview

This report analyzes **taxonomic gaps** in GBIF Sweden's biodiversity occurrence data, comparing GBIF coverage against the Swedish Red List reference taxonomy and examining how data mobilization has changed for different taxonomic groups.

**Key questions addressed:**

- What proportion of Sweden's known species are represented in GBIF?
- Which taxonomic groups have the best/worst coverage?
- How well are threatened species represented?
- How has data mobilization changed for different taxonomic orders over time?
- Which species should be prioritized for data collection?

---

# 1. Overall Taxonomic Coverage

## 1.1 Coverage Summary

```{r load-taxonomic}
# Load taxonomic summary
if (file.exists(here(p_tables, "overview_taxonomic_summary.csv"))) {
  tax_summary <- read_csv(here(p_tables, "overview_taxonomic_summary.csv"), show_col_types = FALSE)
} else if (file.exists(here(p_gaps, "taxonomic_coverage_by_rank.csv"))) {
  tax_summary <- read_csv(here(p_gaps, "taxonomic_coverage_by_rank.csv"), show_col_types = FALSE)
}

# Load dashboard for overall stats
if (file.exists(here(p_tables, "dashboard_summary.csv"))) {
  dashboard <- read_csv(here(p_tables, "dashboard_summary.csv"), show_col_types = FALSE)
  
  cat("**Overall Taxonomic Coverage:**\n\n")
  cat("- Taxa in reference:", comma(dashboard$taxa_in_reference), "\n")
  cat("- Taxa in GBIF:", comma(dashboard$taxa_in_gbif), "\n")
  cat("- Coverage:", dashboard$taxa_pct_coverage, "%\n")
  cat("- Priority taxa:", comma(dashboard$n_priority_taxa), "\n\n")
}
```

## 1.2 Coverage by Taxonomic Rank

```{r coverage-by-rank, fig.height=7}
if (exists("tax_summary")) {
  tax_by_rank <- tax_summary |>
    filter(!is.na(taxonRank), n_ref_total >= 10) |>
    arrange(desc(n_ref_total)) |>
    slice_head(n = 15)
  
  tax_by_rank |>
    select(taxonRank, n_ref_total, n_in_gbif, pct_coverage) |>
    kable(caption = "Taxonomic Coverage by Rank (Top 15)", digits = 1)
  
  ggplot(tax_by_rank, aes(x = fct_reorder(taxonRank, pct_coverage), 
                           y = pct_coverage, fill = pct_coverage)) +
    geom_col() +
    geom_text(aes(label = paste0(round(pct_coverage, 1), "%")), 
              hjust = -0.1, size = 4) +
    coord_flip() +
    scale_fill_viridis_c(option = "viridis", begin = 0.2, end = 0.9) +
    labs(
      title = "GBIF Coverage by Taxonomic Rank",
      subtitle = "Percentage of reference taxa found in GBIF occurrences",
      x = NULL,
      y = "Coverage (%)"
    ) +
    theme(legend.position = "none") +
    ylim(0, 110)
}
```

---

# 2. Threatened Species Coverage

## 2.1 Coverage by Threat Status

```{r threat-coverage}
if (file.exists(here(p_gaps, "taxonomic_coverage_by_threat.csv"))) {
  threat_coverage <- read_csv(here(p_gaps, "taxonomic_coverage_by_threat.csv"), show_col_types = FALSE)
  
  threat_display <- threat_coverage |>
    filter(threatStatus %in% c("CR", "EN", "VU", "NT", "LC")) |>
    mutate(threatStatus = factor(threatStatus, levels = c("CR", "EN", "VU", "NT", "LC")))
  
  threat_display |>
    select(threatStatus, n_ref_total, n_in_gbif, pct_coverage) |>
    kable(caption = "Coverage by IUCN Threat Status", digits = 1)
}
```

```{r threat-plot, fig.height=6}
if (exists("threat_display")) {
  p1 <- ggplot(threat_display, aes(x = threatStatus, y = n_ref_total, fill = threatStatus)) +
    geom_col() +
    geom_text(aes(label = comma(n_ref_total)), vjust = -0.3, size = 4) +
    scale_fill_manual(values = threat_colors) +
    labs(title = "Taxa in Reference", y = "Count", x = NULL) +
    theme(legend.position = "none")
  
  p2 <- ggplot(threat_display, aes(x = threatStatus, y = pct_coverage, fill = threatStatus)) +
    geom_col() +
    geom_text(aes(label = paste0(round(pct_coverage, 1), "%")), vjust = -0.3, size = 4) +
    scale_fill_manual(values = threat_colors) +
    labs(title = "GBIF Coverage", y = "Coverage (%)", x = NULL) +
    theme(legend.position = "none") +
    ylim(0, 110)
  
  p1 + p2 +
    plot_annotation(
      title = "Threatened Species in GBIF Sweden",
      subtitle = "Reference count and coverage by IUCN Red List category"
    )
}
```

## 2.2 Missing Threatened Taxa

```{r missing-threatened}
if (file.exists(here(p_gaps, "taxonomic_missing_taxa.csv"))) {
  missing_taxa <- fread(here(p_gaps, "taxonomic_missing_taxa.csv"))
  
  missing_threatened <- missing_taxa |>
    filter(threatStatus %in% c("CR", "EN", "VU", "NT")) |>
    mutate(threatStatus = factor(threatStatus, levels = c("CR", "EN", "VU", "NT")))
  
  cat("**Missing Threatened Taxa:**\n\n")
  cat("- Critically Endangered (CR):", sum(missing_threatened$threatStatus == "CR"), "\n")
  cat("- Endangered (EN):", sum(missing_threatened$threatStatus == "EN"), "\n")
  cat("- Vulnerable (VU):", sum(missing_threatened$threatStatus == "VU"), "\n")
  cat("- Near Threatened (NT):", sum(missing_threatened$threatStatus == "NT"), "\n\n")
  
  # Show sample of missing CR/EN taxa
  missing_threatened |>
    filter(threatStatus %in% c("CR", "EN")) |>
    arrange(threatStatus, scientificName) |>
    select(scientificName, threatStatus, taxonRank, family, order) |>
    slice_head(n = 25) |>
    kable(caption = "Sample of Missing CR/EN Taxa (Top 25)")
}
```

---

# 3. Coverage by Higher Taxonomy

## 3.1 Coverage by Order

```{r order-coverage, fig.height=8}
if (file.exists(here(p_gaps, "taxonomic_coverage_by_order.csv"))) {
  order_coverage <- read_csv(here(p_gaps, "taxonomic_coverage_by_order.csv"), show_col_types = FALSE) |>
    filter(!is.na(order), order != "", n_ref_total >= 5)
  
  # Top 20 orders by reference count
  top_orders_cov <- order_coverage |>
    arrange(desc(n_ref_total)) |>
    slice_head(n = 20)
  
  ggplot(top_orders_cov, aes(x = fct_reorder(order, pct_coverage), 
                              y = pct_coverage, fill = pct_coverage)) +
    geom_col() +
    geom_text(aes(label = paste0(round(pct_coverage, 0), "%")), 
              hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_fill_viridis_c(option = "plasma", begin = 0.1, end = 0.9) +
    labs(
      title = "GBIF Coverage by Taxonomic Order",
      subtitle = "Top 20 orders by number of reference taxa",
      x = NULL,
      y = "Coverage (%)"
    ) +
    theme(legend.position = "none") +
    ylim(0, 110)
}
```

## 3.2 Coverage by Family

```{r family-coverage, fig.height=10}
if (file.exists(here(p_gaps, "taxonomic_coverage_by_family.csv"))) {
  family_coverage <- read_csv(here(p_gaps, "taxonomic_coverage_by_family.csv"), show_col_types = FALSE) |>
    filter(!is.na(family), family != "", n_ref_total >= 5)
  
  # Top 25 families by reference count
  top_families <- family_coverage |>
    arrange(desc(n_ref_total)) |>
    slice_head(n = 25)
  
  ggplot(top_families, aes(x = fct_reorder(family, pct_coverage), 
                            y = pct_coverage, fill = pct_coverage)) +
    geom_col() +
    geom_text(aes(label = paste0(round(pct_coverage, 0), "%")), 
              hjust = -0.1, size = 3) +
    coord_flip() +
    scale_fill_viridis_c(option = "mako", begin = 0.1, end = 0.9) +
    labs(
      title = "GBIF Coverage by Family",
      subtitle = "Top 25 families by number of reference taxa",
      x = NULL,
      y = "Coverage (%)"
    ) +
    theme(legend.position = "none") +
    ylim(0, 110)
}
```

---

# 4. Data Mobilization by Order (5-Year Periods)

This section examines how data mobilization has changed over time for different taxonomic orders, grouped into 5-year periods.

```{r load-order-time}
order_path <- here(p_derived, "order_time_summary_10km.csv")

if (file.exists(order_path)) {
  order_time <- fread(order_path) |>
    filter(basisofrecord == "all") |>
    mutate(
      yearmonth_chr = str_trim(as.character(yearmonth)),
      year = as.integer(str_sub(yearmonth_chr, 1, 4))
    ) |>
    filter(year >= 1970, year <= current_year, !is.na(order), order != "")
  
  # Create 5-year bins
  order_5yr <- order_time |>
    mutate(
      period_start = floor(year / 5) * 5,
      period = paste0(period_start, "-", period_start + 4)
    ) |>
    group_by(order, period, period_start) |>
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")
  
  # Get top 12 orders by total occurrences
  top_orders <- order_time |>
    group_by(order) |>
    summarise(total = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(total)) |>
    slice_head(n = 12) |>
    pull(order)
  
  order_5yr_top <- order_5yr |>
    filter(order %in% top_orders)
  
  has_order_data <- TRUE
} else {
  cat("Order-level time summary not found.\n")
  has_order_data <- FALSE
}
```

## 4.1 Mobilization Histograms by Order

```{r order-histogram, fig.height=12, fig.width=14}
if (has_order_data) {
  ggplot(order_5yr_top, aes(x = period, y = occurrences, fill = order)) +
    geom_col() +
    facet_wrap(~order, scales = "free_y", ncol = 3) +
    scale_fill_viridis_d(option = "turbo") +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Data Mobilization by Taxonomic Order",
      subtitle = "Occurrence records in 5-year periods (Top 12 orders)",
      x = "5-Year Period",
      y = "Occurrences"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      legend.position = "none",
      strip.text = element_text(face = "bold", size = 11)
    )
}
```

## 4.2 Stacked Area Chart - All Orders Combined

```{r order-stacked, fig.height=8}
if (has_order_data) {
  ggplot(order_5yr_top, aes(x = period_start, y = occurrences, fill = order)) +
    geom_area(position = "stack", alpha = 0.8) +
    scale_fill_viridis_d(option = "turbo") +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Cumulative Data Mobilization by Order",
      subtitle = "Stacked area chart showing relative contributions",
      x = "Period Start Year",
      y = "Total Occurrences",
      fill = "Order"
    ) +
    theme(legend.position = "right")
}
```

## 4.3 Proportional Change Over Time

```{r order-proportional, fig.height=8}
if (has_order_data) {
  order_prop <- order_5yr_top |>
    group_by(period, period_start) |>
    mutate(total_period = sum(occurrences),
           proportion = 100 * occurrences / total_period) |>
    ungroup()
  
  ggplot(order_prop, aes(x = period_start, y = proportion, fill = order)) +
    geom_area(position = "stack") +
    scale_fill_viridis_d(option = "turbo") +
    labs(
      title = "Proportional Contribution by Order Over Time",
      subtitle = "How the share of data from each order has changed",
      x = "Period Start Year",
      y = "Proportion (%)",
      fill = "Order"
    )
}
```

## 4.4 Recent vs Historical Mobilization Change

```{r order-change, fig.height=8}
if (has_order_data) {
  recent_cutoff <- current_year - 10
  historical_cutoff <- current_year - 20
  
  order_change <- order_time |>
    filter(order %in% top_orders) |>
    mutate(
      era = case_when(
        year >= recent_cutoff ~ "Recent",
        year >= historical_cutoff ~ "Historical",
        TRUE ~ NA_character_
      )
    ) |>
    filter(!is.na(era)) |>
    group_by(order, era) |>
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = era, values_from = occurrences) |>
    mutate(
      pct_change = 100 * (Recent - Historical) / Historical,
      direction = ifelse(pct_change >= 0, "Increased", "Decreased")
    ) |>
    arrange(desc(pct_change))
  
  ggplot(order_change, aes(x = fct_reorder(order, pct_change), 
                            y = pct_change, fill = direction)) +
    geom_col() +
    geom_text(aes(label = paste0(ifelse(pct_change >= 0, "+", ""), 
                                  round(pct_change, 0), "%")),
              hjust = ifelse(order_change$pct_change >= 0, -0.1, 1.1), 
              size = 4) +
    coord_flip() +
    scale_fill_viridis_d(option = "viridis", begin = 0.3, end = 0.7) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    labs(
      title = "Change in Data Mobilization by Order",
      subtitle = glue("Comparing {recent_cutoff}-{current_year} vs {historical_cutoff}-{recent_cutoff - 1}"),
      x = NULL,
      y = "% Change in Occurrences",
      fill = NULL
    )
}
```

## 4.5 Mobilization Summary Table

```{r order-summary-table}
if (has_order_data) {
  order_summary_table <- order_change |>
    left_join(
      order_time |>
        group_by(order) |>
        summarise(total_all_time = sum(occurrences, na.rm = TRUE)),
      by = "order"
    ) |>
    arrange(desc(total_all_time)) |>
    select(order, Historical, Recent, pct_change, total_all_time) |>
    mutate(
      across(c(Historical, Recent, total_all_time), comma),
      pct_change = paste0(ifelse(as.numeric(gsub(",", "", pct_change)) >= 0, "+", ""), 
                          round(as.numeric(gsub(",", "", pct_change)), 1), "%")
    )
  
  order_summary_table |>
    kable(
      caption = "Data Mobilization Summary by Order",
      col.names = c("Order", "Historical (10-20yr)", "Recent (0-10yr)", 
                    "Change", "Total All Time")
    )
}
```

---

# 5. Priority Taxa for Data Collection

## 5.1 Priority Taxa Overview

```{r priority-taxa}
if (file.exists(here(p_integrated, "priority_undersampled_taxa.csv"))) {
  priority_taxa <- fread(here(p_integrated, "priority_undersampled_taxa.csv"))
  
  cat("**Priority Taxa Requiring Attention:**", comma(nrow(priority_taxa)), "\n\n")
  
  # Summary by threat status
  priority_threat <- priority_taxa |>
    filter(!is.na(threatStatus)) |>
    count(threatStatus) |>
    arrange(match(threatStatus, c("CR", "EN", "VU", "NT", "LC")))
  
  priority_threat |>
    kable(caption = "Priority Taxa by Threat Status", 
          col.names = c("Threat Status", "Count"))
}
```

```{r priority-plot, fig.height=5}
if (exists("priority_threat")) {
  ggplot(priority_threat, aes(x = threatStatus, y = n, fill = threatStatus)) +
    geom_col() +
    geom_text(aes(label = comma(n)), vjust = -0.3, size = 5) +
    scale_fill_manual(values = threat_colors) +
    labs(
      title = "Priority Taxa by IUCN Threat Status",
      subtitle = "Taxa requiring additional data collection or verification",
      x = "Threat Status",
      y = "Number of Taxa"
    ) +
    theme(legend.position = "none")
}
```

## 5.2 Highest Priority Species (CR & EN)

```{r priority-cr-en}
if (exists("priority_taxa")) {
  priority_taxa |>
    filter(threatStatus %in% c("CR", "EN")) |>
    arrange(threatStatus, scientificName) |>
    select(scientificName, threatStatus, taxonRank, family, order) |>
    slice_head(n = 30) |>
    kable(caption = "Top 30 Critically Endangered and Endangered Priority Taxa")
}
```

## 5.3 Priority Taxa by Family

```{r priority-by-family, fig.height=8}
if (exists("priority_taxa")) {
  priority_by_family <- priority_taxa |>
    filter(!is.na(family), family != "") |>
    count(family, sort = TRUE) |>
    slice_head(n = 20)
  
  ggplot(priority_by_family, aes(x = fct_reorder(family, n), y = n, fill = n)) +
    geom_col() +
    geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
    coord_flip() +
    scale_fill_viridis_c(option = "rocket", direction = -1, begin = 0.2, end = 0.8) +
    labs(
      title = "Priority Taxa by Family",
      subtitle = "Families with most taxa requiring attention",
      x = NULL,
      y = "Number of Priority Taxa"
    ) +
    theme(legend.position = "none")
}
```

---

# 6. Key Findings

```{r summary}
cat("## Taxonomic Gap Summary\n\n")

# Overall coverage
if (exists("dashboard")) {
  cat("**Overall Coverage:**\n")
  cat("- Reference taxa:", comma(dashboard$taxa_in_reference), "\n")
  cat("- Found in GBIF:", comma(dashboard$taxa_in_gbif), 
      "(", dashboard$taxa_pct_coverage, "%)\n")
  cat("- Missing taxa:", comma(dashboard$taxa_in_reference - dashboard$taxa_in_gbif), "\n\n")
}

# Threatened species
if (exists("missing_threatened")) {
  cat("**Threatened Species Gaps:**\n")
  cat("- Missing CR taxa:", sum(missing_threatened$threatStatus == "CR"), "\n")
  cat("- Missing EN taxa:", sum(missing_threatened$threatStatus == "EN"), "\n")
  cat("- Missing VU taxa:", sum(missing_threatened$threatStatus == "VU"), "\n\n")
}

# Mobilization trends
if (has_order_data) {
  cat("**Data Mobilization Trends:**\n")
  increased <- sum(order_change$pct_change > 0)
  decreased <- sum(order_change$pct_change < 0)
  cat("- Orders with increased recent mobilization:", increased, "\n")
  cat("- Orders with decreased recent mobilization:", decreased, "\n")
  
  top_increase <- order_change |> slice_head(n = 1)
  top_decrease <- order_change |> slice_tail(n = 1)
  cat("- Largest increase:", top_increase$order, 
      "(+", round(top_increase$pct_change, 0), "%)\n")
  if (top_decrease$pct_change < 0) {
    cat("- Largest decrease:", top_decrease$order, 
        "(", round(top_decrease$pct_change, 0), "%)\n")
  }
  cat("\n")
}

# Priority taxa
if (exists("priority_taxa")) {
  cat("**Priority Actions:**\n")
  cat("- Total priority taxa:", comma(nrow(priority_taxa)), "\n")
  cr_en <- sum(priority_taxa$threatStatus %in% c("CR", "EN"), na.rm = TRUE)
  cat("- High priority (CR/EN):", comma(cr_en), "\n\n")
}

cat("**Recommendations:**\n")
cat("1. Focus on CR and EN species with no or very few GBIF records\n")
cat("2. Target underrepresented families for systematic surveys\n")
cat("3. Investigate causes of declining mobilization in affected orders\n")
cat("4. Leverage museum collections for historical specimen digitization\n")
cat("5. Coordinate with taxonomic experts for difficult groups\n")
```

---

# Session Info

```{r session}
sessionInfo()
```
