---
title: "GBIF Sweden Biodiversity Data Gaps"
subtitle: "Integrated Assessment: Spatial, Temporal, and Taxonomic Dimensions"
author: "GBIF Sweden"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    code_folding: hide
    fig_width: 14
    fig_height: 9
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 14,
  fig.height = 9,
  dpi = 300
)

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(viridis)
library(patchwork)
library(sf)
library(knitr)
library(data.table)
library(gt)
library(glue)
library(lubridate)

# Professional theme with viridis colors
theme_report <- theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 14, color = "grey40"),
    plot.caption = element_text(size = 11, color = "grey50", hjust = 1),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 13, face = "bold")
  )

theme_set(theme_report)

# Paths
p_gaps <- here("data_proc", "gaps")
p_tables <- here("output", "tables")
p_integrated <- here("output", "tables", "integrated")
p_data_proc <- here("data_proc")
p_derived <- here("data_proc", "derived")

# Current year
current_year <- year(Sys.Date())

# Threat status colors
threat_colors <- c("CR" = "#D32F2F", "EN" = "#E64A19", "VU" = "#F57C00", 
                   "NT" = "#FBC02D", "LC" = "#8BC34A")
```

# Executive Summary

This report presents a comprehensive assessment of biodiversity data gaps in GBIF Sweden's occurrence database, integrating spatial, temporal, and taxonomic dimensions to inform strategic priorities for data collection and mobilization.

```{r load-dashboard}
dashboard <- read_csv(here(p_tables, "dashboard_summary.csv"), show_col_types = FALSE)
```

```{r executive-summary}
exec_summary <- tibble(
  Dimension = c(
    "Spatial Coverage (10km)", 
    "Spatial Coverage (50km)", 
    "Temporal Range",
    "Data Recency", 
    "Taxonomic Coverage", 
    "Priority Taxa"
  ),
  `Current Status` = c(
    glue("{comma(dashboard$cells_10km_with_data)} / {comma(dashboard$cells_10km_total)} cells ({dashboard$cells_10km_pct_coverage}%)"),
    glue("{comma(dashboard$cells_50km_with_data)} / {comma(dashboard$cells_50km_total)} cells ({dashboard$cells_50km_pct_coverage}%)"),
    glue("{dashboard$year_range} years ({dashboard$year_min}–{dashboard$year_max})"),
    glue("{dashboard$median_staleness_months_10km} months median; {dashboard$pct_stale_5y_10km}% cells >5yr stale"),
    glue("{comma(dashboard$taxa_in_gbif)} / {comma(dashboard$taxa_in_reference)} taxa ({dashboard$taxa_pct_coverage}%)"),
    glue("{comma(dashboard$n_priority_taxa)} taxa requiring attention")
  ),
  Assessment = c(
    ifelse(dashboard$cells_10km_pct_coverage >= 90, "✅ Excellent", 
           ifelse(dashboard$cells_10km_pct_coverage >= 75, "✅ Good", "⚠️ Needs Attention")),
    ifelse(dashboard$cells_50km_pct_coverage >= 90, "✅ Excellent", 
           ifelse(dashboard$cells_50km_pct_coverage >= 75, "✅ Good", "⚠️ Needs Attention")),
    "✅ Good",
    ifelse(dashboard$median_staleness_months_10km <= 24, "✅ Recent",
           ifelse(dashboard$median_staleness_months_10km <= 48, "⚠️ Fair", "❌ Needs Update")),
    ifelse(dashboard$taxa_pct_coverage >= 75, "✅ Good",
           ifelse(dashboard$taxa_pct_coverage >= 50, "⚠️ Fair", "❌ Low")),
    ifelse(dashboard$n_priority_taxa < 500, "✅ Manageable", "⚠️ Action Required")
  )
)

exec_summary |>
  gt() |>
  tab_header(
    title = md("**GBIF Sweden Data Gap Assessment**"),
    subtitle = glue("Analysis Date: {dashboard$analysis_date}")
  ) |>
  cols_label(
    Dimension = md("**Dimension**"),
    `Current Status` = md("**Current Status**"),
    Assessment = md("**Assessment**")
  ) |>
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 20,
    heading.subtitle.font.size = 14
  )
```

```{r summary-chart, fig.height=5}
tibble(
  dimension = c("10km Spatial", "50km Spatial", "Taxonomic"),
  coverage = c(dashboard$cells_10km_pct_coverage, 
               dashboard$cells_50km_pct_coverage, 
               dashboard$taxa_pct_coverage)
) |>
  ggplot(aes(x = dimension, y = coverage, fill = dimension)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(coverage, 1), "%")), 
            vjust = -0.5, size = 7, fontface = "bold") +
  scale_fill_viridis_d(option = "viridis", begin = 0.2, end = 0.8) +
  ylim(0, 110) +
  labs(
    title = "Coverage Across Key Dimensions",
    x = NULL, 
    y = "Coverage (%)"
  ) +
  theme(legend.position = "none")
```

---

# 1. Spatial Gaps

## 1.1 Geographic Coverage

```{r spatial-load}
comp_grids <- read_csv(here(p_tables, "comparison_grid_resolutions.csv"), show_col_types = FALSE)
spatial_overview <- read_csv(here(p_tables, "overview_spatial_gap_rates.csv"), show_col_types = FALSE)
```

```{r spatial-overview, fig.height=5}
p1 <- comp_grids |>
  ggplot(aes(x = grid_resolution, y = pct_coverage, fill = grid_resolution)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(round(pct_coverage, 1), "%")),
            vjust = -0.5, size = 6, fontface = "bold") +
  scale_fill_viridis_d(option = "viridis", begin = 0.3, end = 0.7) +
  ylim(0, 110) +
  labs(title = "Coverage %", y = "Coverage (%)", x = NULL) +
  theme(legend.position = "none")

p2 <- comp_grids |>
  select(grid_resolution, n_cells_total, n_cells_with_data) |>
  pivot_longer(-grid_resolution, names_to = "type", values_to = "count") |>
  mutate(type = recode(type, n_cells_total = "Total", n_cells_with_data = "With Data")) |>
  ggplot(aes(x = grid_resolution, y = count, fill = type)) +
  geom_col(position = "dodge", width = 0.5) +
  geom_text(aes(label = comma(count)), 
            position = position_dodge(width = 0.5), vjust = -0.5, size = 5) +
  scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.7) +
  scale_y_continuous(labels = comma) +
  labs(title = "Cell Counts", y = "Cells", x = NULL, fill = NULL)

p1 + p2 + plot_annotation(title = "Spatial Coverage Summary")
```

## 1.2 Geographic Distribution

```{r spatial-map, fig.height=12, fig.width=12}
grid_10km <- st_read(here(p_data_proc, "grids_10km.gpkg"), quiet = TRUE)
spatial_gaps <- fread(here(p_gaps, "spatial_gaps_10km.csv"))

# Find cellcode field
cellcode_candidates <- names(grid_10km)[grepl("cellcode|eeacell", tolower(names(grid_10km)))]
cellcode_field <- if(length(cellcode_candidates) > 0) cellcode_candidates[1] else names(grid_10km)[grepl("code", tolower(names(grid_10km)))][1]

grid_10km <- grid_10km |> mutate(eeacellcode = as.character(.data[[cellcode_field]]))

gaps_all <- spatial_gaps |>
  filter(basisofrecord == "all") |>
  select(eeacellcode, occurrences)

grid_map <- grid_10km |>
  left_join(gaps_all, by = "eeacellcode") |>
  mutate(log_occ = ifelse(is.na(occurrences) | occurrences == 0, NA, log10(occurrences + 1)))

ggplot(grid_map) +
  geom_sf(aes(fill = log_occ), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "log10(occurrences)",
    na.value = "grey85",
    guide = guide_colorbar(barwidth = 20, barheight = 0.8)
  ) +
  labs(
    title = "Geographic Distribution of Occurrence Data",
    subtitle = "10km EEA Grid - Sweden",
    caption = "Grey cells = zero coverage"
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40")
  )
```

---

# 2. Temporal Gaps

## 2.1 Historical Trends

```{r temporal-load}
time10 <- read_csv(here(p_derived, "time_summary_10km.csv"), show_col_types = FALSE) |>
  filter(basisofrecord == "all") |>
  mutate(
    year = as.integer(str_sub(as.character(yearmonth), 1, 4)),
    month = as.integer(str_sub(as.character(yearmonth), 6, 7))
  )

annual <- time10 |>
  group_by(year) |>
  summarise(occurrences = sum(occurrences, na.rm = TRUE)) |>
  filter(year >= 1900, year <= current_year)
```

```{r temporal-trend, fig.height=6}
ggplot(annual, aes(x = year, y = occurrences)) +
  geom_area(fill = viridis(1, option = "viridis", begin = 0.4), alpha = 0.3) +
  geom_line(color = viridis(1, option = "viridis", begin = 0.4), linewidth = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Historical Trend in Biodiversity Data Collection",
    subtitle = "Annual occurrence records from all evidence types",
    x = "Year",
    y = "Total Occurrences"
  )
```

## 2.2 Seasonal Patterns

```{r seasonal-heatmap, fig.height=8}
time10 |>
  filter(year >= 1970) |>
  group_by(year, month) |>
  summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = factor(month), y = year, fill = log10(occurrences + 1))) +
  geom_tile() +
  scale_fill_viridis(option = "inferno", name = "log10(occ+1)") +
  scale_x_discrete(labels = month.abb) +
  labs(
    title = "Seasonal Sampling Intensity (1970-Present)",
    subtitle = "Year × Month heatmap",
    x = "Month",
    y = "Year"
  )
```

## 2.3 Data Recency

```{r recency-map, fig.height=10, fig.width=12}
if (file.exists(here(p_gaps, "temporal_cell_recency_10km.csv"))) {
  cell_recency <- fread(here(p_gaps, "temporal_cell_recency_10km.csv"))
  
  recency_all <- cell_recency |>
    filter(basisofrecord == "all") |>
    select(eeacellcode, staleness_months)
  
  grid_stale <- grid_10km |>
    left_join(recency_all, by = "eeacellcode") |>
    mutate(
      staleness_cat = case_when(
        is.na(staleness_months) ~ "No Data",
        staleness_months <= 12 ~ "< 1 year",
        staleness_months <= 36 ~ "1-3 years",
        staleness_months <= 60 ~ "3-5 years",
        staleness_months <= 120 ~ "5-10 years",
        TRUE ~ "> 10 years"
      ),
      staleness_cat = factor(staleness_cat, levels = c(
        "< 1 year", "1-3 years", "3-5 years", "5-10 years", "> 10 years", "No Data"
      ))
    )
  
  ggplot(grid_stale) +
    geom_sf(aes(fill = staleness_cat), color = "white", linewidth = 0.02) +
    scale_fill_viridis_d(option = "turbo", direction = -1, na.value = "grey80") +
    labs(
      title = "Data Recency Across Sweden",
      subtitle = "Time since last occurrence record per cell",
      fill = "Staleness"
    ) +
    theme_void() +
    theme(
      legend.position = "right",
      legend.key.height = unit(0.8, "cm"),
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40")
    )
}
```

---

# 3. Taxonomic Gaps

## 3.1 Coverage by Threat Status

```{r threat-coverage, fig.height=6}
if (file.exists(here(p_gaps, "taxonomic_coverage_by_threat.csv"))) {
  threat_coverage <- read_csv(here(p_gaps, "taxonomic_coverage_by_threat.csv"), show_col_types = FALSE) |>
    filter(threatStatus %in% c("CR", "EN", "VU", "NT", "LC")) |>
    mutate(threatStatus = factor(threatStatus, levels = c("CR", "EN", "VU", "NT", "LC")))
  
  p1 <- ggplot(threat_coverage, aes(x = threatStatus, y = n_ref_total, fill = threatStatus)) +
    geom_col() +
    geom_text(aes(label = comma(n_ref_total)), vjust = -0.3, size = 5) +
    scale_fill_manual(values = threat_colors) +
    labs(title = "Reference Taxa", y = "Count", x = NULL) +
    theme(legend.position = "none")
  
  p2 <- ggplot(threat_coverage, aes(x = threatStatus, y = pct_coverage, fill = threatStatus)) +
    geom_col() +
    geom_text(aes(label = paste0(round(pct_coverage, 1), "%")), vjust = -0.3, size = 5) +
    scale_fill_manual(values = threat_colors) +
    labs(title = "GBIF Coverage", y = "Coverage (%)", x = NULL) +
    theme(legend.position = "none") +
    ylim(0, 110)
  
  p1 + p2 + plot_annotation(
    title = "Threatened Species Coverage",
    subtitle = "IUCN Red List categories: CR = Critically Endangered, EN = Endangered, VU = Vulnerable, NT = Near Threatened"
  )
}
```

## 3.2 Data Mobilization by Order (5-Year Periods)

```{r order-mobilization, fig.height=10, fig.width=14}
order_path <- here(p_derived, "order_time_summary_10km.csv")

if (file.exists(order_path)) {
  order_time <- fread(order_path) |>
    filter(basisofrecord == "all") |>
    mutate(year = as.integer(str_sub(as.character(yearmonth), 1, 4))) |>
    filter(year >= 1970, year <= current_year, !is.na(order), order != "")
  
  # 5-year bins
  order_5yr <- order_time |>
    mutate(
      period_start = floor(year / 5) * 5,
      period = paste0(period_start, "-", period_start + 4)
    ) |>
    group_by(order, period, period_start) |>
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop")
  
  # Top 9 orders
  top_orders <- order_time |>
    group_by(order) |>
    summarise(total = sum(occurrences), .groups = "drop") |>
    arrange(desc(total)) |>
    slice_head(n = 9) |>
    pull(order)
  
  order_5yr |>
    filter(order %in% top_orders) |>
    ggplot(aes(x = period, y = occurrences, fill = order)) +
    geom_col() +
    facet_wrap(~order, scales = "free_y", ncol = 3) +
    scale_fill_viridis_d(option = "turbo") +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Data Mobilization Trends by Taxonomic Order",
      subtitle = "Occurrence records in 5-year periods (Top 9 orders)",
      x = "5-Year Period",
      y = "Occurrences"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      legend.position = "none",
      strip.text = element_text(face = "bold")
    )
}
```

---

# 4. Priority Recommendations

## 4.1 Geographic Priorities

```{r priority-cells}
priority_zero <- if (file.exists(here(p_integrated, "priority_zero_coverage_cells.csv"))) {
  fread(here(p_integrated, "priority_zero_coverage_cells.csv"))
} else data.frame()

priority_stale <- if (file.exists(here(p_integrated, "priority_stale_cells.csv"))) {
  fread(here(p_integrated, "priority_stale_cells.csv"))
} else data.frame()
```

```{r priority-spatial-summary}
cat("### Spatial Priority Summary\n\n")

cat("**Zero Coverage Cells:**", comma(nrow(priority_zero)), "\n")
cat("- These cells have no occurrence records from any evidence type\n")
cat("- Recommendation: Target for systematic biodiversity surveys\n\n")

cat("**Stale Cells (>5 years):**", comma(nrow(priority_stale)), "\n")
if (nrow(priority_stale) > 0) {
  cat("- Median staleness:", round(median(priority_stale$staleness_months) / 12, 1), "years\n")
}
cat("- Recommendation: Re-survey to update biodiversity status\n")
```

## 4.2 Taxonomic Priorities

```{r priority-taxa}
if (file.exists(here(p_integrated, "priority_undersampled_taxa.csv"))) {
  priority_taxa <- fread(here(p_integrated, "priority_undersampled_taxa.csv"))
  
  cat("### Taxonomic Priority Summary\n\n")
  cat("**Total Priority Taxa:**", comma(nrow(priority_taxa)), "\n\n")
  
  priority_threat_summary <- priority_taxa |>
    filter(threatStatus %in% c("CR", "EN", "VU", "NT")) |>
    count(threatStatus) |>
    arrange(match(threatStatus, c("CR", "EN", "VU", "NT")))
  
  for (i in 1:nrow(priority_threat_summary)) {
    cat("- **", priority_threat_summary$threatStatus[i], ":**", 
        comma(priority_threat_summary$n[i]), "taxa\n")
  }
  
  cat("\n**Recommendation:** Develop targeted monitoring programs for red-listed taxa,\n")
  cat("prioritizing CR and EN species.\n")
}
```

## 4.3 Priority Action Table

```{r action-table}
actions <- tibble(
  Priority = c("High", "High", "Medium", "Medium", "Low"),
  Category = c("Spatial", "Taxonomic", "Temporal", "Taxonomic", "Spatial"),
  Action = c(
    "Survey zero-coverage grid cells",
    "Monitor CR and EN species lacking recent records",
    "Re-survey cells with >5 years since last observation",
    "Digitize museum specimens for underrepresented groups",
    "Expand citizen science to low-coverage regions"
  ),
  Target = c(
    glue("{comma(nrow(priority_zero))} cells"),
    glue("{sum(priority_taxa$threatStatus %in% c('CR', 'EN'), na.rm = TRUE)} taxa"),
    glue("{comma(nrow(priority_stale))} cells"),
    "Multiple families",
    "Northern Sweden"
  )
)

actions |>
  gt() |>
  tab_header(title = "Recommended Priority Actions") |>
  cols_label(
    Priority = md("**Priority**"),
    Category = md("**Category**"),
    Action = md("**Action**"),
    Target = md("**Target**")
  ) |>
  data_color(
    columns = Priority,
    colors = scales::col_factor(
      palette = c("High" = "#E63946", "Medium" = "#F4A261", "Low" = "#2A9D8F"),
      domain = c("High", "Medium", "Low")
    )
  )
```

---

# 5. Data Citations

**GBIF Occurrence Cubes - Sweden**

- 10km grid: https://doi.org/10.15468/dl.wzv3uc
- 50km grid: https://doi.org/10.15468/dl.qyp3uw

**Taxonomic Reference**

SLU Artdatabanken. The Swedish Red List 2020. https://doi.org/10.15468/jhwkpq

**Spatial Reference**

EEA Reference Grids (10km and 50km) - INSPIRE geographical grid systems

---

# 6. Methods

This analysis employed a multi-dimensional gap assessment framework:

1. **Spatial Analysis:** Coverage assessment of EEA reference grid cells at 10km and 50km resolutions
2. **Temporal Analysis:** Historical trends, seasonal patterns, and data recency assessment
3. **Taxonomic Analysis:** Comparison of GBIF occurrences against Swedish Red List reference taxonomy
4. **Integration:** Cross-dimensional priority identification for targeted action

**Software:** R 4.x with sf, dplyr, ggplot2, data.table, viridis, gt

**Analysis Date:** `r dashboard$analysis_date`

---

*For questions or more information, please contact GBIF Sweden.*

---

# Session Info

```{r session, echo=TRUE}
sessionInfo()
```
