---
title: "Taxonomic Gap Analysis"
subtitle: "Species Coverage, Reconciliation, and Sampling Bias"
author: "GBIF Gap Finder"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
  pdf_document:
    toc: true
    latex_engine: xelatex
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6, dpi = 150
)

library(here); library(dplyr); library(tidyr); library(ggplot2)
library(scales); library(knitr); library(kableExtra)

data_path <- here("shiny_app", "gap_finder", "data", "shiny_data.rds")
app_data <- readRDS(data_path)
safe_get <- function(name) if (name %in% names(app_data)) app_data[[name]] else NULL

match_summary  <- safe_get("taxonomic_match_summary")
tax_by_rank    <- safe_get("tax_by_rank")
tax_by_kingdom <- safe_get("tax_by_kingdom")
tax_by_order   <- safe_get("tax_by_order")
tax_by_family  <- safe_get("tax_by_family")
tax_by_threat  <- safe_get("tax_by_threat")
troudet_order  <- safe_get("troudet_bias_order")
troudet_family <- safe_get("troudet_bias_family")
tax_estab      <- safe_get("tax_by_establishment")
dashboard      <- safe_get("dashboard")
metadata       <- safe_get("metadata")

# Normalise column names: script 10 uses n_ref_total, script 11's grp_fn uses n_taxa
norm_tax <- function(df) {
  if (is.null(df)) return(NULL)
  if ("n_ref_total" %in% names(df) && !"n_taxa" %in% names(df))
    df <- df |> rename(n_taxa = n_ref_total)
  df
}
tax_by_rank    <- norm_tax(tax_by_rank)
tax_by_kingdom <- norm_tax(tax_by_kingdom)
tax_by_order   <- norm_tax(tax_by_order)
tax_by_family  <- norm_tax(tax_by_family)
tax_by_threat  <- norm_tax(tax_by_threat)
tax_estab      <- norm_tax(tax_estab)

pal <- list(sage = "#6b8f71", slate = "#5c7a99", sand = "#c4a882",
            coral = "#c47a6c", plum = "#8b6d8f")
theme_gaps <- theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "#666", size = 11),
        panel.grid.minor = element_blank(), legend.position = "bottom")
theme_set(theme_gaps)
```

# Overview

This report analyses how well GBIF occurrence data covers the national taxonomy backbone (**`r metadata$taxonomy_name %||% "National checklist"`**). Every species in the backbone is checked against GBIF records through a 4-tier reconciliation: direct name match, synonym resolution, infraspecific collapse, and GBIF Species API lookup.

```{r recon-stats}
if (!is.null(dashboard)) {
  stats <- tibble(
    Metric = c("Species in backbone", "Species in GBIF",
               "Matched to backbone", "Coverage", "Missing from GBIF"),
    Value = c(comma(dashboard$taxa_in_reference[1]),
              comma(dashboard$taxa_in_gbif[1]),
              comma(dashboard$taxa_in_gbif[1]),
              paste0(dashboard$taxa_pct_coverage[1], "%"),
              comma(dashboard$taxa_missing[1]))
  )
  kable(stats, align = c("l", "r")) |> kable_styling(full_width = FALSE)
}
```

# Coverage by Taxonomic Rank

```{r by-rank, fig.height=5}
if (!is.null(tax_by_rank)) {
  df <- tax_by_rank |>
    filter(n_taxa >= 10) |>
    arrange(desc(n_taxa)) |> head(10)

  ggplot(df, aes(x = reorder(taxonRank, pct_coverage), y = pct_coverage)) +
    geom_col(fill = pal$sage) +
    geom_text(aes(label = paste0(pct_coverage, "% (", comma(n_in_gbif), "/", comma(n_taxa), ")")),
              hjust = -0.05, size = 3.2) +
    coord_flip() +
    scale_y_continuous(limits = c(0, 110)) +
    labs(title = "GBIF Coverage by Taxonomic Rank", x = "", y = "Coverage (%)")
}
```

# Coverage by Kingdom

```{r by-kingdom, fig.height=5}
if (!is.null(tax_by_kingdom)) {
  ggplot(tax_by_kingdom, aes(x = reorder(kingdom, n_taxa), y = pct_coverage)) +
    geom_col(fill = pal$sage) +
    geom_text(aes(label = paste0(pct_coverage, "%")), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_y_continuous(limits = c(0, 110)) +
    labs(title = "Coverage by Kingdom", x = "", y = "Coverage (%)")
}
```

# Coverage by Threat Status

```{r by-threat}
if (!is.null(tax_by_threat) && nrow(tax_by_threat) > 0) {
  df <- tax_by_threat |> filter(threatStatus %in% c("CR", "EN", "VU", "NT", "DD"))
  kable(df |> select(Status = threatStatus, Total = n_taxa,
    `In GBIF` = n_in_gbif, Missing = n_missing, `Coverage %` = pct_coverage),
    align = c("l", rep("r", 4))) |> kable_styling(full_width = FALSE)
}
```

```{r threat-chart, fig.height=4}
if (!is.null(tax_by_threat) && nrow(tax_by_threat) > 0) {
  df <- tax_by_threat |>
    filter(threatStatus %in% c("CR", "EN", "VU", "NT", "DD")) |>
    mutate(threatStatus = factor(threatStatus, c("CR", "EN", "VU", "NT", "DD")))

  threat_colors <- c(CR = pal$coral, EN = pal$sand, VU = "#b8a060",
                     NT = pal$sage, DD = pal$slate)

  ggplot(df, aes(x = threatStatus, y = pct_coverage, fill = threatStatus)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = paste0(pct_coverage, "%")), vjust = -0.3, size = 3.5) +
    scale_fill_manual(values = threat_colors) +
    scale_y_continuous(limits = c(0, 110)) +
    labs(title = "GBIF Coverage by Threat Status", x = "Red List category", y = "Coverage (%)")
}
```

# Most Under-Sampled Orders

```{r undersampled-orders, fig.height=7}
if (!is.null(tax_by_order)) {
  df <- tax_by_order |>
    filter(n_missing > 0) |>
    arrange(desc(n_missing)) |> head(20)

  ggplot(df, aes(x = reorder(order, n_missing), y = n_missing)) +
    geom_col(aes(fill = pct_coverage < 50), show.legend = FALSE) +
    scale_fill_manual(values = c(`FALSE` = pal$sand, `TRUE` = pal$coral)) +
    geom_text(aes(label = paste0(n_missing, " (", pct_coverage, "%)")),
              hjust = -0.05, size = 3) +
    coord_flip() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
    labs(title = "Orders with Most Missing Species",
         subtitle = "Count of backbone species not found in GBIF (coverage % in parentheses)",
         x = "", y = "Species missing from GBIF")
}
```

# Troudet Sampling Bias

The Troudet-style analysis compares each taxonomic group's share of known species against its share of occurrence records. Groups with **negative bias** are under-represented in GBIF relative to their species richness.

```{r troudet, fig.height=7}
if (!is.null(troudet_order)) {
  df <- troudet_order |>
    filter(bias < 0) |>
    arrange(bias) |> head(20) |>
    mutate(gap = n_known_species - n_in_gbif)

  ggplot(df, aes(x = reorder(order, -bias), y = bias)) +
    geom_col(fill = pal$coral) +
    coord_flip() +
    scale_y_continuous(labels = label_comma()) +
    labs(title = "Most Under-Sampled Orders (Troudet Bias)",
         subtitle = "Negative bias = fewer records than expected given species richness",
         x = "", y = "Sampling bias (occurrences)")
}
```

# Establishment Means

```{r establishment, fig.height=4}
if (!is.null(tax_estab)) {
  df <- tax_estab |>
    mutate(label = case_when(
      is.na(establishmentMeans) | establishmentMeans == "" ~ "Unclassified",
      TRUE ~ tools::toTitleCase(establishmentMeans)
    )) |> arrange(desc(n_taxa))

  kable(df |> select(Category = label, Total = n_taxa,
    `In GBIF` = n_in_gbif, Missing = n_missing, `Coverage %` = pct_coverage),
    align = c("l", rep("r", 4))) |> kable_styling(full_width = FALSE)
}
```

---

*Report generated by `gbif_gap_finder` on `r Sys.time()`.*
