---
title: "GBIF Biodiversity Data Gaps"
subtitle: "Integrated Assessment: Spatial, Temporal, and Taxonomic Dimensions"
author: "gbifgaps"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    code_folding: hide
    fig_width: 14
    fig_height: 9
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 3
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}

if (!knitr::is_html_output()) {
  knitr::opts_chunk$set(echo = FALSE)
}

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 14,
  fig.height = 9,
  dpi = 150
)

library(here)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(viridis)
library(patchwork)
library(sf)
library(knitr)
library(data.table)
library(glue)
library(lubridate)

# Professional theme
theme_report <- theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 14, color = "grey40"),
    plot.caption = element_text(size = 11, color = "grey50", hjust = 1),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 13, face = "bold")
  )

theme_set(theme_report)

# Paths
p_gaps <- here("data_proc", "gaps")
p_tables <- here("output", "tables")
p_integrated <- here("output", "tables", "integrated")
p_data_proc <- here("data_proc")
p_derived <- here("data_proc", "derived")

# Current year
current_year <- year(Sys.Date())

# Color palettes
threat_colors <- c("CR" = "#D32F2F", "EN" = "#E64A19", "VU" = "#F57C00", 
                   "NT" = "#FBC02D", "LC" = "#8BC34A", "DD" = "#78909C")

# Safe file reader - returns NULL if file doesn't exist
safe_read_csv <- function(path) {
  tryCatch({
    if (file.exists(path)) read_csv(path, show_col_types = FALSE) else NULL
  }, error = function(e) NULL)
}

safe_fread <- function(path) {
  tryCatch({
    if (file.exists(path)) fread(path) else NULL
  }, error = function(e) NULL)
}

# Safe column accessor
safe_col <- function(df, col, default = NA) {
  if (is.null(df)) return(default)
  if (col %in% names(df)) df[[col]][1] else default
}
```

# Executive Summary

This report presents a comprehensive assessment of biodiversity data gaps in the national GBIF occurrence database, integrating spatial, temporal, and taxonomic dimensions to inform strategic priorities for data collection and mobilization.

```{r load-dashboard}
dashboard <- safe_read_csv(here(p_tables, "dashboard_summary.csv"))
```

```{r executive-summary}
if (!is.null(dashboard) && nrow(dashboard) > 0) {
  # Build the summary table with safe column access
  exec_summary <- tibble(
    Dimension = c(
      "Spatial Coverage (10km)", 
      "Spatial Coverage (50km)", 
      "Temporal Range",
      "Data Recency", 
      "Taxonomic Coverage", 
      "Priority Taxa"
    ),
    `Current Status` = c(
      paste0(comma(safe_col(dashboard, "cells_10km_with_data", 0)), " / ", 
             comma(safe_col(dashboard, "cells_10km_total", 0)), " cells (",
             safe_col(dashboard, "cells_10km_pct_coverage", 0), "%)"),
      paste0(comma(safe_col(dashboard, "cells_50km_with_data", 0)), " / ", 
             comma(safe_col(dashboard, "cells_50km_total", 0)), " cells (",
             safe_col(dashboard, "cells_50km_pct_coverage", 0), "%)"),
      paste0(safe_col(dashboard, "year_range", "N/A"), " years (",
             safe_col(dashboard, "year_min", "?"), "-", safe_col(dashboard, "year_max", "?"), ")"),
      paste0(safe_col(dashboard, "median_staleness_months_10km", "N/A"), " months median; ",
             safe_col(dashboard, "pct_stale_5y_10km", "N/A"), "% cells >5yr stale"),
      paste0(comma(safe_col(dashboard, "taxa_in_gbif", 0)), " / ", 
             comma(safe_col(dashboard, "taxa_in_reference", 0)), " taxa (",
             safe_col(dashboard, "taxa_pct_coverage", 0), "%)"),
      paste0(comma(safe_col(dashboard, "n_priority_taxa", 0)), " taxa requiring attention")
    ),
    Assessment = c(
      { pct <- as.numeric(safe_col(dashboard, "cells_10km_pct_coverage", 0))
        ifelse(pct >= 90, "Excellent", ifelse(pct >= 75, "Good", "Needs Attention")) },
      { pct <- as.numeric(safe_col(dashboard, "cells_50km_pct_coverage", 0))
        ifelse(pct >= 90, "Excellent", ifelse(pct >= 75, "Good", "Needs Attention")) },
      "Good",
      { months <- as.numeric(safe_col(dashboard, "median_staleness_months_10km", 999))
        ifelse(months <= 24, "Recent", ifelse(months <= 48, "Fair", "Needs Update")) },
      { pct <- as.numeric(safe_col(dashboard, "taxa_pct_coverage", 0))
        ifelse(pct >= 75, "Good", ifelse(pct >= 50, "Fair", "Low")) },
      { n <- as.numeric(safe_col(dashboard, "n_priority_taxa", 0))
        ifelse(n < 500, "Manageable", "Action Required") }
    )
  )
  
  exec_summary |> kable()
} else {
  cat("Dashboard summary not found. Please run the analysis pipeline first.\n")
}
```

```{r summary-chart, fig.height=5, fig.width=10}
if (!is.null(dashboard) && nrow(dashboard) > 0) {
  cov_10km <- as.numeric(safe_col(dashboard, "cells_10km_pct_coverage", NA))
  cov_50km <- as.numeric(safe_col(dashboard, "cells_50km_pct_coverage", NA))
  cov_taxa <- as.numeric(safe_col(dashboard, "taxa_pct_coverage", NA))
  
  # Build data frame only with available values
  dims <- c()
  covs <- c()
  if (!is.na(cov_10km)) { dims <- c(dims, "10km Spatial"); covs <- c(covs, cov_10km) }
  if (!is.na(cov_50km)) { dims <- c(dims, "50km Spatial"); covs <- c(covs, cov_50km) }
  if (!is.na(cov_taxa)) { dims <- c(dims, "Taxonomic"); covs <- c(covs, cov_taxa) }
  
  if (length(dims) > 0) {
    coverage_data <- tibble(dimension = dims, coverage = covs, gap = 100 - covs) |>
      pivot_longer(cols = c(coverage, gap), names_to = "type", values_to = "pct") |>
      mutate(type = factor(type, levels = c("gap", "coverage"), labels = c("Gap", "Covered")))
    
    ggplot(coverage_data, aes(x = dimension, y = pct, fill = type)) +
      geom_col(width = 0.6, alpha = 0.9) +
      geom_text(data = coverage_data |> filter(type == "Covered"),
                aes(label = paste0(round(pct, 1), "%")), 
                position = position_stack(vjust = 0.5),
                size = 6, fontface = "bold", color = "white") +
      scale_fill_manual(values = c("Gap" = "#E76F51", "Covered" = "#2A9D8F"), name = NULL) +
      coord_flip() +
      labs(title = "Coverage vs Gaps", x = NULL, y = "Percentage") +
      theme(legend.position = "bottom")
  }
}
```

---

# 1. Spatial Gaps

## 1.1 Geographic Coverage

```{r spatial-load}
comp_grids <- safe_read_csv(here(p_tables, "comparison_grid_resolutions.csv"))
```

```{r spatial-overview, fig.height=5, fig.width=12}
if (!is.null(comp_grids) && nrow(comp_grids) > 0) {
  # Check which columns exist
  has_resolution <- "grid_resolution" %in% names(comp_grids)
  has_coverage <- "pct_coverage" %in% names(comp_grids)
  has_cells <- all(c("n_cells_total", "n_cells_with_data") %in% names(comp_grids))
  
  if (has_resolution && has_coverage) {
    p1 <- ggplot(comp_grids, aes(x = grid_resolution, y = pct_coverage, fill = grid_resolution)) +
      geom_col(width = 0.5, alpha = 0.9) +
      geom_text(aes(label = paste0(round(pct_coverage, 1), "%")),
                vjust = -0.5, size = 6, fontface = "bold") +
      scale_fill_manual(values = c("10km" = "#2A9D8F", "50km" = "#457B9D")) +
      ylim(0, 110) +
      labs(title = "Coverage %", y = "Coverage (%)", x = NULL) +
      theme(legend.position = "none")
    
    if (has_cells) {
      p2 <- comp_grids |>
        mutate(n_cells_empty = n_cells_total - n_cells_with_data) |>
        select(grid_resolution, n_cells_with_data, n_cells_empty) |>
        pivot_longer(-grid_resolution, names_to = "type", values_to = "count") |>
        mutate(type = factor(type, levels = c("n_cells_empty", "n_cells_with_data"),
                             labels = c("Empty", "With Data"))) |>
        ggplot(aes(x = grid_resolution, y = count, fill = type)) +
        geom_col(width = 0.5, alpha = 0.9) +
        scale_fill_manual(values = c("With Data" = "#2A9D8F", "Empty" = "#E76F51")) +
        scale_y_continuous(labels = comma) +
        labs(title = "Cell Counts", y = "Cells", x = NULL, fill = NULL)
      
      p1 + p2 + plot_annotation(title = "Spatial Coverage Summary")
    } else {
      print(p1)
    }
  } else {
    cat("Required columns not found in comparison_grid_resolutions.csv\n")
    cat("Available columns:", paste(names(comp_grids), collapse = ", "), "\n")
  }
} else {
  cat("Spatial comparison data not found.\n")
}
```

## 1.2 Geographic Distribution

```{r spatial-map, fig.height=12, fig.width=10}
grid_path <- here(p_data_proc, "grids_10km.gpkg")
gaps_path <- here(p_gaps, "spatial_gaps_10km.csv")

grid_10km <- NULL

if (file.exists(grid_path) && file.exists(gaps_path)) {
  grid_10km <- tryCatch(st_read(grid_path, quiet = TRUE), error = function(e) NULL)
  spatial_gaps <- safe_fread(gaps_path)
  
  if (!is.null(grid_10km) && !is.null(spatial_gaps)) {
    # Find cellcode field
    cellcode_field <- names(grid_10km)[grepl("cellcode|eeacell", tolower(names(grid_10km)))][1]
    if (is.na(cellcode_field)) {
      cellcode_field <- names(grid_10km)[grepl("code", tolower(names(grid_10km)))][1]
    }
    
    if (!is.na(cellcode_field)) {
      grid_10km <- grid_10km |> mutate(eeacellcode = as.character(.data[[cellcode_field]]))
      
      gaps_all <- spatial_gaps |>
        filter(basisofrecord == "all") |>
        select(eeacellcode, occurrences)
      
      grid_map <- grid_10km |>
        left_join(gaps_all, by = "eeacellcode") |>
        mutate(
          occ_cat = case_when(
            is.na(occurrences) | occurrences == 0 ~ "No data",
            occurrences < 100 ~ "1-99",
            occurrences < 1000 ~ "100-999",
            occurrences < 10000 ~ "1,000-9,999",
            occurrences < 100000 ~ "10,000-99,999",
            TRUE ~ "100,000+"
          ),
          occ_cat = factor(occ_cat, levels = c("No data", "1-99", "100-999", 
                                                "1,000-9,999", "10,000-99,999", "100,000+"))
        )
      
      print(ggplot(grid_map) +
        geom_sf(aes(fill = occ_cat), color = "white", linewidth = 0.05) +
        scale_fill_manual(
          values = c("No data" = "grey85", "1-99" = "#FDE725", "100-999" = "#5DC863",
                     "1,000-9,999" = "#21908C", "10,000-99,999" = "#3B528B", "100,000+" = "#440154"),
          name = "Occurrences"
        ) +
        labs(title = "Geographic Distribution of Occurrence Data",
             subtitle = "10km EEA Grid") +
        theme_void() +
        theme(legend.position = "right", legend.key.height = unit(0.8, "cm"),
              plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
              plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40")))
    }
  }
} else {
  cat("Spatial data files not found.\n")
}
```

---

# 2. Temporal Gaps

## 2.1 Historical Trends

```{r temporal-load}
time_path <- here(p_derived, "time_summary_10km.csv")
time10 <- safe_fread(time_path)
annual <- NULL

if (!is.null(time10) && nrow(time10) > 0 && "basisofrecord" %in% names(time10)) {
  time10 <- time10 |>
    filter(basisofrecord == "all") |>
    mutate(
      year = as.integer(str_sub(as.character(yearmonth), 1, 4)),
      month = as.integer(str_sub(as.character(yearmonth), 6, 7))
    )
  
  annual <- time10 |>
    group_by(year) |>
    summarise(occurrences = sum(occurrences, na.rm = TRUE), .groups = "drop") |>
    filter(year >= 1900, year <= current_year)
}
```

```{r temporal-trend, fig.height=6, fig.width=14}
if (!is.null(annual) && nrow(annual) > 0) {
  ggplot(annual, aes(x = year, y = occurrences)) +
    geom_area(fill = "#2A9D8F", alpha = 0.3) +
    geom_line(color = "#2A9D8F", linewidth = 1.2) +
    geom_vline(xintercept = seq(1950, 2020, 10), linetype = "dotted", alpha = 0.3) +
    scale_y_continuous(labels = comma) +
    scale_x_continuous(breaks = seq(1900, 2025, 25)) +
    labs(title = "Historical Trend in Biodiversity Data Collection",
         subtitle = "Annual occurrence records from all evidence types",
         x = "Year", y = "Total Occurrences")
} else {
  cat("Temporal data not found or empty.\n")
}
```

## 2.2 Data Recency

```{r recency-map, fig.height=12, fig.width=10}
recency_path <- here(p_gaps, "temporal_cell_recency_10km.csv")
if (!file.exists(recency_path)) {
  recency_path <- here(p_gaps, "cell_recency_10km.csv")
}

if (file.exists(recency_path) && !is.null(grid_10km)) {
  cell_recency <- safe_fread(recency_path)
  
  if (!is.null(cell_recency) && "staleness_months" %in% names(cell_recency)) {
    recency_all <- cell_recency |>
      filter(basisofrecord == "all") |>
      select(eeacellcode, staleness_months)
    
    grid_stale <- grid_10km |>
      left_join(recency_all, by = "eeacellcode") |>
      mutate(
        staleness_years = staleness_months / 12,
        staleness_cat = case_when(
          is.na(staleness_months) ~ "No Data",
          staleness_years <= 1 ~ "< 1 year",
          staleness_years <= 3 ~ "1-3 years",
          staleness_years <= 5 ~ "3-5 years",
          staleness_years <= 10 ~ "5-10 years",
          TRUE ~ "> 10 years"
        ),
        staleness_cat = factor(staleness_cat, levels = c(
          "< 1 year", "1-3 years", "3-5 years", "5-10 years", "> 10 years", "No Data"
        ))
      )
    
    ggplot(grid_stale) +
      geom_sf(aes(fill = staleness_cat), color = "white", linewidth = 0.02) +
      scale_fill_viridis_d(option = "rocket", direction = -1, na.value = "grey80",
                           name = "Time Since\nLast Record") +
      labs(title = "Data Staleness Across Study Area",
           subtitle = "Time since last occurrence record per 10km cell") +
      theme_void() +
      theme(legend.position = "right", legend.key.height = unit(0.7, "cm"),
            plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
            plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40"))
  }
} else {
  cat("Recency data not found.\n")
}
```

---

# 3. Taxonomic Gaps

## 3.1 Coverage by Kingdom

```{r taxonomic-load}
match_summary <- safe_fread(here(p_gaps, "taxonomic_match_summary.csv"))
```

```{r kingdom-coverage, fig.height=7, fig.width=12}
if (!is.null(match_summary) && nrow(match_summary) > 0 && 
    all(c("kingdom", "matched_any") %in% names(match_summary))) {
  
  kingdom_coverage <- match_summary |>
    filter(!is.na(kingdom), kingdom != "") |>
    group_by(kingdom) |>
    summarise(
      n_dyntaxa = n(),
      n_gbif = sum(matched_any, na.rm = TRUE),
      n_missing = n_dyntaxa - n_gbif,
      pct_coverage = round(100 * n_gbif / n_dyntaxa, 1),
      .groups = "drop"
    ) |>
    arrange(desc(n_dyntaxa))
  
  kingdom_long <- kingdom_coverage |>
    select(kingdom, n_gbif, n_missing) |>
    pivot_longer(cols = c(n_gbif, n_missing), names_to = "status", values_to = "count") |>
    mutate(status = factor(status, levels = c("n_missing", "n_gbif"),
                           labels = c("Missing from GBIF", "In GBIF")))
  
  ggplot(kingdom_long, aes(x = fct_reorder(kingdom, count, .fun = sum), y = count, fill = status)) +
    geom_col(alpha = 0.9, width = 0.7) +
    geom_text(data = kingdom_coverage,
              aes(x = kingdom, y = n_dyntaxa, label = paste0(pct_coverage, "%"), fill = NULL),
              hjust = -0.2, size = 5, fontface = "bold") +
    coord_flip() +
    scale_fill_manual(values = c("In GBIF" = "#2A9D8F", "Missing from GBIF" = "#457B9D"), name = NULL) +
    scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
    labs(title = "Taxonomic Coverage by Kingdom",
         subtitle = "Species in backbone vs found in GBIF | Percentage = coverage",
         x = NULL, y = "Number of Species") +
    theme(legend.position = "bottom")
} else {
  cat("Taxonomic match summary not found or missing required columns.\n")
}
```

## 3.2 Coverage by Threat Status

```{r threat-coverage, fig.height=6, fig.width=12}
has_threat_data <- FALSE
threat_stats <- NULL

if (!is.null(match_summary) && nrow(match_summary) > 0) {
  # Find threat status column
  threat_col <- intersect(c("threatStatus", "threatStatus_redlist", "threatStatus_dyntaxa"), 
                          names(match_summary))[1]
  
  if (!is.na(threat_col)) {
    threatened_summary <- match_summary |>
      mutate(threatStatus = .data[[threat_col]]) |>
      filter(!is.na(threatStatus), threatStatus != "", 
             threatStatus %in% c("CR", "EN", "VU", "NT", "LC"))
    
    if (nrow(threatened_summary) > 0) {
      has_threat_data <- TRUE
      
      threat_stats <- threatened_summary |>
        group_by(threatStatus) |>
        summarise(
          n_dyntaxa = n(),
          n_gbif = sum(matched_any, na.rm = TRUE),
          n_missing = n_dyntaxa - n_gbif,
          pct_coverage = round(100 * n_gbif / n_dyntaxa, 1),
          .groups = "drop"
        ) |>
        mutate(threatStatus = factor(threatStatus, levels = c("CR", "EN", "VU", "NT", "LC")))
      
      threat_long <- threat_stats |>
        select(threatStatus, n_gbif, n_missing) |>
        pivot_longer(cols = c(n_gbif, n_missing), names_to = "status", values_to = "count") |>
        mutate(status = factor(status, levels = c("n_missing", "n_gbif"),
                               labels = c("Missing", "In GBIF")))
      
      p1 <- ggplot(threat_long, aes(x = threatStatus, y = count, fill = status)) +
        geom_col(alpha = 0.9, width = 0.6) +
        geom_text(data = threat_stats,
                  aes(x = threatStatus, y = n_dyntaxa, label = comma(n_dyntaxa), fill = NULL),
                  vjust = -0.3, size = 5) +
        scale_fill_manual(values = c("In GBIF" = "#2A9D8F", "Missing" = "#E76F51"), name = NULL) +
        scale_y_continuous(expand = expansion(mult = c(0, 0.12)), labels = comma) +
        labs(title = "Taxa by Threat Status", x = NULL, y = "Count") +
        theme(legend.position = "none")
      
      p2 <- ggplot(threat_stats, aes(x = threatStatus, y = pct_coverage, fill = threatStatus)) +
        geom_col(alpha = 0.9, width = 0.6) +
        geom_text(aes(label = paste0(pct_coverage, "%")), vjust = -0.3, size = 5, fontface = "bold") +
        scale_fill_manual(values = threat_colors) +
        scale_y_continuous(limits = c(0, 110), expand = expansion(mult = c(0, 0))) +
        labs(title = "GBIF Coverage (%)", x = NULL, y = "Coverage (%)") +
        theme(legend.position = "none")
      
      print(p1 + p2 + plot_annotation(
        title = "Threatened Species Coverage in GBIF",
        subtitle = "CR = Critically Endangered, EN = Endangered, VU = Vulnerable, NT = Near Threatened, LC = Least Concern"
      ))
    }
  }
}

if (!has_threat_data) {
  cat("No threat status data available. Ensure National Red List is joined to national backbone.\n")
}
```

---

# 4. Priority Recommendations

## 4.1 Key Findings

```{r key-findings, results='asis'}
cat("### Summary of Data Gaps\n\n")

if (!is.null(dashboard) && nrow(dashboard) > 0) {
  cat("**Spatial Coverage**: ", safe_col(dashboard, "cells_10km_pct_coverage", "N/A"), 
      "% of the country's 10km grid cells have occurrence data.\n\n", sep = "")
  cat("**Data Recency**: ", safe_col(dashboard, "pct_stale_5y_10km", "N/A"), 
      "% of cells have not been surveyed in over 5 years.\n\n", sep = "")
  cat("**Taxonomic Coverage**: ", safe_col(dashboard, "taxa_pct_coverage", "N/A"), 
      "% of backbone species have GBIF records.\n\n", sep = "")
}

if (has_threat_data && !is.null(threat_stats)) {
  cr_en_missing <- threat_stats |>
    filter(threatStatus %in% c("CR", "EN")) |>
    summarise(total = sum(n_missing)) |>
    pull(total)
  cat("**Missing Threatened Species**: ", comma(cr_en_missing), 
      " Critically Endangered and Endangered species lack GBIF records.\n\n", sep = "")
}
```

## 4.2 Priority Actions

```{r priority-actions}
priority_zero <- safe_fread(here(p_integrated, "priority_zero_coverage_cells.csv"))
priority_stale <- safe_fread(here(p_integrated, "priority_stale_cells.csv"))

n_zero <- if (!is.null(priority_zero)) nrow(priority_zero) else 0
n_stale <- if (!is.null(priority_stale)) nrow(priority_stale) else 0
n_cr_en <- if (has_threat_data && !is.null(threat_stats)) {
  threat_stats |> filter(threatStatus %in% c("CR", "EN")) |> summarise(sum(n_missing)) |> pull()
} else 0

actions <- tibble(
  Priority = c("High", "High", "Medium", "Medium", "Low"),
  Category = c("Spatial", "Taxonomic", "Temporal", "Taxonomic", "Spatial"),
  Action = c(
    "Survey zero-coverage grid cells",
    "Monitor CR and EN species lacking GBIF records",
    "Re-survey cells with >5 years since last observation",
    "Digitize museum specimens for underrepresented orders",
    "Expand citizen science programs to low-coverage regions"
  ),
  Target = c(
    paste(comma(n_zero), "cells"),
    paste(comma(n_cr_en), "species"),
    paste(comma(n_stale), "cells"),
    "Focus on low-coverage taxonomic groups",
    "Remote and mountainous regions"
  )
)

actions |> kable()
```

---

# 5. Data Sources & Methods

**GBIF Occurrence Cubes**

- 10km grid: https://doi.org/10.15468/dl.wzv3uc
- 50km grid: https://doi.org/10.15468/dl.qyp3uw

**Taxonomic Reference**

- National Taxonomy Backbone: https://doi.org/10.15468/j43wfc
- National Red List: https://doi.org/10.15468/jhwkpq

**Analysis Date**: `r if(!is.null(dashboard)) safe_col(dashboard, "analysis_date", Sys.Date()) else Sys.Date()`

---

*For questions or more information, please contact your GBIF node.*
