---
title: "Species of Concern"
subtitle: "Threatened, Invasive, and Sensitive Species in GBIF"
author: "GBIF Gap Finder"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
  pdf_document:
    toc: true
    latex_engine: xelatex
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6, dpi = 150
)

library(here); library(dplyr); library(tidyr); library(ggplot2)
library(scales); library(sf); library(knitr); library(kableExtra)

data_path <- here("shiny_app", "gap_finder", "data", "shiny_data.rds")
app_data <- readRDS(data_path)
safe_get <- function(name) if (name %in% names(app_data)) app_data[[name]] else NULL

match_summary   <- safe_get("taxonomic_match_summary")
species_scope   <- safe_get("species_scope_lookup")
grid_10km       <- safe_get("grid_10km")
admin_level1    <- safe_get("admin_level1")

# Scope-specific data
threatened_sg   <- safe_get("threatened_spatial_gaps")
threatened_br   <- safe_get("threatened_basis_recent")
invasive_sg     <- safe_get("invasive_spatial_gaps")
invasive_br     <- safe_get("invasive_basis_recent")
invasive_ts     <- safe_get("invasive_time_summary")
sensitive_sg    <- safe_get("sensitive_spatial_gaps")
sensitive_br    <- safe_get("sensitive_basis_recent")

pal <- list(sage = "#2A7F62", slate = "#4477AA", sand = "#CCBB44",
            coral = "#EE6677", plum = "#AA3377")
theme_gaps <- theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "#666", size = 11),
        panel.grid.minor = element_blank(), legend.position = "bottom")
theme_set(theme_gaps)

# Helper: get threat column
get_threat_col <- function(ms) {
  tc <- intersect(c("threatStatus", "threatStatus_redlist", "threatStatus_backbone"), names(ms))
  if (length(tc) > 0) tc[1] else NA
}

# Species-of-concern counts computed exactly as the app's ov_concern boxes
# (match_summary, matched_any = in GBIF), so this report ties out to the app.
concern_stats <- if (!is.null(match_summary)) {
  tc  <- get_threat_col(match_summary)
  thr <- if (!is.na(tc)) match_summary[[tc]] else rep(NA_character_, nrow(match_summary))
  matched <- if ("matched_any"  %in% names(match_summary)) match_summary$matched_any else rep(FALSE, nrow(match_summary))
  inv <- if ("is_invasive"  %in% names(match_summary)) match_summary$is_invasive  else rep(FALSE, nrow(match_summary))
  sen <- if ("is_sensitive" %in% names(match_summary)) match_summary$is_sensitive else rep(FALSE, nrow(match_summary))
  is_thr <- thr %in% c("CR","EN","VU","NT")
  list(thr_total = sum(is_thr, na.rm = TRUE), thr_gbif = sum(is_thr & matched, na.rm = TRUE),
       inv_total = sum(inv, na.rm = TRUE),    inv_gbif = sum(inv & matched, na.rm = TRUE),
       sen_total = sum(sen, na.rm = TRUE),    sen_gbif = sum(sen & matched, na.rm = TRUE),
       unmon_cr  = sum(thr == "CR" & !matched, na.rm = TRUE),
       unmon_en  = sum(thr == "EN" & !matched, na.rm = TRUE),
       unmon_inv = sum(inv & !matched, na.rm = TRUE))
} else NULL

# Per-status Red List coverage from match_summary (matched_any) — the same
# object and definition the app's concern_threat_coverage chart uses, so the
# report ties out to the dashboard. All statuses (incl. DD) are shown; only
# CR/EN/VU/NT are counted as threatened.
threat_cov <- if (!is.null(match_summary)) {
  tc <- get_threat_col(match_summary)
  if (!is.na(tc)) {
    match_summary |> as_tibble() |>
      mutate(threatStatus = .data[[tc]]) |>
      filter(threatStatus %in% c("CR", "EN", "VU", "NT", "DD")) |>
      group_by(threatStatus) |>
      summarise(n_taxa = n(),
                n_in_gbif = sum(matched_any, na.rm = TRUE),
                n_missing = n_taxa - n_in_gbif,
                pct_coverage = round(100 * n_in_gbif / n_taxa, 1),
                .groups = "drop")
  } else NULL
} else NULL
```

# Overview

This report analyses three categories of species requiring special attention: **threatened** species on the national Red List, **invasive** species, and **sensitive** species with restricted location data.

```{r scope-summary}
if (!is.null(concern_stats)) {
  cs <- concern_stats
  df <- tibble(
    Category      = c("Threatened (CR/EN/VU/NT)", "Invasive", "Sensitive"),
    `In backbone` = c(cs$thr_total, cs$inv_total, cs$sen_total),
    `In GBIF`     = c(cs$thr_gbif,  cs$inv_gbif,  cs$sen_gbif),
    Unmonitored   = c(cs$thr_total - cs$thr_gbif, cs$inv_total - cs$inv_gbif, cs$sen_total - cs$sen_gbif)
  )
  kable(df, align = c("l", "r", "r", "r")) |> kable_styling(full_width = FALSE)
}
```

Of the threatened species with **no GBIF records at all**, `r if (!is.null(concern_stats)) concern_stats$unmon_cr else 0` are Critically Endangered (CR) and `r if (!is.null(concern_stats)) concern_stats$unmon_en else 0` Endangered (EN); `r if (!is.null(concern_stats)) concern_stats$unmon_inv else 0` invasive species are entirely unmonitored.

# Threatened Species {.tabset}

## Coverage by Status

```{r threat-coverage}
if (!is.null(threat_cov) && nrow(threat_cov) > 0) {
  df <- threat_cov |> filter(threatStatus %in% c("CR", "EN", "VU", "NT", "DD"))
  threat_colors <- c(CR = pal$coral, EN = pal$sand, VU = "#EE8866",
                     NT = pal$sage, DD = pal$slate)

  ggplot(df, aes(x = factor(threatStatus, c("CR","EN","VU","NT","DD")),
                 y = pct_coverage, fill = threatStatus)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = paste0(pct_coverage, "%")), vjust = -0.3, size = 4) +
    scale_fill_manual(values = threat_colors) +
    scale_y_continuous(limits = c(0, 110)) +
    labs(title = "GBIF Coverage by Red List Category", x = "", y = "Coverage (%)",
         subtitle = "CR/EN/VU/NT are threatened; DD (Data Deficient) is shown for completeness, not counted as threatened")
}
```

## Missing Threatened Species

```{r threat-missing}
if (!is.null(match_summary)) {
  tc <- get_threat_col(match_summary)
  if (!is.na(tc)) {
    missing <- match_summary |>
      filter(.data[[tc]] %in% c("CR", "EN", "VU", "NT"), !matched_any) |>
      group_by(.data[[tc]]) |>
      summarise(n_missing = n(), .groups = "drop") |>
      rename(Status = 1)

    ggplot(missing, aes(x = Status, y = n_missing, fill = Status)) +
      geom_col(show.legend = FALSE) +
      geom_text(aes(label = n_missing), vjust = -0.3) +
      scale_fill_manual(values = c(CR = pal$coral, EN = pal$sand,
                                   VU = "#EE8866", NT = pal$sage)) +
      labs(title = "Missing Threatened Species by Category",
           subtitle = "Backbone species with zero GBIF records",
           x = "", y = "Number of missing species")
  }
}
```

## Spatial Distribution

```{r threat-map, fig.height=9}
if (!is.null(grid_10km) && !is.null(threatened_sg)) {
  sg <- threatened_sg |> filter(basisofrecord == "all")
  map_sf <- grid_10km |>
    left_join(sg |> select(eeacellcode, occurrences, n_species), by = "eeacellcode") |>
    mutate(occ_cat = case_when(
      is.na(occurrences) | occurrences == 0 ~ "No data",
      occurrences <= 10 ~ "1–10", occurrences <= 100 ~ "11–100",
      occurrences <= 1000 ~ "101–1K", TRUE ~ "> 1K"
    ) |> factor(levels = c("1–10", "11–100", "101–1K", "> 1K", "No data")))

  p <- ggplot(map_sf) +
    geom_sf(aes(fill = occ_cat), color = NA) +
    scale_fill_manual(values = c("1–10" = "#CCEEDD", "11–100" = pal$sage,
      "101–1K" = pal$sand, "> 1K" = pal$coral, "No data" = "#e0e0e0")) +
    labs(title = "Threatened Species — Spatial Distribution", fill = "Occurrences") +
    theme_void() + theme(legend.position = "right")
  if (!is.null(admin_level1)) {
    p <- p + geom_sf(data = st_transform(admin_level1, st_crs(map_sf)),
                     fill = NA, color = "#888", linewidth = 0.3)
  }
  print(p)
}
```

## How Recorded

```{r threat-bor, fig.height=4}
if (!is.null(threatened_br)) {
  df <- threatened_br |> filter(basisofrecord != "all") |>
    mutate(bor = gsub("_", " ", basisofrecord) |> tools::toTitleCase()) |>
    arrange(desc(occ_total))

  ggplot(df, aes(x = reorder(bor, occ_total), y = occ_total)) +
    geom_col(fill = pal$sage) +
    geom_text(aes(label = comma(occ_total)), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.2))) +
    labs(title = "Threatened Species — Basis of Record", x = "", y = "Occurrences")
}
```

# Invasive Species {.tabset}

## By Order and Family

```{r inv-by-order, fig.height=6}
if (!is.null(match_summary) && "is_invasive" %in% names(match_summary)) {
  ms_inv <- match_summary |> filter(is_invasive == TRUE)

  if (nrow(ms_inv) > 0 && "order" %in% names(ms_inv)) {
    df <- ms_inv |>
      filter(!is.na(order), order != "") |>
      group_by(order) |>
      summarise(n_total = n(), n_gbif = sum(matched_any, na.rm = TRUE),
                n_missing = sum(!matched_any, na.rm = TRUE), .groups = "drop") |>
      arrange(desc(n_total)) |> head(15)

    df_long <- df |> pivot_longer(c(n_gbif, n_missing), names_to = "status", values_to = "n") |>
      mutate(status = ifelse(status == "n_gbif", "In GBIF", "Missing"))

    ggplot(df_long, aes(x = reorder(order, n_total), y = n, fill = status)) +
      geom_col() +
      scale_fill_manual(values = c("In GBIF" = pal$sage, "Missing" = pal$coral)) +
      coord_flip() +
      labs(title = "Invasive Species by Order", x = "", y = "Number of species", fill = "")
  }
}
```

## Spatial Distribution

```{r inv-map, fig.height=9}
if (!is.null(grid_10km) && !is.null(invasive_sg)) {
  sg <- invasive_sg |> filter(basisofrecord == "all")
  map_sf <- grid_10km |>
    left_join(sg |> select(eeacellcode, occurrences, n_species), by = "eeacellcode") |>
    mutate(occ_cat = case_when(
      is.na(occurrences) | occurrences == 0 ~ "No data",
      occurrences <= 10 ~ "1–10", occurrences <= 100 ~ "11–100",
      occurrences <= 1000 ~ "101–1K", TRUE ~ "> 1K"
    ) |> factor(levels = c("1–10", "11–100", "101–1K", "> 1K", "No data")))

  p <- ggplot(map_sf) +
    geom_sf(aes(fill = occ_cat), color = NA) +
    scale_fill_manual(values = c("1–10" = "#FFAABB", "11–100" = pal$coral,
      "101–1K" = pal$sand, "> 1K" = "#114477", "No data" = "#e0e0e0")) +
    labs(title = "Invasive Species — Spatial Distribution", fill = "Occurrences") +
    theme_void() + theme(legend.position = "right")
  if (!is.null(admin_level1)) {
    p <- p + geom_sf(data = st_transform(admin_level1, st_crs(map_sf)),
                     fill = NA, color = "#888", linewidth = 0.3)
  }
  print(p)
}
```

## Temporal Trend

```{r inv-temporal, fig.height=5}
if (!is.null(invasive_ts)) {
  df <- invasive_ts |>
    filter(basisofrecord == "all") |>
    mutate(yearmonth = as.integer(gsub("-", "", as.character(yearmonth))),
           year = as.integer(substr(as.character(yearmonth), 1, 4))) |>
    group_by(year) |>
    summarise(occ = sum(as.numeric(occurrences), na.rm = TRUE), .groups = "drop") |>
    filter(!is.na(year), year >= 1900)

  ggplot(df, aes(x = year, y = occ)) +
    geom_col(fill = pal$coral) +
    scale_y_continuous(labels = label_comma()) +
    labs(title = "Invasive Species — Observations Over Time", x = "Year", y = "Occurrences")
}
```

# Sensitive Species {.tabset}

## Data Availability

```{r sens-summary}
if (!is.null(match_summary) && "is_sensitive" %in% names(match_summary)) {
  ms_sens <- match_summary |> filter(is_sensitive == TRUE)
  n_sens <- nrow(ms_sens)
  n_in_gbif <- sum(ms_sens$matched_any, na.rm = TRUE)

  sens_stats <- tibble(
    Metric = c("Known sensitive species", "In GBIF", "Missing from GBIF", "Coverage"),
    Value = c(comma(n_sens), comma(n_in_gbif), comma(n_sens - n_in_gbif),
              paste0(round(100 * n_in_gbif / max(n_sens, 1), 1), "%"))
  )
  kable(sens_stats, align = c("l", "r")) |> kable_styling(full_width = FALSE)
}
```

## Generalization Categories

```{r sens-gen}
if (!is.null(match_summary) && "sensitivity_category" %in% names(match_summary)) {
  ms_sens <- match_summary |> filter(is_sensitive == TRUE, !is.na(sensitivity_category))
  if (nrow(ms_sens) > 0) {
    gen_df <- ms_sens |> count(sensitivity_category, sort = TRUE) |>
      rename(Category = sensitivity_category, Species = n)
    kable(gen_df, align = c("l", "r")) |> kable_styling(full_width = FALSE)
  }
}
```

Sensitive species have coordinates generalised by GBIF to protect them from collection pressure, habitat disturbance, or trade. The generalization category (5 km, 25 km, or 50 km) indicates the radius within which coordinates are randomised. At 10 km grid resolution, 5 km-generalised species retain reasonable spatial accuracy; 25 km and 50 km species should be interpreted with caution.

## Spatial Distribution

```{r sens-map, fig.height=9}
if (!is.null(grid_10km) && !is.null(sensitive_sg)) {
  sg <- sensitive_sg |> filter(basisofrecord == "all")
  map_sf <- grid_10km |>
    left_join(sg |> select(eeacellcode, occurrences, n_species), by = "eeacellcode") |>
    mutate(occ_cat = case_when(
      is.na(occurrences) | occurrences == 0 ~ "No data",
      occurrences <= 10 ~ "1–10", occurrences <= 100 ~ "11–100",
      occurrences <= 1000 ~ "101–1K", TRUE ~ "> 1K"
    ) |> factor(levels = c("1–10", "11–100", "101–1K", "> 1K", "No data")))

  p <- ggplot(map_sf) +
    geom_sf(aes(fill = occ_cat), color = NA) +
    scale_fill_manual(values = c("1–10" = "#CCDDF0", "11–100" = pal$slate,
      "101–1K" = pal$plum, "> 1K" = "#114477", "No data" = "#e0e0e0")) +
    labs(title = "Sensitive Species — Spatial Distribution",
         subtitle = "Coordinates may be generalised; interpret with care",
         fill = "Occurrences") +
    theme_void() + theme(legend.position = "right")
  if (!is.null(admin_level1)) {
    p <- p + geom_sf(data = st_transform(admin_level1, st_crs(map_sf)),
                     fill = NA, color = "#888", linewidth = 0.3)
  }
  print(p)
}
```

## How Recorded

```{r sens-bor, fig.height=4}
if (!is.null(sensitive_br)) {
  df <- sensitive_br |> filter(basisofrecord != "all") |>
    mutate(bor = gsub("_", " ", basisofrecord) |> tools::toTitleCase()) |>
    arrange(desc(occ_total))

  ggplot(df, aes(x = reorder(bor, occ_total), y = occ_total)) +
    geom_col(fill = pal$plum) +
    geom_text(aes(label = comma(occ_total)), hjust = -0.1, size = 3.5) +
    coord_flip() +
    scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.2))) +
    labs(title = "Sensitive Species — Basis of Record", x = "", y = "Occurrences")
}
```

---

*Report generated by `gbif_gap_finder` on `r Sys.time()`.*
