---
title: "Taxonomy Reconciliation QA Review"
subtitle: "GBIF Gap Analysis — Verifying 09a Matching Results"
author: "gbifgaps"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    fig_width: 12
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 7,
  dpi = 150
)

if (!knitr::is_html_output()) {
  knitr::opts_chunk$set(echo = FALSE)
}

library(here)
library(data.table)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(scales)
library(viridis)
library(DT)
library(knitr)
library(glue)

theme_qa <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "grey40"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 11, face = "bold")
  )
theme_set(theme_qa)

p_data_proc <- here("data_proc")
```

# Overview

This report reviews the taxonomy reconciliation produced by `09a_reconcile_taxonomy.R`.
It helps you verify match quality and identify species that could be resolved with
manual overrides.

```{r load-data}
recon <- as.data.table(readRDS(here(p_data_proc, "taxonomic_reconciliation.rds")))
taxa  <- as.data.table(readRDS(here(p_data_proc, "taxa_reference_current.rds")))

taxa[, name_std := tolower(trimws(scientificName))]
taxa[, is_accepted := (taxonID == acceptedNameUsageID)]

n_total     <- nrow(recon)
n_matched   <- sum(recon$match_tier != "unmatched")
n_unmatched <- sum(recon$match_tier == "unmatched")
occ_total   <- sum(as.numeric(recon$total_occ))
occ_matched <- recon[match_tier != "unmatched", sum(as.numeric(total_occ))]
```

**Reconciliation summary:** `r comma(n_matched)` of `r comma(n_total)` GBIF species
matched to the national backbone (`r round(100 * n_matched / n_total, 1)`% of species,
`r round(100 * occ_matched / occ_total, 1)`% of occurrences).

```{r tier-summary}
tier_summary <- recon[, .(
  Species = .N,
  Occurrences = sum(as.numeric(total_occ)),
  `% Species` = round(100 * .N / n_total, 1),
  `% Occurrences` = round(100 * sum(as.numeric(total_occ)) / occ_total, 2)
), by = .(Tier = match_tier, Type = match_type)][order(Tier, -Species)]

kable(tier_summary, format.args = list(big.mark = ","))
```

```{r tier-plot, fig.height=5}
tier_occ <- recon[, .(
  occ = sum(as.numeric(total_occ))
), by = match_tier]

tier_order <- c("tier1", "tier2", "tier2_ambiguous", "tier3", "tier4", "unmatched")
tier_labels <- c("Tier 1: Accepted", "Tier 2: Synonym", "Tier 2: Ambiguous",
                  "Tier 3: Infraspecific", "Tier 4: API", "Unmatched")
tier_colors <- c("#2E7D32", "#43A047", "#FDD835", "#66BB6A", "#1565C0", "#B71C1C")

tier_occ[, tier_f := factor(match_tier, levels = tier_order, labels = tier_labels)]
tier_occ <- tier_occ[!is.na(tier_f)]

ggplot(tier_occ, aes(x = tier_f, y = occ / 1e6, fill = tier_f)) +
  geom_col() +
  scale_fill_manual(values = setNames(tier_colors, tier_labels), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(title = "Occurrences by Match Tier",
       x = NULL, y = "Occurrences (millions)") +
  coord_flip()
```


# 1. Ambiguous Synonym Matches {.tabset}

These species matched a backbone synonym name that points to **more than one accepted
taxon**. The script took the first match. Review these to confirm correctness —
especially the high-occurrence ones.

## All Ambiguous Matches

```{r ambiguous-table}
ambig <- recon[match_tier == "tier2_ambiguous",
               .(species, backbone_scientificName, total_occ,
                 match_name_used)][order(-total_occ)]

if (nrow(ambig) > 0) {
  # For each ambiguous match, find ALL possible accepted taxa
  ambig_detail <- rbindlist(lapply(seq_len(nrow(ambig)), function(i) {
    nm <- tolower(trimws(ambig$species[i]))
    # Find all synonym rows for this name
    syn_rows <- taxa[!is_accepted & name_std == nm]
    if (nrow(syn_rows) == 0) return(NULL)

    # Get their accepted taxa
    accepted_ids <- unique(syn_rows$acceptedNameUsageID)
    accepted_names <- taxa[taxonID %in% accepted_ids & is_accepted,
                           .(acceptedNameUsageID = taxonID, accepted_name = scientificName)]

    data.table(
      gbif_species = ambig$species[i],
      total_occ = ambig$total_occ[i],
      chosen_match = ambig$backbone_scientificName[i],
      all_candidates = paste(accepted_names$accepted_name, collapse = " | "),
      n_candidates = nrow(accepted_names)
    )
  }))

  datatable(
    ambig_detail,
    caption = glue("{nrow(ambig)} ambiguous synonym matches"),
    filter = "top",
    options = list(pageLength = 25, order = list(list(1, "desc"))),
    rownames = FALSE
  ) |> formatCurrency("total_occ", currency = "", digits = 0)
} else {
  cat("No ambiguous matches found.")
}
```

## Flagged for Review

These ambiguous matches have **> 5,000 occurrences** and deserve manual verification:

```{r ambiguous-flagged}
if (exists("ambig_detail") && nrow(ambig_detail) > 0) {
  flagged <- ambig_detail[total_occ > 5000][order(-total_occ)]
  if (nrow(flagged) > 0) {
    kable(flagged, format.args = list(big.mark = ","))
  } else {
    cat("No high-occurrence ambiguous matches.")
  }
} else {
  cat("No ambiguous matches to review.")
}
```


# 2. Tier 4 Matches (GBIF API) {.tabset}

These were resolved by querying the GBIF Species API for synonyms. The matches are
generally reliable (the API provides the authoritative synonym chain), but it's good
to scan the high-occurrence ones.

## Top 50 by Occurrences

```{r tier4-table}
t4 <- recon[match_tier == "tier4",
            .(species, backbone_scientificName, match_name_used,
              total_occ)][order(-total_occ)]

datatable(
  head(t4, 50),
  caption = glue("{comma(nrow(t4))} total Tier 4 matches — showing top 50"),
  filter = "top",
  options = list(pageLength = 25),
  rownames = FALSE
) |> formatCurrency("total_occ", currency = "", digits = 0)
```

## Match Patterns

```{r tier4-patterns}
if (nrow(t4) > 0) {
  # Classify: same epithet (genus move) vs different epithet (true synonym)
  t4[, gbif_epithet := word(tolower(species), 2)]
  t4[, backbone_epithet := word(tolower(backbone_scientificName), 2)]
  t4[, pattern := fifelse(gbif_epithet == backbone_epithet,
                          "Genus reclassification", "Different epithet (true synonym)")]

  pattern_summary <- t4[, .(n = .N, occ = sum(as.numeric(total_occ))), by = pattern]
  kable(pattern_summary, format.args = list(big.mark = ","),
        col.names = c("Pattern", "Species", "Occurrences"))
}
```


# 3. Unmatched Species Analysis {.tabset}

## Overview

```{r unmatched-overview}
unmatched <- recon[match_tier == "unmatched"]

unmatched_summary <- unmatched[, .(
  n = .N,
  total_occ = sum(as.numeric(total_occ)),
  max_occ = max(total_occ)
), by = match_type][order(-total_occ)]

kable(unmatched_summary, format.args = list(big.mark = ","),
      col.names = c("Category", "Species", "Total Occurrences", "Max Occurrences"))
```

## Top Unmatched Binomials — Epithet Candidates

For the highest-occurrence unmatched binomials, we check if the **species epithet**
exists in the backbone under a different genus. These are the best candidates for manual
resolution — likely genus reclassifications that the GBIF API missed or that fell
below the API threshold.

```{r epithet-analysis}
# Get unmatched binomials with significant occurrences
unmatched_binom <- unmatched[match_type %in% c("binomial_no_match", "below_api_threshold") &
                               grepl(" ", species)]
unmatched_binom[, `:=`(
  genus   = word(tolower(species), 1),
  epithet = word(tolower(species), 2)
)]

# Build backbone epithet lookup (accepted + synonym names, binomials only)
backbone_binomials <- taxa[nchar(name_std) > 0 &
                            sapply(strsplit(name_std, "\\s+"), length) == 2]
backbone_binomials[, `:=`(
  d_genus   = word(name_std, 1),
  d_epithet = word(name_std, 2)
)]

# For each unmatched species, find backbone names sharing the epithet
epithet_candidates <- rbindlist(lapply(
  seq_len(nrow(unmatched_binom)),
  function(i) {
    ep <- unmatched_binom$epithet[i]
    gn <- unmatched_binom$genus[i]

    # Find backbone names with same epithet but different genus
    matches <- backbone_binomials[d_epithet == ep & d_genus != gn]
    if (nrow(matches) == 0) return(NULL)

    # Get accepted name for each match
    match_info <- merge(
      matches[, .(name_std, taxonID, acceptedNameUsageID, is_accepted)],
      taxa[is_accepted == TRUE, .(taxonID, accepted_scientificName = scientificName)],
      by.x = "acceptedNameUsageID", by.y = "taxonID",
      all.x = TRUE
    )

    # Deduplicate: unique accepted taxa
    accepted_options <- unique(match_info[!is.na(accepted_scientificName),
                                          .(accepted_scientificName)])

    data.table(
      gbif_species     = unmatched_binom$species[i],
      total_occ        = unmatched_binom$total_occ[i],
      epithet          = ep,
      backbone_candidates = paste(unique(accepted_options$accepted_scientificName),
                                  collapse = " | "),
      n_candidates     = nrow(accepted_options),
      match_category   = unmatched_binom$match_type[i]
    )
  }
))
```

```{r epithet-results}
if (!is.null(epithet_candidates) && nrow(epithet_candidates) > 0) {
  # Split into high-confidence (1 candidate) and ambiguous (>1)
  single <- epithet_candidates[n_candidates == 1][order(-total_occ)]
  multi  <- epithet_candidates[n_candidates > 1][order(-total_occ)]
  none   <- unmatched_binom[!species %in% epithet_candidates$gbif_species][order(-total_occ)]

  cat(glue("**{comma(nrow(epithet_candidates))}** unmatched species share an epithet with the backbone:
- **{nrow(single)}** have exactly 1 backbone candidate (high confidence)
- **{nrow(multi)}** have multiple candidates (need manual review)
- **{comma(nrow(none))}** have no epithet match in the backbone at all\n\n"))
} else {
  cat("No epithet candidates found.\n")
}
```

### High-Confidence Candidates (1 backbone match)

These have a single backbone species sharing the epithet — almost certainly genus
reclassifications. **These can likely be added as manual overrides directly.**

```{r single-candidates}
if (exists("single") && nrow(single) > 0) {
  datatable(
    single[, .(gbif_species, backbone_candidates, total_occ, match_category)],
    caption = glue("{nrow(single)} high-confidence epithet matches"),
    filter = "top",
    options = list(pageLength = 50, order = list(list(2, "desc"))),
    rownames = FALSE,
    colnames = c("GBIF Species", "Backbone Match", "Occurrences", "Why Unmatched")
  ) |> formatCurrency("total_occ", currency = "", digits = 0)
} else {
  cat("None found.")
}
```

### Ambiguous Candidates (> 1 backbone match)

These need manual review — the epithet appears under multiple genera in the backbone,
so automated resolution is risky.

```{r multi-candidates}
if (exists("multi") && nrow(multi) > 0) {
  datatable(
    multi[, .(gbif_species, backbone_candidates, n_candidates, total_occ)],
    caption = glue("{nrow(multi)} ambiguous epithet matches"),
    filter = "top",
    options = list(pageLength = 50, order = list(list(3, "desc"))),
    rownames = FALSE,
    colnames = c("GBIF Species", "Backbone Candidates", "N", "Occurrences")
  ) |> formatCurrency("total_occ", currency = "", digits = 0)
} else {
  cat("None found.")
}
```

### No Epithet Match

These species have no shared epithet with any backbone name — they are likely
genuinely absent from the national backbone (foreign vagrants, non-native species,
or taxa the backbone doesn't cover).

```{r no-epithet}
if (exists("none") && nrow(none) > 0) {
  datatable(
    head(none[, .(species, total_occ, match_type)], 100),
    caption = glue("{comma(nrow(none))} species with no epithet in the backbone — top 100"),
    filter = "top",
    options = list(pageLength = 25, order = list(list(1, "desc"))),
    rownames = FALSE,
    colnames = c("GBIF Species", "Occurrences", "Category")
  ) |> formatCurrency("total_occ", currency = "", digits = 0)
} else {
  cat("None — all unmatched binomials share an epithet with the backbone.")
}
```


# 4. Manual Override Template {.tabset}

## Export for Review

This section exports a CSV you can edit to add manual overrides. Fill in the
`accept` column with `TRUE` for matches you've verified, and `backbone_override`
with the correct backbone name if the auto-suggested candidate is wrong.

```{r export-overrides}
override_template <- NULL

if (exists("single") && nrow(single) > 0) {
  override_single <- single[, .(
    gbif_species,
    suggested_backbone = backbone_candidates,
    total_occ,
    source = "epithet_single_match",
    accept = NA,
    backbone_override = NA_character_,
    notes = NA_character_
  )]
  override_template <- override_single
}

if (exists("ambig_detail") && nrow(ambig_detail) > 0) {
  override_ambig <- ambig_detail[, .(
    gbif_species,
    suggested_backbone = chosen_match,
    total_occ,
    source = "tier2_ambiguous",
    accept = NA,
    backbone_override = NA_character_,
    notes = all_candidates
  )]
  override_template <- rbindlist(list(override_template, override_ambig), fill = TRUE)
}

if (!is.null(override_template) && nrow(override_template) > 0) {
  setorder(override_template, -total_occ)

  override_path <- here(p_data_proc, "taxonomy_manual_overrides_template.csv")
  fwrite(override_template, override_path)

  cat(glue("**Exported {nrow(override_template)} candidates** to:
`{override_path}`

Edit this file, set `accept = TRUE` for verified matches, then 09a can
read it as a Tier 0 override on the next run.\n\n"))

  datatable(
    head(override_template, 50),
    caption = "Top 50 override candidates (full list in CSV)",
    filter = "top",
    options = list(pageLength = 25, order = list(list(2, "desc"))),
    rownames = FALSE
  ) |> formatCurrency("total_occ", currency = "", digits = 0)
} else {
  cat("No override candidates to export.")
}
```

## How to Use Overrides

Once you've reviewed the CSV:

1. Open `data_proc/taxonomy_manual_overrides_template.csv`
2. For each row, set `accept = TRUE` if the suggested match is correct
3. If the suggestion is wrong but you know the correct name, fill in `backbone_override`
4. Save as `data_proc/taxonomy_manual_overrides.csv`
5. A future version of 09a can read this file as a "Tier 0" pre-match step


# 5. Summary Statistics

```{r final-summary}
cat(glue("
## Reconciliation Health Check

- **Total GBIF species:** {comma(n_total)}
- **Matched (Tiers 1-4):** {comma(n_matched)} ({round(100 * n_matched / n_total, 1)}%)
- **Occurrence coverage:** {round(100 * occ_matched / occ_total, 1)}%
- **Ambiguous matches needing review:** {nrow(ambig)} ({comma(sum(ambig$total_occ))} occ)
- **Unmatched with epithet candidate:** {if (exists('epithet_candidates') && nrow(epithet_candidates) > 0) comma(nrow(epithet_candidates)) else '0'}
- **Truly unresolvable:** {if (exists('none')) comma(nrow(none)) else 'unknown'}

### Potential improvement from manual overrides

"))

if (exists("single") && nrow(single) > 0) {
  potential_occ <- sum(as.numeric(single$total_occ))
  new_coverage <- round(100 * (occ_matched + potential_occ) / occ_total, 2)
  cat(glue("If all **{nrow(single)}** high-confidence epithet matches are accepted,
occurrence coverage would increase from **{round(100 * occ_matched / occ_total, 2)}%**
to **{new_coverage}%** (+{comma(potential_occ)} occurrences).\n"))
}
```
